---
title: ggplot2 basics
layout: default
output: bookdown::html_chapter
---

Chapter 2

Basics of data visualization using ggplot2, the multi-layer,
multi-panel grammer of graphics.

Chapter outline:
- Brief history of the grammar of graphics
- Basic ggplot
- Multi-layer ggplots
- Multi-panel ggplots

# Brief history of the grammar of graphics

Wilkinson proposed a general grammar of graphics to describe and
create a large class of static data visualizations
[Wilkinson 2006]. Wickham implemented several ideas from that grammar
in the ggplot2 R package [Wickham 2009]. Wilkinson's grammar describes
\vocab{static} graphics, which can be viewed equally well on a
computer screen or on paper.

The main topic of this book is animint, an R package for
\vocab{interactive} graphics. In contrast to static graphics,
interactive graphics are best viewed on a computer with a mouse and
keyboard that can be used to interact with the plot. Since many
concepts from static graphics are also useful for interactive graphics
(e.g. legends), the animint package is implemented as an extension of
ggplot2. In this chapter we will introduce the main features of
ggplot2 which will also be useful for interactive plot design in later
chapters.

# A basic ggplot with one geom and one facet

The R code below produces a scatterplot of life expectancy and
fertility rate for each country in the WorldBank data set.

```{r}
data(WorldBank, package="animint")
library(ggplot2)
p <- ggplot()+
  geom_point(aes(x=life.expectancy, y=fertility.rate, color=region),
             data=subset(WorldBank, year==1975))
print(p)
```

Each ggplot contain exactly one \vocab{layer}, a `geom_point` which
displays a circle for every row the in data set specified. The `aes`
specifies the mapping of data variables to visual properties. The `x`
and `y` properties are used for the horizontal and vertical position,
and the `color` property specifies the color each point.

This ggplot can be rendered by animint via

```{r}
library(animint)
viz <- list(scatter=p)
structure(viz, class="animint")
```

The first line loads the animint package, the second line assigns the
ggplot `p` to a named element of the `viz` list, and the third line
gives the list the class `animint`. When executing this code on the R
command line, R will run the `print.animint` function via the S3
object system (http://adv-r.had.co.nz/OO-essentials.html#s3). That
will run the `animint2dir` function, which converts the animint `viz`
list to a directory of data and code files that can be rendered in a
web browser.

When viewed in a web browser, the plot should look mostly the same as
the version in R. One difference is that the region legend is
interactive: clicking a legend entry will hide or show the points of
that color.

TODO: discuss theme_animint, title, etc?

# Multi-layer data visualization (multiple geoms)

# Multi-panel data visualization (facets)

# Chapter summary and exercises
