---
title: Sharing
layout: default
output: bookdown::html_chapter
---

# Chapter 5, Sharing animints

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch05-figures/")
```

This chapter explains several methods for sharing your interactive
data visualizations on the web. After reading this chapter, you will
be able to share and publish animints

* on your personal computer.
* on [gist.github.com](https://gist.github.com/) and view them on
  [bl.ocks.org](http://bl.ocks.org/).
* in [R Markdown](http://rmarkdown.rstudio.com/) documents.
* using any web server, including
  [GitHub pages](https://pages.github.com/).

## Compile to a local directory {#animint2dir}

When experimenting with different interactive data visualization
designs, it is useful to preview them on your personal computer before
publishing on the web. This section discusses two methods for
compiling animints to a local directory.

So far in previous chapters we have only discussed one method for
creating interactive data visualizations. If `viz` is an animint (list
of ggplots and options), then running `structure(viz,
class="animint")` on the R command line compiles that animint to a
temporary directory. After `structure` attaches the `"animint"` class
to the `viz` list, R calls the `print.animint` S3 method, which
compiles the animint to a new temporary directory on your personal
computer.

Rather than saving each animint to a separate temporary directory, you
can specify an output directory using the `animint2dir` function. If
you want to save `viz` in the `"path/to/figure-2"` directory, use
`animint2dir(viz, "path/to/figure-2")`. If that directory does not
exist, then it will be created. If that directory does exist, then any
files in that directory will be overwritten. To view the data viz,
navigate to `path/to/figure-2/index.html` in a web browser.

## Publish on GitHub Gist {#animint2gist}

This section explains how to publish an animint on the web via
[GitHub Gist](https://gist.github.com/). Gist was originally meant for
sharing code, but it also works for sharing animints.

To publish an animint on Gist, you need two things: (1) a GitHub
account, and (2) the `gistr` package. First, if you don't have a
GitHub account, you can
[sign up for free](https://github.com/join). Second, install the
`gistr` package:

```{r}
if(!require(gistr))install.packages("gistr")
```

Now you should be able to use `animint2gist(viz)` to publish `viz` as
a new gist. That command will post the files to a new gist, for
example https://gist.github.com/tdhock/2b029fe9abc8eb300f9f/. It will
then open a new web browser window at the corresponding bl.ocks.org
URL, for example http://bl.ocks.org/tdhock/raw/2b029fe9abc8eb300f9f/

You can use the bl.ocks.org URL to share your data visualization with
others.

Beware that Gist imposes some limits. In March 2016 Gist imposed a
maximum of 300 files per gist, and a maximum of 1 megabyte per
file. The animint2gist function should stop with an error if your
animint exceeds either of these limits. If you run into this problem,
you may be able to use the [chunk_vars](Ch06-other.html#chunk_vars)
option to edit the number and size of files produced by
animint. Otherwise, try publishing your animint using one of the other
methods below.

## Publish in R Markdown {#Rmd}

To include an animint in an R Markdown document, use `structure(viz,
class="animint")` inside of an R code chunk. R will run the
`knit_print.animint` S3 method, which compiles the animint to a local
directory, named based on the name of the R code chunk. For example a
code chunk named `viz-facets` will be saved in the directory
`vizfacets`. Make sure to put each animint in its own code chunk.

## Publish on a web server {#web-server}

Since animints are just directories with HTML, TSV, and JavaScript
files, you can publish them on any web server by simply copying the
directory to that server.

For example I executed the code in
https://github.com/tdhock/animint/blob/master/inst/examples/WorldBank-facets.R
to create the `WorldBank-facets` directory on my personal computer. I
copied that directory to my lab web server using `rsync -r
WorldBank-facets/ cbio.ensmp.fr:public_html/WorldBank-facets/` and so
I can view it on http://cbio.ensmp.fr/~thocking/WorldBank-facets/

If you don't have access to a personal web server, try using
[GitHub pages](https://pages.github.com/).

## Chapter summary and exercises {#exercises}

This chapter explained how to share animints on the web.

Exercises:

* Use `animint2gist` to publish a visualization of a data set from
  your domain of expertise. Post a link to the data viz (and your
  source code) on the
  [Animint Gallery wiki page](https://github.com/tdhock/animint/wiki/Gallery).

Next, [Chapter 6](Ch06-other.html) explains the different options that
can be used to customize an animint.
