---
title: Sharing
layout: default
output: bookdown::html_chapter
---

# Chapter 5, Sharing animints

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch05-figures/")
```

This chapter explains several methods for sharing your interactive
data visualizations on the web. After reading this chapter, you will
be able to view animints

* from a local directory on your personal computer.
* in [R Markdown](http://rmarkdown.rstudio.com/) documents.
* using any web server, including [Netlify
  Drop](https://app.netlify.com/drop).
* published using [GitHub pages](https://pages.github.com/), and
  organized into a group called a gallery.

## Compile to a local directory {#animint2dir}

When experimenting with different interactive data visualization
designs, it is useful to preview them on your personal computer before
publishing on the web. This section discusses two methods for
compiling animints to a local directory.

So far in previous chapters we have only discussed one method for
creating interactive data visualizations. If `viz` is an animint (list
of ggplots and options with class animint), then printing it on the R
command line compiles that animint to a temporary
directory, using code like this,

```{r}
set.seed(1)
ten.points <- data.frame(x=0:9, y=rnorm(10))
animint(
  point=ggplot()+
    geom_point(aes(
      x, y), 
      data=ten.points)
)
```

The code above saves the animint to a new temporary directory. Rather
than saving each animint to a separate temporary directory, you can
specify an output directory via the `out.dir` argument to
`animint`. If you want to save the animint in the
`"Ch05-sharing-ten-points"` directory, use:

```{r}
animint(
  point=ggplot()+
    geom_point(aes(
      x, y), 
      data=ten.points),
  out.dir="Ch05-sharing-ten-points"
)
```

If the parent of `out.dir` does not exist, that is an error (you can
use `dir.create` to create the parent if necessary). If `out.dir` does
not exist, then it will be created. If `out.dir` does exist (and
contains a file named `animint.js`), then any files in that directory
will be overwritten. To view the data viz, navigate to
`Ch05-sharing-ten-points/index.html` in a web browser (which should be
done automatically / by default). If the web page is blank, you may
need to configure your browser to allow execution of local JavaScript
code, [as explained on our
FAQ](https://github.com/tdhock/animint2/wiki/FAQ#web-browser-on-local-indexhtml-file-is-blank).

Internally, R calls the `print.animint` S3 method, which calls
`animint2dir` to compile it to a new temporary directory on your
personal computer. Generally we advise to avoid calling `animint2dir`
directly, but it is useful if you want to avoid opening lots of
similar browser windows when repeatedly revising and rendering an
animint. You can prevent the default behavior of opening a browser
window via:

```{r eval=F}
viz <- animint(
  point=ggplot()+
    geom_point(aes(
      x, y), 
      data=ten.points))
animint2dir(
  viz, 
  out.dir="Ch05-sharing-ten-points-again", 
  open.browser=FALSE)
```

## Publish in R Markdown {#Rmd}

To include an animint in an R Markdown document, use `animint(...)`
inside of an R code chunk. R will run the `knit_print.animint` S3
method, which compiles the animint to a local directory, named based
on the name of the R code chunk. For example a code chunk named
`viz-facets` will be saved in the directory `vizfacets`. Make sure to
put each animint in its own code chunk (do not put two animints in the
same code chunk).

## Publish on a web server {#web-server}

Since animints are just directories with HTML, TSV, and JavaScript
files, you can publish them on any web server by simply copying the
directory to that server.

For example I executed the code in
https://github.com/animint/animint2/blob/master/inst/examples/WorldBank-facets.R
to create the `WorldBank-facets` directory on my personal computer. I
copied that directory to my lab web server using `rsync -r
WorldBank-facets/ monsoon.hpc.nau.edu:genomic-ml/WorldBank-facets/` and so
I can view it on https://rcdata.nau.edu/genomic-ml/WorldBank-facets/

If you don't have access to a personal/lab web server, try using
one of the methods described below, which are free for anyone.

## Publish on Netlify Drop {#netlifydrop}

[Netlify Drop](https://app.netlify.com/drop) is for hosting static web
sites. To publish your data viz there, simply drag a directory to that
web page (it can be a directory resulting from `animint2dir`, or from
`rmarkdown::render` if your animint is inside Rmd, as described
above).  After the upload completes, you will be provided a link which
can be used to view the files in that directory. No registration/login
is required, but if you do not register an account, your data viz will
be deleted after one hour. You can register for a free account if you
want your data viz to be available longer.

## Publish on GitHub Pages {#pages}

[GitHub Pages](https://pages.github.com/) is a service that provides
static web site hosting, and can be used to publish animints. To
publish an animint on GitHub Pages, you need a GitHub account, and the
packages `gert` (for running git from R), `gh` (for using the GitHub
API from R), and `gitcreds` (for interacting with the git credential
store, easy authentication when pushing to GitHub). First, install
those packages. If you don't have a GitHub account, you can [sign up
for free](https://github.com/join). Then make sure to tell R what
name/email to use for git commits:

```r
gert::git_config_global_set("user.name", "<your_full_name>")
gert::git_config_global_set("user.email", "<your_email>")
```

You can then use `animint2pages(viz, "new_repo")` function to publish
your data `viz` to the `gh-pages` branch of
`your_github_username/new_repo` (note that `your_github_username` is
not specified in code, because `gitcreds` will get that information
from the git credential store).  It should print a message which tells
you the URL/link where your data viz will be accessible. It takes a
few minutes (usually not more than five) from the time you run
animint2pages, to the time the data viz is published for viewing on
GitHub Pages.

If you want to update an animint that has already been published on
GitHub Pages, you can simply run `animint2pages(updated_viz,
"existing_repo")`, which will update the `gh-pages` branch of the
specified repository. 

Beware that GitHub Pages imposes a limit of 100MB per file. For most
animints, this limit should not be a problem. If your data viz
contains a TSV file over 100MB, consider using the
[`chunk_vars`](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch06-other.html#geom-options)
option to break that TSV file into several smaller files.

## Organizing animints in a gallery {#gallery}

A gallery is a collection of meta-data about animints that have been
published to GitHub Pages. For example, the main animint gallery is
<https://animint.github.io/gallery/> which is a web page that has
links to various animints, organized into a table. 

* There is one row for each animint.
* The first column `viz.link` shows a screenshot of the animint, which
  links to the animint web page.
* The second column `repo.link` shows the name of the data viz (taken
  from the `title` option of the animint), and links to the GitHub
  repo's `gh-pages` branch.
* The third column `source.link` shows the last commit time of the
  animint, and links to the R source code used to create that animint
  (taken from the `source` option of the animint).

A gallery is defined as a github repo that should have two source
files in the `gh-pages` branch: 

* `repos.txt` (list of github repositories which contain animints, one
  owner/repo per line), and
* `index.Rmd` (source for web page with links to animints).

To add a new animint to the gallery which is published in the
`gh-pages` branch of `your_github_username/gallery_repo`, you can use
the following method:

* Create a new animint in R code, and make sure to define the `source`
  and `title` options.
* use `animint2pages(viz, "viz_repo")` to publish that animint to the
  `gh-pages` branch of `your_github_username/viz_repo`.
* take a screenshot of that animint, and commit/push that screenshot
  as a file named `Capture.PNG` (case-sensitive), in the `gh-pages`
  branch of `your_github_username/viz_repo`.
* add `your_github_username/viz_repo` to the `repos.txt` file in the
  `gh-pages` branch of `your_github_username/gallery_repo`.
* run the R code `animint2::update_gallery("path/to/gallery_repo")`
  (note that a clone of the gallery repo must be present on the system
  where you run this function, and the GitHub remote must be named
  `origin`). It will read `gallery_repo/repos.txt`, read meta-data
  (`title`, `source`, `Capture.PNG`) from each repo which is not
  already present in `gallery_repo/meta.csv`, write updated meta-data
  files to the gallery, render `gallery_repo/index.Rmd` to
  `gallery_repo/index.html`, commit, and push to `origin`.
* the updated gallery should be viewable on the web, at
  <https://your_github_username.github.io/gallery_repo> after a few
  minutes (usually no more than five).

## Chapter summary and exercises {#exercises}

This chapter explained how to share animints on the web.

Exercises:

* Create an animint using the options mentioned in this chapter:
  `out.dir` (name of directory to save animint on your computer),
  `source` (link to R source code used to create animint), `title`
  (description of animint).
* Use Netlify Drop to publish that animint on the web.
* Use `animint2pages` to publish that animint to a new GitHub
  repository. Create a screenshot and save it as `Capture.PNG` in the
  `gh-pages` branch of that repository. Add that repository to the
  main animint gallery
  [repos.txt](https://github.com/animint/gallery/blob/gh-pages/repos.txt)
  by submitting a Pull Request on GitHub.
* Create your own animint gallery repository, and add two or more of
  your own animints to that gallery.
  
Next, [Chapter 6](Ch06-other.html) explains the different options that
can be used to customize an animint.
