---
title: Other features
layout: default
output: bookdown::html_chapter
---

# Chapter 6, Other animint features

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch06-figures/")
```

This chapter explains several new features that animint introduces to
the grammar of graphics. After reading this chapter, you will
understand how to customize your animint graphics via

* the `href`, `tooltip`, `id`, `clickSelects.variable`,
  `clickSelects.value`, `showSelected.variable`, and
  `showSelected.value` aesthetics for observation-specific
  characteristics;
* the `chunk_vars` geom-specific option;
* plot-specific height and width options; and
* 

## Observation-specific options (new aesthetics)

### Review of previously introduced aesthetics

`aes(key)`

`aes(showSelected)`

`aes(clickSelects)`

### Hyperlinks using `aes(href)`

### Tooltips using `aes(tooltip)`

### CSS id using `aes(id)`

### Data-driven selector names using *.variable and *.value aesthetics

## The `chunk_vars` geom-specific compilation option

So far there is only one option for customization at the geom
level. The `chunk_vars` option should be specified as an argument to a
`geom_*` function, and its value should be a character vector of
selection variable names. When the data viz is compiled, these
selection variables will be used to break this geom's data table into
separate TSV files.

```{r}
data(breakpoints)
only.error <- subset(breakpoints$error,type=="E")
only.segments <- subset(only.error,bases.per.probe==bases.per.probe[1])
signal.colors <- c(estimate="#0adb0a",
                   latent="#0098ef")
tsvSizes <- function(segment.chunk.vars){
  viz <- 
    list(signal=ggplot()+
           geom_point(aes(position, signal, showSelected=bases.per.probe),
                      data=breakpoints$signals)+
           geom_line(aes(position, signal), colour=signal.colors[["latent"]],
                     data=breakpoints$imprecision)+
           geom_segment(aes(first.base, mean, xend=last.base, yend=mean,
                            showSelected=segments,
                            showSelected2=bases.per.probe),
                        colour=signal.colors[["estimate"]],
                        chunk_vars=segment.chunk.vars,
                        data=breakpoints$segments)+
           geom_vline(aes(xintercept=base,
                          showSelected=segments,
                          showSelected2=bases.per.probe),
                      colour=signal.colors[["estimate"]],
                      linetype="dashed",
                      data=breakpoints$breaks),
         error=ggplot()+
           geom_vline(aes(xintercept=segments, clickSelects=segments),
                      data=only.segments, lwd=17, alpha=1/2)+
           geom_line(aes(segments, error, group=bases.per.probe,
                         clickSelects=bases.per.probe),
                     data=only.error, lwd=4))
  info <- animint2dir(viz, open.browser=FALSE)
  cmd <- paste("du -ks", info$out.dir)
  kb.dt <- fread(cmd)
  setnames(kb.dt, c("kb", "dir"))
  tsv.vec <- Sys.glob(paste0(info$out.dir, "/*.tsv"))
  is.geom3 <- grepl("geom3", tsv.vec)
  data.frame(kb=kb.dt$kb, geom3.tsv=sum(is.geom3), other.tsv=sum(!is.geom3))
}

chunk_vars_list <- list(
  neither=c(),
  bases.per.probe=c("bases.per.probe"),
  segments=c("segments"),
  both=c("segments", "bases.per.probe"))

sizes.list <- lapply(chunk_vars_list, tsvSizes)
do.call(rbind, sizes.list)
```

## Plot-specific options

### Plot height and width

The `theme_animint` function is used to attach animint-specific
options to ggplot objects. There are options for specifying width and
height of the plot in pixels.

TODO example.

### Size scale in pixels

`scale_size_animint`

## Global data viz options

Global data viz options any named elements of the `viz` list that are
not ggplots.

### Review of previously introduced global options

duration, time, selector.types, first

### Web page title with the title option

### Show or hide selection menus with the selectize option

## Chapter summary and exercises

This chapter explain how to share animints on the web.

Exercises:

* Use `animint2gist` to publish a visualization of a data set from
  your domain of expertise. Post a link to the data viz (and your
  source code) on the
  [Animint Gallery wiki page](https://github.com/tdhock/animint/wiki/Gallery).
