---
title: Other features
layout: default
output: bookdown::html_chapter
---

# Chapter 6, Other animint features

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch06-figures/")
```

This chapter explains several new features that animint introduces to
the grammar of graphics. After reading this chapter, you will
understand how to customize your animint graphics via

* the `href`, `tooltip`, `id`, `clickSelects.variable`,
  `clickSelects.value`, `showSelected.variable`, and
  `showSelected.value` aesthetics for observation-specific
  characteristics;
* the `chunk_vars` geom-specific option;
* plot-specific legends and height/width options; and
* global data viz options.

## Observation-specific options (new aesthetics) {#new-aesthetics}

This section explains the new aesthetics that are recognized by
animint but not ggplot2.

### Review of previously introduced aesthetics {#review-aesthetics}

First we discuss the new aesthetics that we have already introduced in
previous chapters.

[Chapter 3](Ch03-showSelected.html#selecting-with-menus) introduced
`aes(showSelected)` for designating a geom which shows only
the selected subset of its data.

[Chapter 3](Ch03-showSelected.html#duration-key) also introduced
`aes(key)` to designate a variable to use for smooth transitions that
are interpretable.

[Chapter 4](Ch04-clickSelects.html#clickSelects-aes) introduced
`aes(clickSelects)` to designate a geom which can be clicked to change
a selection variable.

### Hyperlinks using `aes(href)` {#aes-href}

The code below uses animint to draw a map of the United States. 

```{r Ch06-viz-href}
library(animint)
USpolygons <- map_data("state")
viz.href <- list(
  map=ggplot()+
    ggtitle("click a state to read its Wikipedia page")+
    coord_equal()+
    geom_polygon(
      aes(x=long, y=lat, group=group,
          href=paste0("http://en.wikipedia.org/wiki/", region)),
      data=USpolygons, fill="black", colour="grey")
  )
structure(viz.href, class="animint")
```

Try clicking a state in the data viz above. You should see the
corresponding wikipedia page open in a new tab.

### Tooltips using `aes(tooltip)` {#aes-tooltip}

Tooltips are little windows of text information that appear when you
hover the cursor over something on the screen. In animint you can use
`aes(tooltip)` to designate the observation-specific message that 
appears. For example we use it to display the population and country
name in the scatterplot of the World Bank data below.

```{r Ch06-viz-tooltip}
data(WorldBank, package="animint")
WorldBank1975 <- subset(WorldBank, year == 1975)
viz.tooltip <- list(
  scatter=ggplot()+
  geom_point(
    aes(x=life.expectancy, y=fertility.rate,
        tooltip=paste(country, "population =", population)),
    WorldBank1975))
structure(viz.tooltip, class="animint")
```

Try hovering the cursor over one of the data points. You should see a
small box appear with the country name and population for that data
point.

Note that a tooltip of the form "variable value" is specified by
default for each geom with `aes(clickSelects)`. For example a geom
with `aes(clickSelects=year)` shows the default tooltip "year
1984" for an observation with year 1984. You can change this
default by explicitly specifying `aes(tooltip)`.

### HTML id attribute using `aes(id)` {#aes-id}

Since everything plotted by animint is rendered as an
[SVG](http://www.w3schools.com/html/html5_svg.asp) element in a web
page, you may want to specify a
[HTML id attribute](http://www.w3schools.com/tags/att_global_id.asp)
using `aes(id)` as below.

```{r Ch06-viz-id}
viz.id <- list(
  map=ggplot()+
    ggtitle("click a state to read its Wikipedia page")+
    coord_equal()+
    geom_polygon(
      aes(x=long, y=lat, group=group,
          id=gsub(" ", "_", paste(region, group))),
      data=USpolygons, fill="black", colour="grey")
  )
structure(viz.id, class="animint")
```

Note how `gsub` is used to convert spaces to underscores, since a
well-defined id must not include spaces. Note also that `paste` is
used to add a group number, since there may be more than one polygon
per state/region, and each id must be unique on a web page. The
animint developers use this feature
[for testing the animint JavaScript renderer code](https://github.com/tdhock/animint/wiki/Testing).

### Data-driven selector names using *.variable and *.value aesthetics {#variable-value}

Usually selector names are defined in `aes`, but that becomes
inconvenient if you have many selectors in your data viz. To
illustrate we consider a theoretical example (the code in this section
is not directly executable). Say you want to use 20 different selector
variable names, `selector1value` ... `selector20value`. The usual way
to define your data viz would be

```{r, eval=FALSE}
viz <- list(
  points=ggplot()+
    geom_point(aes(clickSelects=selector1value), data=data1)+
    ...
    geom_point(aes(clickSelects=selector20value),data=data20)
)
```

However that method is bad since it violates the DRY principle (Don't
Repeat Yourself). Another way to do that would be to use a for loop:

```{r, eval=FALSE}
viz <- list(points=ggplot())
for(selector.name in paste0("selector", 1:20, "value")){
  data.for.selector <- all.data.list[[selector.name]]
  viz$points <- viz$points +
    geom_point(aes_string(clickSelects=selector.name),
               data=data.for.selector)
}
```

That method is bad since it is slow to construct `viz`, and the
compiled viz potentially takes up a lot of disk space since there 
is at least one TSV file created for each `geom_point`. The preferable
method is to use `clickSelects.variable` and `clickSelects.value`
aesthetics:

```{r, eval=FALSE}
viz <- list(
  points=ggplot()+
    geom_point(aes(clickSelects.variable=selector.name,
                   clickSelects.value=selector.value),
               data=all.data)
)
```

The animint compiler looks through the data.frame `all.data` and
create selectors for each of the distinct values of
`all.data$selector.name`. Clicking one of the data points updates the
corresponding selector with the value indicated in
`all.data$selector.value`.

You can similarly use one geom with `showSelected.variable` and
`showSelected.value` instead of a bunch of different geoms with
`showSelected`.

The `*.variable` and `*.value` aesthetics are useful not only to avoid
repetition in the definition of the data viz, but also because they
are more computationally efficient. For a detailed example with
timings and disk space measurements, see
[Chapter 14](Ch14-variable-value.html).

## Geom options {#geom-options}

In animint there is only one option for customization at the geom
level: `chunk_vars`. 

### The `chunk_vars` geom-specific compilation option {#chunk-vars}

The `chunk_vars` option defines the selection variables that are used
to split the data set into separate chunks (TSV files) to
download. There is one TSV file created for each combination of values
of the `chunk_vars` variables. More selection variables specified in
`chunk_vars` means to split the data set into more TSV files, each of
a smaller size.

The `chunk_vars` option should be specified as an argument to a
`geom_*` function, and its value should be a character vector of
selection variable names. When `chunk_vars=character(0)`, a character
vector of length zero, all of the data is stored in a single TSV
file. When `chunk_vars` is set to all of the `showSelected` variable
names, then a TSV file is created for each combination of values of
those variables.

In general the animint compiler chooses a sensible default for
`chunk_vars`, but you may want to specify `chunk_vars` if the data viz
is loading slowly, or taking up too much space on disk. If the data
viz is loading slowly, you should add selection variables to
`chunk_vars` to reduce the size of the first TSV file to download. If
the data viz takes up too much space on disk, you should remove
selection variables from `chunk_vars` to decrease the number of TSV
files. Lots of small TSV files can take more disk space than a single
TSV file because some filesystems store a constant amount of metadata
for every file.

To illustrate the usage of `chunk_vars`, consider the following
visualization of the `breakpoints` data set.

TODO: sketch.

The sketch above consists of two plots. We begin by creating the plot
of error curves on the left.

```{r Ch06-errorPlot}
data(breakpoints, package="animint")
only.error <- subset(breakpoints$error, type=="E")
only.segments <- subset(only.error,bases.per.probe==bases.per.probe[1])
library(data.table)
fp.fn.names <- rbind(
  data.table(error.type="false positives", type="FP"),
  data.table(error.type="false negatives", type=c("I", "FN")))
setkey(fp.fn.names, type)
error.dt <- data.table(breakpoints$error)
setkey(error.dt, type)
error.type.dt <- error.dt[fp.fn.names]
fp.fn.dt <- error.type.dt[, list(
  error.value=sum(error)
), by=.(error.type, segments, bases.per.probe)]
errorPlot <- ggplot()+
  ggtitle("select data and segments")+
  theme_bw()+
  geom_tallrect(aes(xmin=segments-0.5, xmax=segments+0.5,
                    clickSelects=segments),
                data=only.segments,
                alpha=1/2)+
  geom_line(aes(segments, error.value, color=error.type,
                group=paste(bases.per.probe, error.type),
                showSelected=bases.per.probe),
            data=fp.fn.dt,
            size=5)+
  scale_color_manual(values=c(
    "false positives"="red", "false negatives"="blue"))+
  geom_line(aes(segments, error, group=bases.per.probe,
                clickSelects=bases.per.probe),
            data=only.error,
            size=4)+
  scale_x_continuous(breaks=c(1, 6, 10, 20))
errorPlot
```

The plot above includes a `geom_tallrect` with `clickSelects=segments`
and a `geom_line` with `clickSelects=bases.per.probe`. It will be used
to select the data and model in the plot below.

```{r Ch06-signalPlot}
signalPlot <- ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  geom_point(aes(position/1e5, signal,
                 showSelected=bases.per.probe),
             shape=1,
             data=breakpoints$signals)+
  geom_segment(aes(first.base/1e5, mean, xend=last.base/1e5, yend=mean,
                   showSelected=segments,
                   showSelected2=bases.per.probe),
               color="green",
               data=breakpoints$segments)
signalPlot+facet_grid(segments ~ bases.per.probe)
```

The non-interactive plot above has 80 facets, one for each combination
of the two `showSelected` variables, `bases.per.probe`, and
`segments`. Below we make an interactive version in which only one of
these facets will be shown.

```{r Ch06-viz-chunk-vars}
viz.chunk.vars <- list(
  error=errorPlot,
  signal=signalPlot+
  geom_vline(aes(xintercept=base/1e5,
                 showSelected=segments,
                 showSelected2=bases.per.probe),
             color="green",
             chunk_vars=character(),
             linetype="dashed",
             data=breakpoints$breaks)
)
structure(viz.chunk.vars, class="animint")
```

Click the "Show download status table" button, and you should see
counts of chunks (TSV files). Note that `geom6_vline_signal` has only
1 chunk, since `chunk_vars=character()` is specified for the
`geom_vline` in the R code above. If another value of `chunk_vars` was
specified, it would create a different number of TSV files, but the
appearance of the data viz should be the same.

Below we use the `du` command line program to determine the disk usage
of the data viz for different choices of `chunk_vars`.

```{r tsvSizes}
tsvSizes <- function(segment.chunk.vars){
  viz <- list(
    error=errorPlot,
    signal=signalPlot+
      geom_vline(aes(xintercept=base/1e5,
                     showSelected=segments,
                     showSelected2=bases.per.probe),
                 color="green",
                 chunk_vars=segment.chunk.vars,
                 linetype="dashed",
                 data=breakpoints$breaks)
  )
  info <- animint2dir(viz, open.browser=FALSE)
  cmd <- paste("du -ks", info$out.dir)
  kb.dt <- fread(cmd)
  setnames(kb.dt, c("kb", "dir"))
  tsv.vec <- Sys.glob(paste0(info$out.dir, "/*.tsv"))
  is.geom6 <- grepl("geom6", tsv.vec)
  data.frame(kb=kb.dt$kb, geom6.tsv=sum(is.geom6), other.tsv=sum(!is.geom6))
}
chunk_vars_list <- list(
  neither=c(),
  bases.per.probe=c("bases.per.probe"),
  segments=c("segments"),
  both=c("segments", "bases.per.probe"))
sizes.list <- lapply(chunk_vars_list, tsvSizes)
do.call(rbind, sizes.list)
```

The table above includes counts of kilobytes for the data viz, along
with counts of TSV files for `geom6_vline_signal` and the other
geoms. Note how the choice of `chunk_vars` affects the number of TSV
files and the disk space usage. Since `chunk_vars` was only specified
for `geom6_vline_signal`, the number of TSV files for the other geoms
does not change. When both `segments` and `bases.per.probe` are
specified for `chunk_vars`, there are 76 TSV files for
`geom6_vline_signal`, and the data viz takes 1084 kilobytes. In
contrast, `chunk_vars=character()` produces only one TSV file for
`geom6_vline_signal`, and the data viz takes only 820 kilobytes.

In conclusion, the geom-specific `chunk_vars` option defines the
number of TSV files created for each geom. When deciding the value of
`chunk_vars`, you should consider both disk usage and loading time. A
few large files take up less disk space but are slower to download
than many small files.

## Plot-specific options {#plot-options}

This section discusses options which are specific to one ggplot of a
data viz. The `theme_animint` function is used to attach animint
options to ggplot objects.

### Plot height and width {#plot-height-width}

The `width` and `height` options are for specifying the dimensions (in
pixels) of a ggplot rendered by animint. For example, consider the
following re-design of the plot of the United States:

```{r Ch06-viz-width-height}
viz.width.height <- list(
  map=ggplot()+
    theme_animint(width=750, height=500)+
    theme(axis.line=element_blank(),
          axis.text=element_blank(), 
          axis.ticks=element_blank(),
          axis.title=element_blank(),
          panel.border=element_blank(),
          panel.background=element_blank(),
          panel.grid.major=element_blank(),
          panel.grid.minor=element_blank())+
    geom_polygon(
      aes(x=long, y=lat, group=group),
      data=USpolygons, fill="black", colour="grey")
)
structure(viz.width.height, class="animint")
```

Note that the plot above was rendered with a width of 750 pixels and a
height of 500 pixels, due to the `theme_animint` options. If either of
these options is not specified for any ggplot, then animint uses a
default of 400 pixels.

Also note that `theme` was used to specify several blank
elements. This has the effect of removing the axes and background, and
is generally useful for rendering maps.

### Size scale in pixels {#scale-size-animint}

The `scale_size_animint` scale should be used in all ggplots where you
specify `aes(size)`. To see why, consider the following examples.

```{r Ch06-viz-scale-size}
scatter1975 <- ggplot()+
  geom_point(
    aes(x=life.expectancy, y=fertility.rate, size=population),
    WorldBank1975,
    shape=21,
    color="red",
    fill="black")
viz.scale.size <- list(
  ggplotDefault=scatter1975+
    ggtitle("no scale specified"),
  animintDefault=scatter1975+
    ggtitle("scale_size_animint()")+
    scale_size_animint(),
  animintOptions=scatter1975+
    ggtitle("scale_size_animint(pixel.range, breaks)")+
    scale_size_animint(pixel.range=c(5, 15), breaks=10^(10:1)))
structure(viz.scale.size, class="animint")
```

The first ggplot above has no scale specified, so it uses the default
ggplot2 scale, which has two problems. The first problem is that it
seems that all countries have about the same size except the two
really big countries. That problem can be fixed by simply adding
`scale_size_animint()` to the ggplot, which results in the second plot
above. However, a second problem is that the legend entries do not
show the full range of the data. That problem is fixed in the third
plot above, by manually specifying the `breaks` to use for legend
entries. Note that the `pixel.range` argument can also be used to
specify the radius of the largest and smallest circles.

## Global data viz options {#global-options}

Global data viz options any named elements of the `viz` list that are
not ggplots.

### Review of previously introduced global options {#review-global}

[Chapter 3](Ch03-showSelected.html#duration-key) introduced the
`duration` option for specifying the duration of smooth transitions.

[Chapter 3](Ch03-showSelected.html#animation-time) introduced the
`time` option for specifying a selection variable which is
automatically updated (animation).

[Chapter 4](Ch04-clickSelects.html#selector-types-first) introduced
the `selector.types` option for specifying multiple selection
variables, and the `first` option for specifying the selection when
the data viz is first rendered.

### Web page title with the title option {#title-option}

The `title` option should be a character string, and will be used to
set the `<title>` element of the web page. It does not make sense to
use the `title` option in an Rmd document such as this page. You can
and should use it with `animint2dir` or `animint2gist`, as in the code
below.

```{r Ch06-viz-title}
viz.title <- viz.scale.size
viz.title$title <- "Several size scales"
animint2dir(viz.title, "Ch06-title")
```

Note that `viz.scale.size` already has three ggplots, each with a
`ggtitle`. Adding the global `title` option has the effect of defining
a title for [the web page](Ch06-title/index.html).

### Show or hide selection menus with the selectize option {#selectize-option}

The selectize option should be a named list of logical values. Names
should be selector variables, and values should indicate whether or
not you would like to render a selection menu via
[selectize.js](http://selectize.github.io/selectize.js/). By default,
animint will render a selection menu for every selection variable,
with two exceptions:

* selection variables that are defined using
  [.variable and .value aesthetics](Ch06-other.html#variable-value).
* selection variables that have a lot of values (they are slow to
  render).
  
These defaults should work well for the vast majority of animints. For
those who are interested to see an example of how the `selectize`
option works, please see the
[PredictedPeaks test in the animint source code](https://github.com/tdhock/animint/blob/master/tests/testthat/test-renderer2-PredictedPeaks.R).
  
## Chapter summary and exercises {#exercises}

This chapter explained several options for customizing animints at the
observation, geom, plot, and global level.

Exercises:

* Use `animint2gist` to publish a visualization of a data set from
  your domain of expertise. Post the source code for your data viz in
  a GitHub repository. Post links to your data viz and its source code
  on the
  [Animint Gallery wiki page](https://github.com/tdhock/animint/wiki/Gallery).
* Create other versions of `viz.chunk.vars` with different values of
  `chunk_vars` for the `geom_point` and `geom_segment`. How does the
  choice of `chunk_vars` affect the appearance of the visualization?
  The disk space? The loading time?
