---
title: World Bank data viz
layout: default
output: bookdown::html_chapter
---

# Chapter 8, World Bank data viz

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch08-figures/")
```

In this chapter we will explore several data visualizations of the
World Bank data set.

Chapter outline:

* We begin by loading the World Bank data set and defining some helper
  functions for creating a multi-panel ggplot with several geoms.
* We then create a time series plot for life expectancy.
* We then add a scatterplot of life expectancy versus fertility rate
  as a second panel.
* We then add a third panel with a time series for fertility rate.

## Load data and define helper functions {#load}

First we load the WorldBank data set, and consider only the subset
which has both non-missing values for both life.expectancy and
fertility.rate.

```{r}
library(animint)
data(WorldBank)
library(data.table)
not.na <- data.table(
  WorldBank)[!(is.na(life.expectancy) | is.na(fertility.rate)), ]
```

We will also be plotting the population variable using a size
legend. Before plotting, we will make sure that none of the values are
missing.

```{r}
not.na[is.na(not.na$population), ]
```

The table above shows that there are three rows with missing values
for the population variable. They are for the country Kuwait during
1992-1994. The table below shows the data
from the neighboring years, 1991-1995.

```{r}
not.na[country == "Kuwait" & 1991 <= year & year <= 1995, ]
```

The table above shows that the population of Kuwait decreased over the
period 1991-1995, consistent with the Gulf War of that time period. We
fill in those missing values below.

```{r}
not.na[is.na(population), population := 1700000]
not.na[country == "Kuwait" & 1991 <= year & year <= 1995, ]
```

Next, we define the following helper function, which will be used to
add columns to data sets in order to assign geoms to facets. 

```{r}
FACETS <- function(df, top, side){
  data.frame(df,
             top=factor(top, c("Fertility rate", "Years")),
             side=factor(side, c("Years", "Life expectancy")))
}
```

Note that the factor levels will specify the order of the facets in
the ggplot. This is an example of the
[addColumn then facet idiom](Ch17-appendix.html#addColumn-then-facet).
Below, we define three helper functions, one for each facet.

```{r}
TS.RIGHT <- function(df)FACETS(df, "Years", "Life expectancy")
SCATTER <- function(df)FACETS(df, "Fertility rate", "Life expectancy")
TS.ABOVE <- function(df)FACETS(df, "Fertility rate", "Years")
```

## First time series plot {#first-ts}

First we define a data set with one row for each year, which we will
use for selecting years using a `geom_tallrect` in the background.

```{r}
years <- unique(not.na[, .(year)])
```

We define the ggplot with a `geom_tallrect` in the background, and a
`geom_line` for the time series.

```{r}
ts.right <- ggplot()+
  geom_tallrect(aes(xmin=year-1/2, xmax=year+1/2,
                    clickSelects=year),
                data=TS.RIGHT(years), alpha=1/2)+
  geom_line(aes(year, life.expectancy, group=country, colour=region,
                clickSelects=country),
            data=TS.RIGHT(not.na), size=4, alpha=3/5)
ts.right
```

Note that we specified `clickSelects=year` so that clicking a tallrect
will change the selected year, and `clickSelects=country` so that
clicking a line will select or de-select a country. Also note that we
used `TS.RIGHT` to specify columns that we will use in the facet
specification (next section).

## Add a scatterplot facet {#add-scatter}

We begin by simply adding facets to the previous time series plot.

```{r}
ts.facet <- ts.right+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(side ~ top, scales="free")+
  xlab("")+
  ylab("")
ts.facet
```

We set the `panel.margin` to 0, which is always a good idea to
[save space in a ggplot with facets](Ch17-appendix.html#space-saving-facets).
We use `scales="free"` and hide the axis labels, in an example of the
[addColumn then facet idiom](Ch17-appendix.html#addColumn-then-facet). Instead,
we use the facet label to show the variable encoded on each
axis. Below, we add a scatterplot facet with a point for each year and
country.

```{r}
ts.scatter <- ts.facet+
  geom_point(aes(fertility.rate, life.expectancy, clickSelects=country,
                 showSelected=year, colour=region, size=population,
                 key=country), # key aesthetic for animated transitions!
             data=SCATTER(not.na))+
  scale_size_animint(pixel.range=c(2, 20), breaks=10^(9:5))
ts.scatter
```

Note how we use `scale_size_animint` to specify the range of sizes in
pixels, and the breaks in the legend. Also note that we use `SCATTER`
to specify `top` and `side` columns which are used in the facet
specification. We also render this ggplot interactively below.

```{r Ch08-viz-ts-scatter}
viz.ts.scatter <- list(TS=ts.scatter)
structure(viz.ts.scatter, class="animint")
```

Note that single selection is used by default for both year and
country.

## Adding another time series facet {#add-ts}

Below we add widerects for selecting years, and paths for showing
fertility rate.

```{r}
scatter.both <- ts.scatter+
  geom_widerect(aes(ymin=year-1/2, ymax=year+1/2,
                    clickSelects=year),
                data=TS.ABOVE(years), alpha=1/2)+
  geom_path(aes(fertility.rate, year, group=country, colour=region,
                clickSelects=country),
            data=TS.ABOVE(not.na), size=4, alpha=3/5)
scatter.both
```

Note that `TS.ABOVE` was used to specify facet columns `top` and
`side`. We render an interactive version below.

```{r Ch08-viz-scatter-both}
viz.scatter.both <- list(
  title="World Bank data (multiple selection, facets)",
  scatterBoth=scatter.both+
    theme_animint(width=1000, height=800),
  duration=list(year=1000),
  time=list(variable="year", ms=3000),
  first=list(year=1975, country=c("United States", "Vietnam")),
  selector.types=list(country="multiple"))
structure(viz.scatter.both, class="animint")
```

## Chapter summary and exercises {#exercises}

We showed how to create a multi-layer, multi-panel (but single-plot)
visualization of the World Bank data.

Exercises:

* Add a points on each time series plot, with size proportional to
  population as in the scatterplot. The points should appear only when
  the country is selected, and clicking the points should de-select
  that country.
* Add text labels to the time series plot on the right, with names for
  each country. Each label should appear only when the country is
  selected, and should disappear after clicking on the label.
* Add a text label to the scatterplot to indicate the selected year.
* Add text labels to the scatterplot, with names for each
  country. Each label should appear only when the country is selected,
  and should disappear after clicking on the label.

Next, [Chapter 9](Ch09-Montreal-bikes.html) explains how to visualize the
Montreal bike data set.
