---
title: Montreal bikes data viz
layout: default
output: bookdown::html_chapter
---

# Chapter 9, Montreal bikes data viz

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch09-figures/")
```

In this chapter we will explore several data visualizations of the
Montreal bike data set.

Chapter outline:

* We begin with some static data visualizations.
* We create an interactive visualization of accident frequency over
  time.
* We create a interactive data viz with four plots, showing monthly
  accident trends, daily details, and a map of counter locations.

## Static figures {#static}

We begin by loading the `montreal.bikes` data set, which is not
available in the CRAN release of animint2, in order to save space on
CRAN. Therefore to access this data set, you will need to install
animint2 from GitHub:

```{r ghinstall}
tryCatch({
  data(montreal.bikes, package="animint2")
}, warning=function(w){
  devtools::install_github("tdhock/animint2")
})
```

We begin by examining the accidents data table.

```{r accidents}
library(animint2)
data(montreal.bikes) #only present if installed from github
Sys.setlocale(locale="en_US.UTF-8")
if(! "Brébeuf" %in% montreal.bikes$counter.counts$location){
  Encoding(levels(montreal.bikes$counter.counts$location)) <- "UTF-8"
  print(table(montreal.bikes$counter.counts$location))
}

library(data.table)
accidents.dt <- data.table(montreal.bikes$accidents)
str(accidents.dt)
```

Each accident has data about its date, time, location, and counts of
death and slight/severe injury. Some of the values are in French
(e.g. position Voie de circulation, location En intersection, etc).

We calculate the time period of the accidents below.

```{r timeRange}
accidents.dt[, date.POSIXct := suppressWarnings(strptime(
  date.str, "%Y-%m-%d"))]
accidents.dt[, month.str := strftime(date.POSIXct, "%Y-%m")]
range(accidents.dt$month.str)
```

Below we also compute the range of months for the bike counter data table.

```{r counterRange}
(counts.dt <- data.table(montreal.bikes$counter.counts))
counts.dt[, month.str := strftime(date, "%Y-%m")]
range(counts.dt$month.str)
```

The bike counts are time series data which we visualize below.

```{r counterviz}
counts.dt[, loc.lines := gsub("[- _]", "\n", location)]
ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(loc.lines ~ .)+
  geom_point(aes(
    date, count, color=count==0),
    shape=21,
    data=counts.dt)+
  scale_color_manual(values=c("TRUE"="grey", "FALSE"="black"))
```

Plotting with `geom_point` makes it easy to see the difference between
zeros and missing values.

We will compute a summary of all accidents per month in this time
period, so we first create a data table for each month below. (and
make sure to set the locale to C for English month names)

```{r months}
uniq.month.vec <- unique(c(
  accidents.dt$month.str,
  counts.dt$month.str))
months <- data.table(month.str=uniq.month.vec)
months[, month01.str := paste0(month.str, "-01")]
months[, month01.POSIXct := suppressWarnings(strptime(month01.str, "%Y-%m-%d"))]
one.day <- 60 * 60 * 24
months[, next.POSIXct := month01.POSIXct + one.day * 31]
months[, next01.str := paste0(strftime(next.POSIXct, "%Y-%m"), "-01")]
months[, next01.POSIXct := suppressWarnings(strptime(
  next01.str, "%Y-%m-%d"))]
months[, month.str := strftime(month01.POSIXct, "%B %Y")]
month.levs <- months[order(month01.POSIXct), month.str]
months[, month := factor(month.str, month.levs)]
```

Note that we created a `month` column which is a factor ordered by
`month.levs`.

```{r}
accidents.dt[, month.text := strftime(date.POSIXct, "%B %Y")]
accidents.dt[, month := factor(month.text, month.levs)]
accidents.dt[, month.POSIXct := suppressWarnings(
  strptime(paste0(month.str, "-15"), "%Y-%m-%d"))]
stopifnot(!is.na(accidents.dt$month.POSIXct))
accidents.per.month <- accidents.dt[, list(
  total.accidents=.N,
  total.people=sum(deaths+people.severely.injured+people.slightly.injured),
  deaths=sum(deaths),
  people.severely.injured=sum(people.severely.injured),
  people.slightly.injured=sum(people.slightly.injured)
), by=.(month, month.str, month.text, month.POSIXct)]
accidents.per.month[, next.POSIXct := month.POSIXct + one.day * 30]
accidents.per.month[, month01.str := paste0(
  strftime(month.POSIXct, "%Y-%m"), "-01")]
accidents.per.month[, month01.POSIXct := suppressWarnings(
  strptime(month01.str, "%Y-%m-%d"))]
accidents.per.month[, next01.str := paste0(
  strftime(next.POSIXct, "%Y-%m"), "-01")]
accidents.per.month[, next01.POSIXct := suppressWarnings(
  strptime(next01.str, "%Y-%m-%d"))]
```

We plot the accidents per month below.

```{r accidentsPerMonth}
accidents.tall <- melt(
  accidents.per.month,
  measure.vars=c(
    "deaths", "people.severely.injured", "people.slightly.injured"),
  variable.name="severity",
  value.name="people")
severity.colors <- c(
    "people.slightly.injured"="#FEE0D2",#lite red
    "people.severely.injured"="#FB6A4A",
    deaths="#A50F15")#dark red
ggplot()+
  theme_bw()+
  geom_bar(aes(
    month.POSIXct, people, fill=severity),
    stat="identity",
    data=accidents.tall)+
  scale_fill_manual(values=severity.colors)
```

```{r}
accidents.dt[, severity.str := ifelse(
  0 < deaths, "deaths", ifelse(
    0 < people.severely.injured,
    "people.severely.injured",
    "people.slightly.injured"))]
accidents.dt[, severity := factor(severity.str, names(severity.colors))]
counts.dt[, month.text := strftime(date, "%B %Y")]
counts.dt[, day.of.the.month := as.integer(strftime(date, "%d"))]
counts.dt[, month := factor(month.text, month.levs)]
counts.dt[, month.POSIXct := suppressWarnings(
  strptime(paste0(month.str, "-15"), "%Y-%m-%d"))]
counts.per.month <- counts.dt[, list(
  days=.N,
  mean.per.day=mean(count),
  count=sum(count)
), by=.(location, month, month.str, month.POSIXct)][0 < count,]
counts.per.month[, month01.str := paste0(month.str, "-01")]
counts.per.month[, month01.POSIXct := suppressWarnings(strptime(month01.str, "%Y-%m-%d"))]
counts.per.month[, next.POSIXct := month01.POSIXct + one.day * 31]
counts.per.month[, next01.str := paste0(strftime(next.POSIXct, "%Y-%m"), "-01")]
counts.per.month[, next01.POSIXct := suppressWarnings(strptime(next01.str, "%Y-%m-%d"))]
counts.per.month[, days.in.month := round(next01.POSIXct - month01.POSIXct)]
counts.per.month[days < days.in.month, {
  list(location, month, days, days.in.month)
}]
```

As shown above, some months do not have observations for all days.

## Interactive viz of accident frequency {#regression}

```{r Ch09-viz-regression}
complete.months <- counts.per.month[days == days.in.month]
month.labels <- counts.per.month[, {
  .SD[which.max(count), ]
}, by=location]
day.labels <- counts.dt[, {
  .SD[which.max(count), ]
}, by=.(location, month)]
city.wide.cyclists <- counts.per.month[0 < count, list(
  locations=.N,
  count=sum(count)
), by=.(month, month.str, month.POSIXct)]
city.wide.cyclists[, month01.str := paste0(month.str, "-01")]
city.wide.cyclists[, month01.POSIXct := suppressWarnings(
  strptime(month01.str, "%Y-%m-%d"))]
city.wide.cyclists[, next.POSIXct := month01.POSIXct + one.day * 31]
city.wide.cyclists[, next01.str := paste0(
  strftime(next.POSIXct, "%Y-%m"), "-01")]
city.wide.cyclists[, next01.POSIXct := suppressWarnings(
  strptime(next01.str, "%Y-%m-%d"))]
month.str.vec <- strftime(seq(
  strptime("2012-01-15", "%Y-%m-%d"),
  strptime("2013-01-15", "%Y-%m-%d"),
  by="month"), "%Y-%m")
city.wide.complete <- complete.months[0 < count, list(
  locations=.N,
  count=sum(count)
), by=.(month, month.str, month.POSIXct)]
city.wide.complete[, month01.str := paste0(month.str, "-01")]
setkey(city.wide.complete, month.str)
scatter.cyclists <- city.wide.complete[month.str.vec]
scatter.accidents <- accidents.per.month[scatter.cyclists, on=list(month.str)]
scatter.not.na <- scatter.accidents[!is.na(locations),]
scatter.max <- scatter.not.na[locations==max(locations),]
fit <- lm(total.accidents ~ count - 1, scatter.max)
scatter.max[, mean(total.accidents/count)]
scatter.max[, pred.accidents := predict(fit)]
animint(
  regression=ggplot()+
    theme_bw()+
    ggtitle("Numbers of accidents and cyclists")+
    geom_line(aes(
      count, pred.accidents),
      color="grey",
      data=scatter.max)+
    geom_point(aes(
      count, total.accidents),
      shape=1,
      clickSelects="month",
      size=5,
      alpha=0.75,
      data=scatter.max)+
    ylab("Total bike accidents (all Montreal locations)")+
    xlab("Total cyclists (all Montreal locations)"),
  timeSeries=ggplot()+
    theme_bw()+
    ggtitle("Time series of accident frequency")+
    xlab("Month")+
    geom_point(aes(
      month.POSIXct, total.accidents/count),
      clickSelects="month",             
      size=5,
      alpha=0.75,
      data=scatter.max))
```

The data viz above shows two data visualizations of city-wide accident
frequency over time. The plot on the left shows that the number of
accidents grows with the number of cyclists. The plot on the right
shows the frequency of accidents over time.

## Interactive viz with map and details {#details}

The plot below is a dotplot of accidents for each month. Each dot
represents one person who got in an accident.

```{r}
accidents.cumsum <- accidents.dt[order(date.POSIXct, month, severity)]
accidents.cumsum[, accident.i := seq_along(severity), by=.(date.POSIXct, month)]
accidents.cumsum[, day.of.the.month := as.integer(strftime(date.POSIXct, "%d"))]
ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "cm"))+
  facet_wrap("month")+
  geom_text(aes(15, 25, label=month), data=accidents.per.month)+
  scale_fill_manual(values=severity.colors)+
  scale_x_continuous("day of the month", breaks=c(1, 10, 20, 30))+
  geom_point(aes(
    day.of.the.month, accident.i, fill=severity),
    shape=21,
    data=accidents.cumsum)
```

```{r}
counter.locations <- data.table(montreal.bikes$counter.locations)
counter.locations[, lon := coord_X]
counter.locations[, lat := coord_Y]
loc.name.code <- c(
  "Berri1"="Berri",
  "Brebeuf"="Brébeuf",
  CSC="Côte-Sainte-Catherine",
  "Maisonneuve_1"="Maisonneuve 1",
  "Maisonneuve_2"="Maisonneuve 2",
  "Parc"="du Parc",
  PierDup="Pierre-Dupuy",
  "Rachel/Papineau"="Rachel",
  "Saint-Urbain"="Saint-Urbain",
  "Totem_Laurier"="Totem_Laurier")
counter.locations[, location := loc.name.code[nom_comptage]]
velo.counts <- table(counts.dt$location)
(show.locations <- counter.locations[names(velo.counts), on=list(location)])
```

The counter locations above will be plotted below. Note that we use
`showSelected=month` and `clickSelects=location`.

```{r}
map.lim <- show.locations[, list(
  range.lat=range(lat),
  range.lon=range(lon)
)]
diff.vec <- sapply(map.lim, diff)
diff.mat <- c(-1, 1) * matrix(diff.vec, 2, 2, byrow=TRUE)
scale.mat <- as.matrix(map.lim) + diff.mat
location.colors <-
  c("#8DD3C7", "#FFFFB3", "#BEBADA", "#FB8072", "#80B1D3", "#FDB462", 
    "#B3DE69", "#FCCDE5", "#D9D9D9", "#BC80BD", "#CCEBC5", "#FFED6F")
names(location.colors) <- show.locations$location
counts.per.month.loc <- counts.per.month[show.locations, on=list(location)]
bike.paths <- data.table(montreal.bikes$path.locations)
some.paths <- bike.paths[
  scale.mat[1, "range.lat"] < lat &
    scale.mat[1, "range.lon"] < lon &
    lat < scale.mat[2, "range.lat"] &
    lon < scale.mat[2, "range.lon"]]
mtl.map <- ggplot()+
  theme_bw()+
  theme(
    panel.margin=grid::unit(0, "lines"),
    axis.line=element_blank(), axis.text=element_blank(), 
    axis.ticks=element_blank(), axis.title=element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank())+
  coord_equal(xlim=map.lim$range.lon, ylim=map.lim$range.lat)+
  scale_color_manual(values=location.colors)+
  scale_x_continuous(limits=scale.mat[, "range.lon"])+
  scale_y_continuous(limits=scale.mat[, "range.lat"])+
  geom_path(aes(
    lon, lat,
    tooltip=TYPE_VOIE,
    group=paste(feature.i, path.i)),
    color="grey",
    data=some.paths)+
  guides(color="none")+
  geom_text(aes(
    lon, lat,
    label=location),
    clickSelects="location",
    data=show.locations)
mtl.map
```

The plot below shows the time period that each counter was in
operation. Note that we use `geom_tallrect` with `clickSelects` to
select the month.

```{r}
location.ranges <- counts.per.month[0 < count, list(
  min=min(month.POSIXct),
  max=max(month.POSIXct)
), by=location]
accidents.range <- accidents.dt[, data.table(
  location="accidents",
  min=min(date.POSIXct),
  max=max(date.POSIXct))]
MonthSummary <- ggplot()+
  theme_bw()+
  theme_animint(width=400, height=200)+
  xlab("range of dates in data")+
  ylab("data type")+
  scale_color_manual(values=location.colors)+
  guides(color="none")+
  geom_segment(aes(
    min, location,
    xend=max, yend=location,
    color=location),
    clickSelects="location",
    data=location.ranges, alpha=3/4, size=10)+
  geom_segment(aes(
    min, location,
    xend=max, yend=location),
    color=severity.colors[["deaths"]],
    data=accidents.range,
    size=10)
print(MonthSummary)
```

The plot below shows the bike counts at each location and day.

```{r}
dates <- counts.dt[, list(
  locations=sum(!is.na(count))
), by=list(date)][0 < locations]
dates[, min.date := date-one.day/2]
dates[, max.date := date+one.day/2]
location.labels <- counts.dt[, .SD[which.max(count)], by=list(location)]
TimeSeries <- ggplot()+
  theme_bw()+
  geom_tallrect(aes(xmin=date-one.day/2, xmax=date+one.day/2,
                    clickSelects=date),
                data=dates, alpha=1/2)+
  geom_line(aes(date, count, group=location,
                showSelected=location,
                clickSelects=location),
            data=counts.dt)+
  scale_color_manual(values=location.colors)+
  geom_point(aes(date, count, color=location,
                 showSelected=location,
                 clickSelects=location),
             data=counts.dt)+
  geom_text(aes(date, count+200, color=location, label=location,
                showSelected=location,
                clickSelects=location),
            data=location.labels)
print(TimeSeries)
```

The plot below shows the same data but for each month.

```{r}
MonthSeries <- ggplot()+
  guides(color="none", fill="none")+
  theme_bw()+
  geom_tallrect(aes(
    xmin=month01.POSIXct, xmax=next01.POSIXct),
    clickSelects="month",    
    data=months,
    alpha=1/2)+
  geom_line(aes(
    month.POSIXct, count, group=location,
    color=location),
    showSelected="location",
    clickSelects="location",
    data=counts.per.month)+
  scale_color_manual(values=location.colors)+
  scale_fill_manual(values=location.colors)+
  xlab("month")+
  ylab("bike counts per month")+
  geom_point(aes(
    month.POSIXct, count, fill=location,
    tooltip=paste(
      count, "bikers counted at",
      location, "in", month)),
    showSelected="location",
    clickSelects="location",
    size=5,
    color="black",
    data=counts.per.month)+
  geom_text(aes(
    month.POSIXct, count+5000, color=location, label=location),
    showSelected="location",
    clickSelects="location",
    data=month.labels)
print(MonthSeries)
```

```{r}
counter.title <- "mean cyclists per day"
accidents.title <- "city-wide accidents"
MonthFacet <- ggplot()+
  ggtitle("All data, select month")+
  guides(color="none", fill="none")+
  theme_bw()+
  facet_grid(facet ~ ., scales="free")+
  theme(panel.margin=grid::unit(0, "lines"))+
  geom_tallrect(aes(
    xmin=month01.POSIXct, xmax=next01.POSIXct),
    clickSelects="month",
    data=data.table(
      city.wide.cyclists,
      facet=counter.title),
    alpha=1/2)+
  geom_line(aes(
    month.POSIXct, mean.per.day, group=location,
    color=location),
    showSelected="location",
    clickSelects="location",
    data=data.table(counts.per.month, facet=counter.title))+
  scale_color_manual(values=location.colors)+
  xlab("month")+
  ylab("")+
  geom_point(aes(
    month.POSIXct, mean.per.day, color=location,
    tooltip=paste(
      count, "cyclists counted at",
      location, "in",
      days, "days of", month,
      sprintf("(mean %d cyclists/day)", as.integer(mean.per.day)))),
    showSelected="location",
    clickSelects="location",
    size=5,
    fill="grey",
    data=data.table(counts.per.month, facet=counter.title))+
  geom_text(aes(
    month.POSIXct, mean.per.day+300, color=location, label=location),
    showSelected="location",
    clickSelects="location",
    data=data.table(month.labels, facet=counter.title))+
  scale_fill_manual(values=severity.colors, breaks=names(severity.colors))+
  geom_bar(aes(
    month.POSIXct, people,
    fill=severity),
    showSelected="severity",
    stat="identity",
    position="identity",
    color=NA,
    data=data.table(accidents.tall, facet=accidents.title))+
  geom_tallrect(aes(
    xmin=month01.POSIXct, xmax=next01.POSIXct,
    tooltip=paste(
      ifelse(deaths==0, "",
      ifelse(deaths==1,
             "1 death,",
             paste(deaths, "deaths,"))),
      ifelse(people.severely.injured==0, "",
      ifelse(people.severely.injured==1,
             "1 person severely injured,",
             paste(people.severely.injured,
                   "people severely injured,"))),
      people.slightly.injured,
                      "people slightly injured in",
                      month)),
    clickSelects="month",
    alpha=0.5,
    data=data.table(accidents.per.month,
                    facet=accidents.title))
MonthFacet
```

```{r}
days.dt <- data.table(day.POSIXct=with(months, seq(
  min(month01.POSIXct),
  max(next01.POSIXct),
  by="day")))
days.dt[, day.of.the.week := strftime(day.POSIXct, "%a")]
## The following only works in locales with English days of the week.
weekend.dt <- days.dt[day.of.the.week %in% c("Sat", "Sun"),]
weekend.dt[, month.text := strftime(day.POSIXct, "%B %Y")]
weekend.dt[, month := factor(month.text, month.levs)]
weekend.dt[, day.of.the.month := as.integer(strftime(day.POSIXct, "%d"))]
counter.title <- "cyclists per day"
DaysFacet <- ggplot()+
  ggtitle("Selected month (weekends in grey)")+
  geom_tallrect(aes(
    xmin=day.of.the.month-0.5, xmax=day.of.the.month+0.5,
    key=paste(day.POSIXct)),
    showSelected="month",
    color="grey",
    data=weekend.dt)+
  guides(color="none")+
  theme_bw()+
  facet_grid(facet ~ ., scales="free")+
  theme(panel.margin=grid::unit(0, "lines"))+
  geom_line(aes(
    day.of.the.month, count, group=location,
    key=location,
    color=location),
    showSelected=c("location", "month"),
    clickSelects="location",
    chunk_vars=c("month"),
    data=data.table(counts.dt, facet=counter.title))+
  scale_color_manual(values=location.colors)+
  ylab("")+
  geom_point(aes(
    day.of.the.month, count, color=location,
    key=paste(day.of.the.month, location),
    tooltip=paste(
      count, "cyclists counted at",
      location, "on",
      date)),
    showSelected=c("location", "month"),
    clickSelects="location",
    size=5,
    chunk_vars=c("month"),
    fill="grey",
    data=data.table(counts.dt, facet=counter.title))+
  scale_fill_manual(values=severity.colors, breaks=names(severity.colors))+
  geom_text(aes(
    15, 23, label=month, key=1),
    showSelected="month",
    data=data.table(months, facet=accidents.title))+
  scale_x_continuous("day of the month", breaks=c(1, 10, 20, 30))+
  geom_text(aes(
    day.of.the.month, count+500, color=location, label=location,
    key=location),
    showSelected=c("location", "month"),
    clickSelects="location",
    data=data.table(day.labels, facet=counter.title))+
  geom_point(aes(
    day.of.the.month, accident.i,
    key=paste(date.str, accident.i),
    tooltip=paste(
      ifelse(deaths==0, "",
      ifelse(deaths==1,
             "1 death,",
             paste(deaths, "deaths,"))),
      ifelse(people.severely.injured==0, "",
      ifelse(people.severely.injured==1,
             "1 person severely injured,",
             paste(people.severely.injured,
                   "people severely injured,"))),
      people.slightly.injured,
      "people slightly injured at",
      ifelse(is.na(street.number), "", street.number),
      street, "/", cross.street,
      date.str, time.str),
    fill=severity),
    showSelected="month",
    size=4,
    chunk_vars=c("month"),
    data=data.table(accidents.cumsum, facet=accidents.title))
DaysFacet
```

```{r Ch09-viz-mtl}
animint(
  MonthFacet,
  DaysFacet,
  MonthSummary,
  selector.types=list(severity="multiple"),
  duration=list(month=2000),
  first=list(
    location="Berri",
    month="September 2012"),
  time=list(variable="month", ms=5000))#buggy.
```

## Chapter summary and exercises {#exercises}

Exercises:

* Change location to a multiple selection variable.
* Add a plot for the map to the data viz.
* On the map, draw a circle for each location, with size that changes
  based on the `count` of the accidents in the currently selected
  `month`.
* On the `MonthSummary` plot, add a background
  rectangle that can be used to select the `month`.
* Remove the `MonthSummary` plot and add a similar visualization as a
  third panel in the `MonthFacet` plot.

Next, [Chapter 10](Ch10-nearest-neighbors.html) explains how to visualize the
K-Nearest-Neighbors machine learning model.
