---
title: Support Vector Machines
layout: default
output: bookdown::html_chapter
---

# Chapter 12, Support Vector Machines

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch12-figures/")
```

This goal of this chapter is to create an interactive data
visualization that explains the
[Support Vector Machine](https://en.wikipedia.org/wiki/Support_vector_machine),
a machine learning model for binary classification.

Chapter summary:

* We begin by simulating some data for binary classification in two
  dimensions, and making some static plots.
* In the second section, we make an interactive data visualization to
  show how the linear Support Vector Machine decision boundary changes
  as a function of the cost hyper-parameter.
* In the last section, we make an interactive data visualization to
  show how the decision boundary of the polynomial kernel Support
  Vector Machine changes as a function of the two hyper-parameters
  (cost and degree).

## Generate and plot some data {#generate}

We begin by generating two input features, `x1` and `x2`.

```{r}
library(data.table)
N <- 50
set.seed(1)
getInput <- function(){
  c(#rnorm(N, sd=0.3),
    runif(N, -1, 1),
    runif(N, -1, 1)    
    )
}
data.dt <- data.table(
  x1=getInput(),
  x2=getInput())
library(animint2)
ggplot()+
  geom_point(aes(
    x1, x2),
    data=data.dt)
```

The plot below shows the same data, after computing two additional
input features (the squares of the original two inputs).

```{r}
data.dt[, x1.sq := x1^2 ]
data.dt[, x2.sq := x2^2 ]
ggplot()+
  geom_point(aes(
    x1.sq, x2.sq),
    data=data.dt)
```

In our simulation, we assume that the output score `f` is a linear
function of `x1.sq`, and ignores `x2.sq`. The plot below visualizes
the output scores using the point fill aesthetic.

```{r}
data.dt[, f := x1.sq]
true.decision.boundary <- 0.2
ggplot()+
  theme_bw()+
  scale_fill_gradient2(midpoint=true.decision.boundary)+
  geom_point(aes(
    x1.sq, x2.sq, fill=f),
    shape=21,
    color="grey",
    data=data.dt)
```

In particular, we assume that the label `y` is negative (-1) if
`x1.sq + noise < threshold`, and positive (1) otherwise. The plot below
visualizes the scores and labels, as a function of the input feature
`x1`. It also shows the true score function in black. Of course, we
would not be able to make this visualization with real data (only the
labels are known in real data, not the scores).

```{r}
data.dt[, f.noise := f+rnorm(N, 0, 0.2)]
data.dt[, y.num := ifelse(f.noise<true.decision.boundary, -1, 1)]
data.dt[, y := factor(y.num)]
table(data.dt$y)
scores <- data.table(x1=seq(-1, 1, l=101))
scores[, x1.sq := x1^2 ]
scores[, f := x1.sq ]
x1.boundaries <- data.table(boundary=c(1, -1)*sqrt(true.decision.boundary))
ggplot()+
  scale_y_continuous(breaks=seq(0, 1, by=0.2))+
  geom_vline(aes(xintercept=boundary), color="grey50", data=x1.boundaries)+
  geom_line(aes(x1, f), data=scores)+
  geom_point(aes(x1, f.noise, color=y),
             shape=21,
             fill=NA,
             data=data.dt)
```

The plot below shows the scores and labels, as a function of the
squared feature `x1.sq`. It is clear that the score function that we
want to learn is linear in `x1.sq`.

```{r}
x1sq.boundary <- data.table(boundary=true.decision.boundary)
ggplot()+
  scale_y_continuous(breaks=seq(0, 1, by=0.2))+
  scale_x_continuous(breaks=seq(0, 1, by=0.2))+
  geom_vline(aes(
    xintercept=boundary),
    color="grey50",
    data=x1sq.boundary)+
  geom_line(aes(x1.sq, f), data=scores)+
  geom_point(aes(
    x1.sq, f.noise, color=y),
    shape=21,
    fill=NA,
    data=data.dt)
```

Next, we visualize the labels in the two-dimensional squared feature
space. It is clear that the decision boundary is linear in this space.

```{r}
ggplot()+
  scale_y_continuous(breaks=seq(0, 1, by=0.2))+
  scale_x_continuous(breaks=seq(0, 1, by=0.2))+
  geom_vline(aes(
    xintercept=boundary),
    color="grey50",
    data=x1sq.boundary)+
  geom_point(aes(
    x1.sq, x2.sq, color=y),
    shape=21,
    fill=NA,
    data=data.dt)
```

The plot below shows the input feature space (`x1` and `x2`). It is
clear that the decision boundary is non-linear in `x1`.

```{r}
ggplot()+
  scale_y_continuous(breaks=seq(-1, 1, by=0.2))+
  scale_x_continuous(breaks=seq(-1, 1, by=0.2))+
  geom_vline(aes(
    xintercept=boundary),
    color="grey50",
    data=x1.boundaries)+
  geom_point(aes(
    x1, x2, color=y),
    shape=21,
    fill=NA,
    data=data.dt)
```

The animint below uses `clickSelects` to show which points in the
input and squared space correspond. We just need to create an
`data.i` variable that has a unique ID for each data point.

```{r, Ch12-viz-simple}
data.dt[, data.i := 1:.N]
YVAR <- function(dt, y.var){
  dt$y.var <- factor(y.var, c("x2", "x2.sq", "f"))
  dt
}
animint(
  input=ggplot()+
    ggtitle("input feature space")+
    theme_bw()+
    theme(panel.margin=grid::unit(0, "lines"))+
    facet_grid(y.var ~ ., scales="free")+
    scale_x_continuous(breaks=seq(-1, 1, by=0.2))+
    ylab("")+
    guides(color="none")+
    geom_vline(aes(
      xintercept=boundary),
      color="grey50",
      data=x1.boundaries)+
    geom_point(aes(
      x1, x2, color=y),
      clickSelects="data.i",
      size=4,
      alpha=0.7,
      data=YVAR(data.dt, "x2"))+
    geom_line(aes(
      x1, f),
      data=YVAR(scores, "f"))+
    geom_point(aes(
      x1, f.noise, color=y),
      clickSelects="data.i",
      size=4,
      alpha=0.7,
      data=YVAR(data.dt, "f")),
  square=ggplot()+
    ggtitle("squared feature space")+
    theme_bw()+
    theme(panel.margin=grid::unit(0, "lines"))+
    facet_grid(y.var ~ ., scales="free")+
    ylab("")+
    scale_x_continuous(breaks=seq(0, 1, by=0.2))+
    geom_vline(aes(
      xintercept=boundary),
      color="grey50",
      data=x1sq.boundary)+
    geom_point(aes(
      x1.sq, x2.sq, color=y),
      clickSelects="data.i",
      size=4,
      alpha=0.7,
      data=YVAR(data.dt, "x2.sq"))+
    geom_line(aes(
      x1.sq, f),
      data=YVAR(scores, "f"))+
    geom_point(aes(
      x1.sq, f.noise, color=y),
      clickSelects="data.i",
      size=4,
      alpha=0.7,
      data=YVAR(data.dt, "f")))
```

Note how we used two multi-panel plots with the
[addColumn then facet](Ch17-appendix.html#addColumn-then-facet) idiom,
rather than creating four separate plots. This emphasizes the fact
that some plots/facets have a common `x1` or `x1.sq` axis. Note that
we also hid the color legend in the first plot, since it is sufficient
to just have one color legend.

### Linear SVM {#linear-svm}

```{r}
train.i <- 1:N
data.dt[, set := "test"]
data.dt <- data.dt[train.i, set := "train"]
table(data.dt$set)
train.dt <- data.dt[set=="train",]
ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(set ~ .)+
  geom_vline(aes(
    xintercept=boundary),
    color="grey50",
    data=x1.boundaries)+
  scale_y_continuous(breaks=seq(-1, 1, by=0.2))+
  scale_x_continuous(breaks=seq(-1, 1, by=0.2))+
  geom_point(aes(
    x1, x2, color=y),
    data=data.dt)
```

We begin by fitting a linear SVM to the train data in the squared
feature space, and visualizing the true labels `y` along with the
predicted labels `pred.y`.

```{r}
library(kernlab)
squared.mat <- train.dt[, cbind(x1.sq, x2.sq)]
y.vec <- train.dt$y
fit <- ksvm(squared.mat, y.vec, kernel="vanilladot")
train.dt$pred.y <- predict(fit)
ggplot()+
  geom_point(aes(
    x1.sq, x2.sq, color=pred.y, fill=y),
    shape=21,
    size=4,
    stroke=2,
    data=train.dt)
```

It is clear from the plot above that there are several mis-classified
train data points. In the plot below we visualize the decision
boundary and margin.

```{r}
predF <- function(fit, X){
  fit.sc <- scaling(fit)$x.scale
  if(is.null(fit.sc)){
    fit.sc <- list(
      "scaled:center"=c(0,0),
      "scaled:scale"=c(1,1))
  }
  mu <- fit.sc[["scaled:center"]]
  sigma <- fit.sc[["scaled:scale"]]
  X.sc <- scale(X, mu, sigma)
  kernelMult(
    kernelf(fit),
    X.sc,
    xmatrix(fit)[[1]],
    coef(fit)[[1]])-b(fit)
}
xsq.vec <- seq(0, 1, l=41)
grid.sq.dt <- data.table(expand.grid(
  x1.sq=xsq.vec,
  x2.sq=xsq.vec))
grid.sq.dt[, pred.f := predF(fit, cbind(x1.sq, x2.sq))]
train.dt[, train.error := ifelse(y==pred.y, "correct", "error")]
ggplot()+
  theme_bw()+
  scale_color_manual(values=c(error="black", correct=NA))+
  geom_point(aes(
    x1.sq, x2.sq, fill=y, color=train.error),
    shape=21,
    stroke=1,
    size=4,
    data=train.dt)+
  geom_vline(aes(
    xintercept=boundary), color="grey50",
    data=x1sq.boundary)+
  geom_contour(aes(
    x1.sq, x2.sq, z=pred.f),
    breaks=0,
    color="black",
    data=grid.sq.dt)+
  geom_contour(aes(
    x1.sq, x2.sq, z=pred.f),
    breaks=c(-1, 1),
    color="black",
    linetype="dashed",
    data=grid.sq.dt)
```

The plot above shows the true decision boundary using a grey
`vline`. It also uses `geom_contour` to display the decision boundary
(solid black line, predicted score 0) and the margin (dashed black
line, predicted score -1 and 1). Since the decision boundary and
margin are linear in this space, we can also use `geom_abline` to
display them. To do that we need to do some math, and work out the
equations for the slope and intercepts of those lines (as a function
of the learned bias `b(fit)` and `weight.vec`, as well as the scale
parameters `mu` and `sigma`).

```{r}
## The equation of the margin lines is x2 = m2 + s2/w2[c+b+w1*m1/s1]
## -s2*w1/(w2*s1)*x1 for c=1 and -1. x is input feature, m is mean, s
## is scale, w is learned weight.
fit.sc <- scaling(fit)$x.scale
if(is.null(fit.sc)){
  fit.sc <- list(
    "scaled:center"=c(0,0),
    "scaled:scale"=c(1,1))
}
mu <- fit.sc[["scaled:center"]]
sigma <- fit.sc[["scaled:scale"]]
weight.vec <- colSums(xmatrix(fit)[[1]]*coef(fit)[[1]])
predF.linear <- function(fit, X){
  X.sc <- scale(X, mu, sigma)
  X.sc %*% weight.vec - b(fit)
}
abline.dt <- data.table(
  y=factor(c(-1,0,1)),
  boundary=c("margin", "decision", "margin"),
  intercept=mu[2]+sigma[2]/weight.vec[2]*(
    c(-1, 0, 1)+b(fit)+weight.vec[1]*mu[1]/sigma[1]),
  slope=-weight.vec[1]*sigma[2]/(weight.vec[2]*sigma[1]))
ggplot()+
  theme_bw()+
  scale_linetype_manual(values=c(margin="dashed", decision="solid"))+
  geom_abline(aes(
    slope=slope, intercept=intercept, linetype=boundary),
    color="green",
    size=1,
    data=abline.dt)+
  geom_point(aes(
    x1.sq, x2.sq, color=y),
    shape=21,
    fill=NA,
    size=4,
    data=train.dt)+
  geom_contour(aes(
    x1.sq, x2.sq, z=pred.f),
    breaks=0,
    color="black",
    data=grid.sq.dt)+
  geom_contour(aes(
    x1.sq, x2.sq, z=pred.f),
    breaks=c(-1, 1),
    color="black",
    linetype="dashed",
    data=grid.sq.dt)
```

The plot above confirms that our computation of the slope and
intercepts (green lines) agrees with the contours (black lines). In
the plot below, we show the learned `alpha` coefficients, and add a
`geom_segment` to visualize the slack.

```{r}
train.dt[, alpha := 0]
train.row.vec <- as.integer(rownames(xmatrix(fit)[[1]]))
train.dt[train.row.vec, alpha := kernlab::alpha(fit)[[1]] ]
train.dt[, status := ifelse(
  alpha==0, "alpha=0",
  ifelse(alpha==1, "alpha=C", "0<alpha<C"))]
slack.slope <- weight.vec[2]*sigma[1]/(weight.vec[1]*sigma[2])
slack.dt <- train.dt[alpha==1,]
slack.join <- abline.dt[slack.dt, on=list(y)]
slack.join[, x1.sq.margin := (
  x2.sq-slack.slope*x1.sq-intercept)/(slope-slack.slope)]
slack.join[, x2.sq.margin := slope*x1.sq.margin + intercept]
sv.colors <- c(
  "alpha=0"="white",
  "0<alpha<C"="black",
  "alpha=C"="grey")
ggplot()+
  theme_bw()+
  scale_linetype_manual(values=c(margin="dashed", decision="solid"))+
  geom_vline(aes(
    xintercept=boundary), color="violet",
    data=x1sq.boundary)+
  geom_abline(aes(
    slope=slope, intercept=intercept, linetype=boundary),
    size=1,
    data=abline.dt)+
  geom_segment(aes(
    x1.sq, x2.sq,
    xend=x1.sq.margin, yend=x2.sq.margin),
    color="grey",
    data=slack.join)+
  scale_fill_manual(values=sv.colors, breaks=names(sv.colors))+
  geom_point(aes(
    x1.sq, x2.sq, color=y, fill=status),
    shape=21,
    size=4,
    data=train.dt)
```

The plot above shows the slack in grey segments, and the decision and
margin lines in black. The Bayes decision boundary is shown in the
background as a vertical violet line. The support vectors are the
points with non-zero alpha coefficients. Black filled support vectors
are on the margin, and grey support vectors are on the wrong side of
the margin (and have non-zero slack). The plot below shows the model
that was learned in the original feature space,

```{r}
n.grid <- 41
x.vec <- seq(-1, 1, l=n.grid)
grid.dt <- data.table(expand.grid(
  x1=x.vec,
  x2=x.vec))
getBoundaryDF <- function(score.vec, level.vec=c(-1, 0, 1)){
  stopifnot(length(score.vec) == n.grid * n.grid)
  several.paths <- contourLines(
    x.vec, x.vec,
    matrix(score.vec, n.grid, n.grid),
    levels=level.vec)
  contour.list <- list()
  for(path.i in seq_along(several.paths)){
    contour.list[[path.i]] <- with(several.paths[[path.i]], data.table(
      path.i,
      level.num=as.numeric(level),
      level.fac=factor(level, level.vec),
      boundary=ifelse(level==0, "decision", "margin"),
      x1=x, x2=y))
  }
  do.call(rbind, contour.list)
}
grid.dt[, pred.f := predF(fit, cbind(x1^2, x2^2))]
boundaries <- grid.dt[, getBoundaryDF(pred.f)]
ggplot()+
  scale_linetype_manual(values=c(margin="dashed", decision="solid"))+
  geom_vline(aes(
    xintercept=boundary),
    color="violet",
    data=x1.boundaries)+
  geom_path(aes(
    x1, x2, group=path.i, linetype=boundary),
    size=1,
    data=boundaries)+
  scale_fill_manual(values=sv.colors, breaks=names(sv.colors))+
  scale_size_manual(values=c(correct=2, error=4))+
  geom_point(aes(
    x1, x2, color=y,
    size=train.error,
    fill=status),
    shape=21,
    data=train.dt)
```

The goal below will be to make an animint that shows
how the decision boundary, margin, and slack change as a function of
the cost parameter.

```{r Ch12-viz-linear}
modelInfo.list <- list()
predictions.list <- list()
slackSegs.list <- list()
modelLines.list <- list()
inputBoundaries.list <- list()
setErrors.list <- list()
cost.by <- 0.2
for(cost.param in 10^seq(-1, 1, by=cost.by)){
  fit <- ksvm(
    squared.mat, y.vec, kernel="vanilladot", scaled=FALSE, C=cost.param)
  fit.sc <- scaling(fit)$x.scale
  if(is.null(fit.sc)){
    fit.sc <- list(
      "scaled:center"=c(0,0),
      "scaled:scale"=c(1,1))
  }
  mu <- fit.sc[["scaled:center"]]
  sigma <- fit.sc[["scaled:scale"]]
  weight.vec <- colSums(xmatrix(fit)[[1]]*coef(fit)[[1]])
  grid.sq.dt[, pred.f := predF(fit, cbind(x1.sq, x2.sq))]
  data.dt[, pred.y := predict(fit, cbind(x1.sq, x2.sq))]
  one.error <- data.dt[, list(errors=sum(y!=pred.y)), by=set]
  setErrors.list[[paste(cost.param)]] <- data.table(
    cost.param, one.error)
  train.dt[, pred.f := predF(fit, cbind(x1^2, x2^2))]
  grid.dt[, pred.f := predF(fit, cbind(x1^2, x2^2))]
  boundaries <- getBoundaryDF(grid.dt$pred.f)
  inputBoundaries.list[[paste(cost.param)]] <- data.table(
    cost.param, boundaries)
  train.dt$alpha <- 0
  train.row.vec <- as.integer(rownames(xmatrix(fit)[[1]]))
  train.dt[train.row.vec, alpha := kernlab::alpha(fit)[[1]] ]
  train.dt[, status := ifelse(
    alpha==0, "alpha=0",
    ifelse(alpha==cost.param, "alpha=C", "0<alpha<C"))]
  ## The equation of the margin lines is x2 = m2 + s2/w2[c+b+w1*m1/s1]
  ## -s2*w1/(w2*s1)*x1 for c=1 and -1. x is input feature, m is mean, s
  ## is scale, w is learned weight.
  slack.slope <- weight.vec[2]*sigma[1]/(weight.vec[1]*sigma[2])
  abline.dt <- data.table(
    y=factor(c(-1,0,1)),
    boundary=c("margin", "decision", "margin"),
    intercept=mu[2]+sigma[2]/weight.vec[2]*(
      c(-1, 0, 1)+b(fit)+weight.vec[1]*mu[1]/sigma[1]),
    slope=-weight.vec[1]*sigma[2]/(weight.vec[2]*sigma[1]))
  slack.dt <- train.dt[alpha==cost.param]
  slack.join <- abline.dt[slack.dt, on=list(y)]
  slack.join[, x1.sq.margin := (
    x2.sq-slack.slope*x1.sq-intercept)/(slope-slack.slope)]
  slack.join[, x2.sq.margin := slope*x1.sq.margin + intercept]
  norm.weights <- as.numeric(weight.vec %*% weight.vec)
  modelInfo.list[[paste(cost.param)]] <- data.table(
    cost.param,
    total.slack=slack.join[, sum(1-pred.f*y.num)],
    norm.weights,
    margin.size=2/sqrt(norm.weights))
  predictions.list[[paste(cost.param)]] <- data.table(
    cost.param, train.dt)
  slackSegs.list[[paste(cost.param)]] <- data.table(
    cost.param, slack.join)
  modelLines.list[[paste(cost.param)]] <- data.table(
    cost.param, abline.dt)
}
inputBoundaries <- do.call(rbind, inputBoundaries.list)
predictions <- do.call(rbind, predictions.list)
slackSegs <- do.call(rbind, slackSegs.list)
modelLines <- do.call(rbind, modelLines.list)
modelInfo <- do.call(rbind, modelInfo.list)
setErrors <- do.call(rbind, setErrors.list)
modelInfo.tall <- melt(modelInfo, id.vars="cost.param")
grid.sq.dt$boundary <- "true"
setErrors$variable <- "errors"
inputBoundaries[, boundary := ifelse(level.num==0, "decision", "margin")]
slackSegs$boundary <- "margin"
set.label.select <- data.table(
  cost.param=range(setErrors$cost.param),
  set=c("test", "train"),
  hjust=c(1, 0))
set.labels <- setErrors[set.label.select, on=list(cost.param, set)]
animint(
  selectModel=ggplot()+
    ggtitle("select regularization parameter")+
    geom_tallrect(aes(
      xmin=log10(cost.param)-cost.by/2,
      xmax=log10(cost.param)+cost.by/2),
      clickSelects="cost.param",
      alpha=0.5,
      data=modelInfo)+
    theme_bw()+
    theme(panel.margin=grid::unit(0, "lines"))+
    facet_grid(variable ~ ., scales="free")+
    geom_line(aes(
      log10(cost.param), errors,
      group=set, color=set),
      data=setErrors)+
    geom_text(aes(
      log10(cost.param), errors-1, label=set,
      hjust=hjust,
      color=set),
      data=set.labels)+
    guides(color="none")+
    geom_line(aes(
      log10(cost.param), log10(value)),
      data=modelInfo.tall),
  inputSpace=ggplot()+
    ggtitle("Input space features")+
    scale_fill_manual(values=sv.colors, breaks=names(sv.colors))+
    geom_vline(aes(
      xintercept=boundary),
      color="violet",
      data=x1.boundaries)+
    guides(color="none", fill="none", linetype="none")+
    scale_linetype_manual(values=c(
      "-1"="dashed",
      "0"="solid",
      "1"="dashed"))+
    geom_path(aes(
      x1, x2,
      group=path.i,
      linetype=level.fac),
      showSelected=c("boundary", "cost.param"),
      color="black",
      data=inputBoundaries)+
    geom_point(aes(
      x1, x2, color=y, fill=status),
      showSelected=c("status", "y", "data.i", "cost.param"),
      size=5,
      color="grey",
      data=predictions)+
    geom_point(aes(
      x1, x2, color=y, fill=status),
      showSelected=c("cost.param", "status", "y"),
      clickSelects="data.i",
      size=3,
      data=predictions),
  kernelSpace=ggplot()+
    ggtitle("Kernel space features")+
    geom_vline(aes(
      xintercept=boundary), color="violet",
      data=x1sq.boundary)+
    coord_cartesian(xlim=c(0, 1), ylim=c(0, 1))+
    geom_abline(aes(
      slope=slope, intercept=intercept, linetype=boundary),
      showSelected="cost.param",
      color="black",
      data=modelLines)+
    scale_linetype_manual(values=c(
      decision="solid",
      margin="dashed",
      true="solid"))+
    geom_point(aes(
      x1.sq, x2.sq, color=y, fill=status),
      showSelected=c("data.i", "cost.param"),
      size=5,
      color="grey",
      data=predictions)+
    geom_point(aes(
      x1.sq, x2.sq, color=y, fill=status),
      clickSelects="data.i",
      showSelected="cost.param",
      size=3,
      data=predictions)+
    scale_fill_manual(values=sv.colors, breaks=names(sv.colors))+
    geom_segment(aes(
      x1.sq, x2.sq, 
      xend=x1.sq.margin, yend=x2.sq.margin),
      showSelected=c("cost.param", "boundary"),
      color="grey",
      data=slackSegs))
```

### Non-linear polynomial kernel SVM {#nonlinear-polynomial-kernel-svm}

In the previous section we fit a linear kernel in the squared feature
space, which resulted in learning a function which is non-linear in
terms of the original feature space. In this section we directly fit a
non-linear polynomial kernel in the original space.

```{r Ch12-viz-poly}
predictions.list <- list()
inputBoundaries.list <- list()
setErrors.list <- list()
cost.by <- 0.2
orig.mat <- train.dt[, cbind(x1, x2)]
for(cost.param in 10^seq(-1, 3, by=cost.by)){
  for(degree.num in seq(1, 6, by=1)){
    k <- polydot(degree.num, offset=0)
    fit <- ksvm(
      orig.mat, y.vec, kernel=k, scaled=FALSE, C=cost.param)
    grid.dt[, pred.f := predF(fit, cbind(x1, x2))]
    grid.dt[, pred.y := predict(fit, cbind(x1, x2))]
    grid.dt[, stopifnot(sign(pred.f) == pred.y)]
    data.dt[, pred.y := predict(fit, cbind(x1, x2))]
    one.error <- data.dt[, list(errors=sum(y != pred.y)), by=set]
    setErrors.list[[paste(cost.param, degree.num)]] <- data.table(
      cost.param, degree.num, one.error)
    boundaries <- getBoundaryDF(grid.dt$pred.f)
    if(is.data.frame(boundaries) && nrow(boundaries)){
      cost.deg <- paste(cost.param, degree.num)
      inputBoundaries.list[[cost.deg]] <- data.table(
        cost.param, degree.num, boundaries)
    }
    train.dt[, alpha := 0]
    train.row.vec <- as.integer(rownames(xmatrix(fit)[[1]]))
    train.dt[train.row.vec, alpha := kernlab::alpha(fit)[[1]] ]
    train.dt[, status := ifelse(
      alpha==0, "alpha=0",
      ifelse(alpha==cost.param, "alpha=C", "0<alpha<C"))]
    predictions.list[[paste(cost.param, degree.num)]] <- data.table(
      cost.param, degree.num, train.dt)
  }
}
inputBoundaries <- do.call(rbind, inputBoundaries.list)
predictions <- do.call(rbind, predictions.list)
setErrors <- do.call(rbind, setErrors.list)

testErrors <- setErrors[set=="test"]
testErrors$select <- "degree"
setErrors$select <- "cost"

animint(
  selectModel=ggplot()+
    ggtitle("select hyper parameters")+
    geom_tallrect(aes(
      xmin=log10(cost.param)-cost.by/2,
      xmax=log10(cost.param)+cost.by/2),
      clickSelects="cost.param",
      alpha=0.5,
      data=setErrors[degree.num==1 & set=="train",])+
    theme_bw()+
    theme(panel.margin=grid::unit(0, "lines"))+
    facet_grid(select ~ ., scales="free")+
    ylab("")+
    geom_line(aes(
      log10(cost.param), errors,
      key=set,
      group=set,
      color=set),
      showSelected="degree.num",
      data=setErrors)+
    scale_fill_gradient("TestErrors", low="white", high="red")+
    geom_tile(aes(
      log10(cost.param), degree.num, fill=errors),
      clickSelects="degree.num",
      data=testErrors),
  inputSpace=ggplot()+
    theme_bw()+
    ggtitle("Input space features")+
    scale_fill_manual(values=sv.colors, breaks=names(sv.colors))+
    geom_vline(aes(
      xintercept=boundary),
      color="violet",
      data=x1.boundaries)+
    scale_linetype_manual(values=c(
      margin="dashed",
      decision="solid"))+
    geom_path(aes(
      x1, x2,
      group=path.i,
      linetype=boundary),
      showSelected=c("degree.num", "cost.param"),
      color="black",
      data=inputBoundaries)+
    geom_point(aes(
      x1, x2, color=y, fill=status),
      showSelected=c("cost.param", "degree.num"),
      size=3,
      data=predictions))
```

## Chapter summary and exercises {#exercises}

We used ggplots to visualize the Support Vector Machine model for
binary classification. We used animint and interactivity to show how
the SVM decision boundary changes as a function of the model
hyper-parameters.

Exercises:

* Use `rbfdot` as the kernel function. Compute train and test error,
  then add a new panel to the "select hyper parameters" plot.
* By default ggplot2 uses the same two colors for the **y** and
  **set** legends, which could be confusing. Change the colors in one
  of the two legends so that they are different.

Next, [Chapter 13](Ch13-poisson-regression.html) explains how to
visualize the Poisson regression model.
