---
title: K-Nearest-Neighbors example
layout: default
output: bookdown::html_chapter
---

# Chapter 13, Poisson regression

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch13-figures/")
```

This goal of this chapter is to create an interactive data
visualization that explains
[Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression),
a machine learning model for predicting an integer-valued output from
inputs that are real-valued vectors. This is a "linear regression"
model since it learns a linear function from the inputs to the
output. Like least squares regression, Poisson regression can be
formulated as a maximum likelihood problem. However, it differs from
least squares linear regression since it uses a Poisson distribution
to model the output labels, instead of a Gaussian distribution. This
modeling choice is appropriate when output labels are non-negative
integers. 

## Plot the probability mass function and select the Poisson mean parameter

The goal of this section is to create a data visualization that shows
the probability mass function for a selected Poisson mean parameter.

TODO: sketch point(prob versus count), tallrect mean.

```{r}
library(data.table)
poisson.mean.diff <- 0.5
poisson.mean.vec <- seq(0, 10, by=poisson.mean.diff)
quantile.max <- 0.99
poisson.prob.list <- list()
for(poisson.mean in poisson.mean.vec){
  count.max <- qpois(quantile.max, poisson.mean)
  count <- 0:count.max
  probability <- dpois(count, poisson.mean)
  cum.prob <- cumsum(probability)
  poisson.prob.list[[paste(poisson.mean)]] <- data.table(
    poisson.mean,
    count,
    probability,
    cum.prob)
}
poisson.prob <- do.call(rbind, poisson.prob.list)
poisson.prob
```

The static data viz below shows one facet for each Poisson
distribution.

```{r}
mean.tallrects <- data.table(
  poisson.mean=poisson.mean.vec,
  min=poisson.mean.vec - poisson.mean.diff/2,
  max=poisson.mean.vec + poisson.mean.diff/2)
library(animint)
prob.mass <- ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "cm"))+
  geom_tallrect(aes(xmin=min, xmax=max, clickSelects=poisson.mean),
                alpha=0.6,
                data=mean.tallrects)+
  geom_point(aes(count, probability, showSelected=poisson.mean,
                 tooltip=sprintf("prob(count = %d) = %f", count, probability)),
             color="red",
             size=5,
             data=poisson.prob)
prob.mass+
  facet_wrap("poisson.mean")
```

Note that we used `alpha=0.6` with `geom_tallrect`, which means that
the tallrect for the selected mean has 0.6 opacity, and the other
tallrects have 0.1 opacity. Note also that we use `color="red"` and
`size=5` with `geom_point` so that it is easier to see the points on a
grey background, and to hover the cursor over the points to see the
tooltip. We next create an interactive version with animint.

```{r Ch13-viz-prob}
viz.prob <- list(prob=prob.mass)
structure(viz.prob, class="animint")
```

You can click the viz above to change the mean of the Poisson
distribution. You can also hover the cursor over a data point to see
its probability. Note that for integer values of the Poisson mean,
there are two counts that are the most probable (the mode of the
Poisson distribution). For example the Poisson distribution with a
mean of 7 attains its maximum probability of about 0.149 at count
values of 6 and 7.

## Add a panel for the cumulative distribution function

To add a panel for the cumulative distribution function, we will
re-make the ggplot below. When we specify the data sets, we will use
the [addColumn then facet](ChXX-appendix.html#addColumn-then-facet)
idiom to add a `panel` variable.

```{r Ch13-viz-cum-prob}
addPanel <- function(dt, panel){
  data.table(dt, panel=factor(panel, c("probability", "cum prob")))
}
quantile.max.dt <- data.table(quantile.max)
viz.cum.prob <- list(
  prob=ggplot()+
    theme_bw()+
    theme(panel.margin=grid::unit(0, "cm"))+
    facet_grid(panel ~ ., scales="free")+
    geom_hline(aes(yintercept=quantile.max),
               color="grey",
               data=addPanel(quantile.max.dt, "cum prob"))+
    geom_tallrect(aes(xmin=min, xmax=max, clickSelects=poisson.mean),
                  alpha=0.6,
                  data=mean.tallrects)+
    geom_point(aes(count, probability, showSelected=poisson.mean,
                   tooltip=sprintf(
                     "prob(count = %d) = %f", count, probability)),
               color="red",
               size=5,
               data=addPanel(poisson.prob, "probability"))+
    geom_point(aes(count, cum.prob, showSelected=poisson.mean,
                   tooltip=sprintf(
                     "prob(count <= %d) = %f", count, cum.prob)),
               color="red",
               size=5,
               data=addPanel(poisson.prob, "cum prob"))
)
structure(viz.cum.prob, class="animint")
```

Note how we used `addPanel` to add a `panel` variable to all the data
sets for each geom except `geom_tallrect`. Using `panel` as a facet
variable has the effect of drawing each geom in only one panel, except
the `geom_tallrect` which is drawn in each panel.

Note that we also used a `geom_hline` to show 0.99, the cumulative
distribution function threshold that was used to determine the set of
points to plot for each Poisson distribution. This is an example of
"show your arbitrary choices," one of the general principles of
designing good interactive data visualizations.

## Add a plot of the Poisson loss and a selector for label value

Next we will compute the Poisson loss for several values of the output
label.

```{r}
PoissonLoss <- function(count, seg.mean){
  stopifnot(is.numeric(count))
  stopifnot(is.numeric(seg.mean))
  if(any(seg.mean < 0)){
    stop("PoissonLoss undefined for negative segment mean")
  }
  if(length(seg.mean)==1)seg.mean <- rep(seg.mean, length(count))
  if(length(count)==1)count <- rep(count, length(seg.mean))
  stopifnot(length(seg.mean) == length(count))
  not.integer <- round(count) != count
  is.negative <- count < 0
  loss <- ifelse(
    not.integer | is.negative, Inf,
    ifelse(seg.mean == 0, ifelse(count == 0, 0, Inf),
           seg.mean - count * log(seg.mean)
           ## This term makes all the minima zero.
           -ifelse(count == 0, 0, count - count*log(count))))
  loss
}
```

Below we compute the loss for several values of count, using the
[list of data tables idiom](ChXX-appendix.html#list-of-data-tables).

```{r}
count.vec <- unique(poisson.prob$count)
count.range <- range(count.vec)
mean.vec <- seq(count.range[1], count.range[2], l=100)
loss.min.list <- list()
loss.fun.list <- list()
for(count in count.vec){
  loss <- PoissonLoss(count, mean.vec)
  loss.fun.list[[paste(count)]] <- data.table(
    count, poisson.mean=mean.vec, loss)
  loss.min.list[[paste(count)]] <- data.table(count, loss=0)
}
loss.fun <- do.call(rbind, loss.fun.list)
loss.min <- do.call(rbind, loss.min.list)
```

Next we make a data viz with an additional panel.

```{r Ch13-viz-loss}
viz.loss <- list(
  prob=ggplot()+
    theme_bw()+
    theme(panel.margin=grid::unit(0, "cm"))+
    facet_grid(panel ~ ., scales="free")+
    geom_hline(aes(yintercept=quantile.max),
               color="grey",
               data=addPanel(quantile.max.dt, "cum prob"))+
    geom_point(aes(count, probability, showSelected=poisson.mean,
                   clickSelects=count,
                   tooltip=sprintf(
                     "prob(count = %d) = %f", count, probability)),
               color="red",
               size=5,
               alpha=0.7,
               data=addPanel(poisson.prob, "probability"))+
    geom_point(aes(count, cum.prob, showSelected=poisson.mean,
                   clickSelects=count,
                   tooltip=sprintf(
                     "prob(count <= %d) = %f", count, cum.prob)),
               color="red",
               size=5,
               alpha=0.7,
               data=addPanel(poisson.prob, "cum prob")),
  loss=ggplot()+
    geom_line(aes(poisson.mean, loss, showSelected=count),
              data=loss.fun)+
    geom_point(aes(count, loss, showSelected=count),
               data=loss.min)+
    geom_tallrect(aes(xmin=min, xmax=max, clickSelects=poisson.mean),
                  alpha=0.6,
                  data=mean.tallrects))
structure(viz.loss, class="animint")
```

The data viz above shows the probability on the left and the Poisson
loss on the right.

## Chapter summary and exercises

Next, [Chapter 14](Ch14-PeakSegJoint.html) explains how to use the
`.variable` and `.value` aesthetics to visualize the PeakSegJoint
machine learning model.
