---
title: PeakSegJoint
layout: default
output: bookdown::html_chapter
---

# Chapter 14, PeakSegJoint

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch14-figures/")
```

This chapter explains how to use
[.variable and .value](Ch06-other.html#variable-value) aesthetics for
creating data-driven selector names. The example data come from the
[PeakSegJoint package](https://github.com/tdhock/PeakSegJoint).

```{r download-PSJ}
library(animint)
if(!file.exists("PSJ.RData")){
  u <- paste0(
    "http://github.com/tdhock/animint-examples",
    "/blob/master/data/PSJ.RData?raw=true"
    )
  require(httr)
  request <- GET(u)
  stop_for_status(request)
  writeBin(content(request), "PSJ.RData")
}
load("PSJ.RData")
```

## Compute data to plot

The section below computes some common data that we will use in two
data visualizations below.

```{r Ch14-setup-data}
res.error <- PSJ$error.total.chunk
ann.colors <-
  c(noPeaks="#f6f4bf",
    peakStart="#ffafaf",
    peakEnd="#ff4c4c",
    peaks="#a445ee")
## prob.regions are the black segments that show which regions are
## mapped to which segmentation problems.
all.regions <- do.call(rbind, PSJ$regions.by.problem)
prob.regions.names <-
  c("bases.per.problem", "problem.i", "problem.name",
    "chromStart", "chromEnd")
prob.regions <- unique(data.frame(all.regions)[, prob.regions.names])
prob.regions$sample.id <- "problems"

all.modelSelection <- do.call(rbind, PSJ$modelSelection.by.problem)
modelSelection.errors <-
  all.modelSelection[!is.na(all.modelSelection$errors), ]
penalty.range <-
  with(all.modelSelection, c(min(max.log.lambda), max(min.log.lambda)))
penalty.mid <- mean(penalty.range)

coverage.counts <- table(PSJ$coverage$sample.id)
facet.rows <- length(coverage.counts)+1
dvec <- diff(log(res.error$bases.per.problem))
dval <- exp(mean(dvec))
dval2 <- (dval-1)/2 + 1
res.error$min.bases.per.problem <- res.error$bases.per.problem/dval2
res.error$max.bases.per.problem <- res.error$bases.per.problem*dval2

modelSelection.labels <- unique(with(all.modelSelection, {
  data.frame(problem.name=problem.name,
             bases.per.problem=bases.per.problem,
             problemStart=problemStart,
             problemEnd=problemEnd,
             min.log.lambda=penalty.mid,
             peaks=max(peaks)+0.5)
}))
```

## Define data viz using for loops

The R code below constructs a data viz using for loops. TODO break up
into pieces and comment each piece to explain what we are doing?

```{r Ch14-for-construct}
print(timing.for.construct <- system.time({
  viz.for <- list(
    coverage=ggplot()+
      geom_segment(aes(chromStart/1e3, problem.i,
                       xend=chromEnd/1e3, yend=problem.i,
                       showSelected=bases.per.problem,
                       clickSelects=problem.name),
                   data=prob.regions)+
      ggtitle("select problem")+
      geom_text(aes(chromStart/1e3, problem.i,
                    showSelected=bases.per.problem,
                    label=sprintf("%d problems mean size %.1f kb",
                                  problems, mean.bases/1e3)),
                data=PSJ$problem.labels,
                hjust=0)+
      geom_segment(aes(problemStart/1e3, problem.i,
                       showSelected=bases.per.problem,
                       clickSelects=problem.name,
                       xend=problemEnd/1e3, yend=problem.i),
                   size=5,
                   data=PSJ$problems)+
      scale_y_continuous("aligned read coverage",
                         breaks=function(limits){
                           floor(limits[2])
                         })+
      scale_linetype_manual("error type",
                            limits=c("correct", 
                                     "false negative",
                                     "false positive"
                                     ),
                            values=c(correct=0,
                                     "false negative"=3,
                                     "false positive"=1))+
      scale_x_continuous(paste("position on chr11",
                               "(kilo bases = kb)"))+
      coord_cartesian(xlim=c(118167.406, 118238.833))+
      geom_tallrect(aes(xmin=chromStart/1e3, xmax=chromEnd/1e3,
                        fill=annotation),
                    alpha=0.5,
                    color="grey",
                    data=PSJ$filled.regions)+
      scale_fill_manual(values=ann.colors)+
      theme_bw()+
      theme_animint(width=1500, height=facet.rows*100)+
      theme(panel.margin=grid::unit(0, "cm"))+
      facet_grid(sample.id ~ ., labeller=function(var, val){
        sub("McGill0", "", sub(" ", "\n", val))
      }, scales="free")+
      geom_line(aes(base/1e3, count),
                data=PSJ$coverage,
                color="grey50"),

    resError=ggplot()+
      ggtitle("select problem size")+
      ylab("minimum percent incorrect regions")+
      geom_tallrect(aes(xmin=min.bases.per.problem,
                        xmax=max.bases.per.problem,
                        clickSelects=bases.per.problem),
                    alpha=0.5,
                    data=res.error)+
      scale_x_log10()+
      geom_line(aes(bases.per.problem, errors/regions*100,
                    color=chunks, size=chunks),
                data=data.frame(res.error, chunks="this"))+
      geom_line(aes(bases.per.problem, errors/regions*100,
                    color=chunks, size=chunks),
                data=data.frame(PSJ$error.total.all, chunks="all")),

    modelSelection=ggplot()+
      geom_segment(aes(min.log.lambda, peaks,
                       xend=max.log.lambda, yend=peaks,
                       showSelected=problem.name,
                       showSelected2=bases.per.problem),
                   data=data.frame(all.modelSelection, what="peaks"),
                   size=5)+
      geom_text(aes(min.log.lambda, peaks,
                    showSelected=problem.name,
                    showSelected2=bases.per.problem,
                    label=sprintf("%.1f kb in problem %s",
                    (problemEnd-problemStart)/1e3, problem.name)),
                data=data.frame(modelSelection.labels, what="peaks"))+
      geom_segment(aes(min.log.lambda, as.integer(errors),
                       xend=max.log.lambda, yend=as.integer(errors),
                       showSelected=problem.name,
                       showSelected2=bases.per.problem),
                   data=data.frame(modelSelection.errors, what="errors"),
                   size=5)+
      ggtitle("select number of samples with 1 peak")+
      ylab("")+
      facet_grid(what ~ ., scales="free"),

    title="Animint compiler with for loops",

    first=PSJ$first)

  ## For every problem there is a selector (called problem.dot) for the
  ## number of peaks in that problem. So in this for loop we add a few
  ## layers with aes_string(clickSelects=problem.dot) or
  ## aes_string(showSelected=problem.dot) to the coverage and
  ## modelSelection plots.
  for(problem.dot in names(PSJ$modelSelection.by.problem)){
    regions.dt <- PSJ$regions.by.problem[[problem.dot]]
    regions.dt[[problem.dot]] <- regions.dt$peaks
    if(!is.null(regions.dt)){
      viz.for$coverage <- viz.for$coverage+
        geom_tallrect(aes_string(xmin="chromStart/1e3",
                                 xmax="chromEnd/1e3",
                                 linetype="status",
                                 showSelected=problem.dot,
                                 showSelected2="bases.per.problem"),
                      data=data.frame(regions.dt),
                      fill=NA,
                      color="black")
    }
    if(problem.dot %in% names(PSJ$peaks.by.problem)){
      peaks <- PSJ$peaks.by.problem[[problem.dot]]
      peaks[[problem.dot]] <- peaks$peaks
      prob.peaks.names <-
        c("bases.per.problem", "problem.i", "problem.name",
          "chromStart", "chromEnd", problem.dot)
      prob.peaks <- unique(data.frame(peaks)[, prob.peaks.names])
      prob.peaks$sample.id <- "problems"
      viz.for$coverage <- viz.for$coverage +
        geom_segment(aes_string("chromStart/1e3", "0",
                                xend="chromEnd/1e3", yend="0",
                                clickSelects="problem.name",
                                showSelected=problem.dot,
                                showSelected2="bases.per.problem"),
                     data=peaks, size=7, color="deepskyblue")+
        geom_segment(aes_string("chromStart/1e3", "problem.i",
                                xend="chromEnd/1e3", yend="problem.i",
                                clickSelects="problem.name",
                                showSelected=problem.dot,
                                showSelected2="bases.per.problem"),
                     data=prob.peaks, size=7, color="deepskyblue")
    }
    modelSelection.dt <- PSJ$modelSelection.by.problem[[problem.dot]]
    modelSelection.dt[[problem.dot]] <- modelSelection.dt$peaks
    viz.for$modelSelection <- viz.for$modelSelection+
      geom_tallrect(aes_string(xmin="min.log.lambda", 
                               xmax="max.log.lambda", 
                               clickSelects=problem.dot,
                               showSelected="problem.name",
                               showSelected2="bases.per.problem"),
                    data=modelSelection.dt, alpha=0.5)
  }
}))
```

Note the timing of the code above. It takes over a minute just to
evaluate the R code that defines this data viz, since it has so many
geoms. Next, we compile the data visualization.

```{r Ch14-for-compile}
print(timing.for.compile <- system.time({
  animint2dir(viz.for, "Ch14-for")
}))
```

Note that the compilation also takes over a minute, since there are so
many geoms. The data viz can be viewed on
[Ch14-for/index.html](Ch14-for/index.html). In the next section we
will create the same data viz, but more efficiently.

## Define data viz using variable and value aesthetics

In this section we use `.variable` and `.value` aesthetics to create a
more efficient version of the previous data visualization. In general,
any data visualization defined using for loops in R code can be made
more efficient by instead using `.variable` and `.value` aesthetics.

```{r Ch14-viz-var-val}
sample.peaks <- do.call(rbind, PSJ$peaks.by.problem)
prob.peaks.names <-
  c("bases.per.problem", "problem.i", "problem.name", "peaks",
    "chromStart", "chromEnd")
problem.peaks <- unique(data.frame(sample.peaks)[, prob.peaks.names])
problem.peaks$sample.id <- "problems"

peakvar <- function(position){
  paste0(gsub("[-:]", ".", position), "peaks")
}

print(timing.var.val.construct <- system.time({
  viz.var.val <-
    list(coverage=ggplot()+
           geom_segment(aes(chromStart/1e3, problem.i,
                            xend=chromEnd/1e3, yend=problem.i,
                            showSelected=bases.per.problem,
                            clickSelects=problem.name),
                        data=prob.regions)+
           ggtitle("select problem")+
           geom_text(aes(chromStart/1e3, problem.i,
                         showSelected=bases.per.problem,
                         label=sprintf("%d problems mean size %.1f kb",
                           problems, mean.bases/1e3)),
                     data=PSJ$problem.labels,
                     hjust=0)+
           geom_segment(aes(problemStart/1e3, problem.i,
                            showSelected=bases.per.problem,
                            clickSelects=problem.name,
                            xend=problemEnd/1e3, yend=problem.i),
                        size=5,
                        data=PSJ$problems)+
           scale_y_continuous("aligned read coverage",
                              breaks=function(limits){
                                floor(limits[2])
                              })+
           scale_linetype_manual("error type",
                                 limits=c("correct", 
                                   "false negative",
                                   "false positive"
                                          ),
                                 values=c(correct=0,
                                   "false negative"=3,
                                   "false positive"=1))+
           scale_x_continuous(paste("position on chr11",
                                    "(kilo bases = kb)"))+
           coord_cartesian(xlim=c(118167.406, 118238.833))+
           geom_tallrect(aes(xmin=chromStart/1e3, xmax=chromEnd/1e3,
                             fill=annotation),
                         alpha=0.5,
                         color="grey",
                         data=PSJ$filled.regions)+
           scale_fill_manual(values=ann.colors)+
           theme_bw()+
           theme_animint(width=1500, height=facet.rows*100)+
           theme(panel.margin=grid::unit(0, "cm"))+
           facet_grid(sample.id ~ ., labeller=function(var, val){
             sub("McGill0", "", sub(" ", "\n", val))
           }, scales="free")+
           geom_line(aes(base/1e3, count),
                     data=PSJ$coverage,
                     color="grey50")+
           geom_tallrect(aes(xmin=chromStart/1e3,
                             xmax=chromEnd/1e3,
                             linetype=status,
                             showSelected.value=peaks,
                             showSelected.variable=peakvar(problem.name),
                             showSelected2=bases.per.problem),
                         data=all.regions,
                         fill=NA,
                         color="black")+
           geom_segment(aes(chromStart/1e3, 0,
                            xend=chromEnd/1e3, yend=0,
                            clickSelects=problem.name,
                            showSelected.variable=peakvar(problem.name),
                            showSelected.value=peaks,
                            showSelected2=bases.per.problem),
                        data=sample.peaks, size=7, color="deepskyblue")+
           geom_segment(aes(chromStart/1e3, problem.i,
                            xend=chromEnd/1e3, yend=problem.i,
                            clickSelects=problem.name,
                            showSelected.variable=peakvar(problem.name),
                            showSelected.value=peaks,
                            showSelected2=bases.per.problem),
                        data=problem.peaks, size=7, color="deepskyblue"),

         resError=ggplot()+
           ggtitle("select problem size")+
           ylab("minimum percent incorrect regions")+
           geom_tallrect(aes(xmin=min.bases.per.problem,
                             xmax=max.bases.per.problem,
                             clickSelects=bases.per.problem),
                         alpha=0.5,
                         data=res.error)+
           scale_x_log10()+
           geom_line(aes(bases.per.problem, errors/regions*100,
                         color=chunks, size=chunks),
                     data=data.frame(res.error, chunks="this"))+
           geom_line(aes(bases.per.problem, errors/regions*100,
                         color=chunks, size=chunks),
                     data=data.frame(PSJ$error.total.all, chunks="all")),

         modelSelection=ggplot()+
           geom_segment(aes(min.log.lambda, peaks,
                            xend=max.log.lambda, yend=peaks,
                            showSelected=problem.name,
                            showSelected2=bases.per.problem),
                        data=data.frame(all.modelSelection, what="peaks"),
                        size=5)+
           geom_text(aes(min.log.lambda, peaks,
                         showSelected=problem.name,
                         showSelected2=bases.per.problem,
                         label=sprintf("%.1f kb in problem %s",
                           (problemEnd-problemStart)/1e3, problem.name)),
                     data=data.frame(modelSelection.labels, what="peaks"))+
           geom_segment(aes(min.log.lambda, as.integer(errors),
                            xend=max.log.lambda, yend=as.integer(errors),
                            showSelected=problem.name,
                            showSelected2=bases.per.problem),
                        data=data.frame(modelSelection.errors, what="errors"),
                        size=5)+
           ggtitle("select number of samples with 1 peak")+
           ylab("")+
           geom_tallrect(aes(xmin=min.log.lambda, 
                             xmax=max.log.lambda, 
                             clickSelects.variable=
                               peakvar(problem.name),
                             clickSelects.value=peaks,
                             showSelected=problem.name,
                             showSelected2=bases.per.problem),
                         data=all.modelSelection, alpha=0.5)+
           facet_grid(what ~ ., scales="free"),
         
         title="Animint compiler with .variable .value aesthetics",

         first=PSJ$first)
### For every problem there is a selector (called problem.name) for
### the number of peaks in that problem. The animint2dir compiler
### creates a selection variable for every unique value of
### clickSelects.variable and showSelected.variable (and it uses
### clickSelects.value and showSelected.value to set/update the
### selected value/geoms).
}))
```

It is clear that it takes much less time to evaluate the R code above
which uses the `.variable` and `.value` aesthetics. We 

```{r Ch14-var-val-compile}
print(timing.var.val.compile <- system.time({
  animint2dir(viz.var.val, "Ch14-var-val")
}))
```

The animint produced above can be viewed on
[Ch14-var-val/index.html](Ch14-var-val/index.html). Note that it
should appear to be the same as the other data viz above. The timings
above show that `.variable` and `.value` aesthetics are much faster
than for loops, in both the definition and compilation steps.

## Disk usage comparison

In this section we compute the disk usage of both methods.

```{r}
library(data.table)
viz.dirs.vec <- c("Ch14-for", "Ch14-var-val")
viz.dirs.text <- paste(viz.dirs.vec, collapse=" ")
(cmd <- paste("du -ks", viz.dirs.text))
kb.dt <- fread(cmd)
setnames(kb.dt, c("kilobytes", "path"))
kb.dt
```

The table above shows that the data viz defined using for loops takes
about twice as much disk space as the data viz that used `.variable`
and `.value` aesthetics.

## Chapter summary and exercises

The table below summarizes the disk usage and timings presented in
this chapter. It is clear that `.variable` and `.value` aesthetics are
more efficient in both respects, and should be used instead of for
loops.

```{r}
data.frame(
  kilobytes=kb.dt$kilobytes,
  construct.seconds=c(
    timing.for.construct[["elapsed"]],
    timing.var.val.construct[["elapsed"]]),
  compile.seconds=c(
    timing.for.compile[["elapsed"]],
    timing.var.val.compile[["elapsed"]]),
  row.names=c("for.loop", "variable.value"))
```

Exercises:

* Use `.variable` and `.value` aesthetics to create a visualization
  of some data from your domain of expertise.
