---
title: Named clickSelects/showSelected
layout: default
output: bookdown::html_chapter
---

# Chapter 14, Named clickSelects/showSelected

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch14-figures/")
```

This chapter explains how to use
[named clickSelects/showSelected variables](Ch06-other.html#data-driven-selectors) for
creating data-driven selector names. This feature makes it easier to
write animint code, and makes it faster to compile.

Chapter outline:

* We begin by downloading the PSJ data set and computing the data to
  plot.
* We show one method of defining an animint with many selectors, using
  for loops. This method is technically correct, but computationally
  inefficient.
* We then explain the preferred method for defining an animint with
  many selectors, using named clickSelects/showSelected. This
  method is more computationally efficient, and easier to code.

## Download data set {#download}

The example data come from the
[PeakSegJoint package](https://github.com/tdhock/PeakSegJoint). The
code below downloads the data set.

```{r download-PSJ}
if(!file.exists("PSJ.RData")){
  u <- paste0(
    "http://github.com/tdhock/animint-examples",
    "/blob/master/data/PSJ.RData?raw=true"
    )
  library(httr)
  request <- GET(u)
  stop_for_status(request)
  writeBin(content(request), "PSJ.RData")
}
load("PSJ.RData")
```

## Compute data to plot {#compute-data-to-plot}

The section below computes some common data that we will use in two
data visualizations below.

```{r Ch14-setup-data}
res.error <- PSJ$error.total.chunk
ann.colors <- c(
  noPeaks="#f6f4bf",
  peakStart="#ffafaf",
  peakEnd="#ff4c4c",
  peaks="#a445ee")
## prob.regions are the black segments that show which regions are
## mapped to which segmentation problems.
library(data.table)
all.regions <- data.table(do.call(rbind, PSJ$regions.by.problem))
prob.regions.names <- c(
  "bases.per.problem", "problem.i", "problem.name",
  "chromStart", "chromEnd")
prob.regions <- unique(data.frame(all.regions)[, prob.regions.names])
prob.regions$sample.id <- "problems"
all.modelSelection <- data.table(do.call(
  rbind, PSJ$modelSelection.by.problem))
modelSelection.errors <- all.modelSelection[!is.na(errors)]
penalty.range <- all.modelSelection[, c(
  min(max.log.lambda), max(min.log.lambda))]
penalty.mid <- mean(penalty.range)
coverage.counts <- table(PSJ$coverage$sample.id)
facet.rows <- length(coverage.counts)+1
dvec <- diff(log(res.error$bases.per.problem))
dval <- exp(mean(dvec))
dval2 <- (dval-1)/2 + 1
res.error$min.bases.per.problem <- res.error$bases.per.problem/dval2
res.error$max.bases.per.problem <- res.error$bases.per.problem*dval2
modelSelection.labels <- unique(all.modelSelection[, data.table(
  problem.name=problem.name,
  bases.per.problem=bases.per.problem,
  problemStart=problemStart,
  problemEnd=problemEnd,
  min.log.lambda=penalty.mid,
  peaks=max(peaks)+0.5)])
```

## Define data viz using for loops {#define-using-for-loops}

The R code below constructs a data viz using for loops. 

```{r Ch14-for-construct}
library(animint2)
print(timing.for.construct <- system.time({
  viz.for <- list(
    coverage=ggplot()+
      geom_segment(aes(
        chromStart/1e3, problem.i,
        xend=chromEnd/1e3, yend=problem.i),
        showSelected="bases.per.problem",
        clickSelects="problem.name",
        data=prob.regions)+
      ggtitle("select problem")+
      geom_text(aes(
        chromStart/1e3, problem.i,
        label=sprintf("%d problems mean size %.1f kb",
                      problems, mean.bases/1e3)),
        showSelected="bases.per.problem",
        data=PSJ$problem.labels,
        hjust=0)+
      geom_segment(aes(
        problemStart/1e3, problem.i,
        xend=problemEnd/1e3, yend=problem.i),
        showSelected="bases.per.problem",
        clickSelects="problem.name",
        size=5,
        data=PSJ$problems)+
      scale_y_continuous(
        "aligned read coverage",
        breaks=function(limits){
          floor(limits[2])
        })+
      scale_linetype_manual(
        "error type",
        limits=c(
          "correct", 
          "false negative",
          "false positive"),
        values=c(
          correct=0,
          "false negative"=3,
          "false positive"=1))+
      scale_x_continuous(paste(
        "position on chr11",
        "(kilo bases = kb)"))+
      coord_cartesian(xlim=c(118167.406, 118238.833))+
      geom_tallrect(aes(
        xmin=chromStart/1e3, xmax=chromEnd/1e3,
        fill=annotation),
        alpha=0.5,
        color="grey",
        data=PSJ$filled.regions)+
      scale_fill_manual(values=ann.colors)+
      theme_bw()+
      theme_animint(width=1500, height=facet.rows*100)+
      theme(panel.margin=grid::unit(0, "cm"))+
      facet_grid(sample.id ~ ., labeller=function(df){
        df$sample.id <- sub("McGill0", "", sub(" ", "\n", df$sample.id))
        df
      }, scales="free")+
      geom_line(aes(
        base/1e3, count),
        data=PSJ$coverage,
        color="grey50"),
    resError=ggplot()+
      ggtitle("select problem size")+
      ylab("minimum percent incorrect regions")+
      geom_tallrect(aes(
        xmin=min.bases.per.problem,
        xmax=max.bases.per.problem),
        clickSelects="bases.per.problem",
        alpha=0.5,
        data=res.error)+
      scale_x_log10()+
      geom_line(aes(
        bases.per.problem, errors/regions*100,
        color=chunks, size=chunks),
        data=data.frame(res.error, chunks="this"))+
      geom_line(aes(
        bases.per.problem, errors/regions*100,
        color=chunks, size=chunks),
        data=data.frame(PSJ$error.total.all, chunks="all")),
    modelSelection=ggplot()+
      geom_segment(aes(
        min.log.lambda, peaks,
        xend=max.log.lambda, yend=peaks),
        showSelected=c("bases.per.problem", "problem.name"),
        data=data.frame(all.modelSelection, what="peaks"),
        size=5)+
      geom_text(aes(
        min.log.lambda, peaks,
        label=sprintf(
          "%.1f kb in problem %s",
          (problemEnd-problemStart)/1e3, problem.name)),
        showSelected=c("problem.name", "bases.per.problem"),
        data=data.frame(modelSelection.labels, what="peaks"))+
      geom_segment(aes(
        min.log.lambda, as.integer(errors),
        xend=max.log.lambda, yend=as.integer(errors)),
        showSelected=c("bases.per.problem", "problem.name"),
        data=data.frame(modelSelection.errors, what="errors"),
        size=5)+
      ggtitle("select number of samples with 1 peak")+
      ylab("")+
      facet_grid(what ~ ., scales="free"),
    title="Animint compiler with for loops",
    first=PSJ$first)
  ## For every problem there is a selector (called problem.dot) for the
  ## number of peaks in that problem. So in this for loop we add a few
  ## layers with aes_string(clickSelects=problem.dot) or
  ## aes_string(showSelected=problem.dot) to the coverage and
  ## modelSelection plots.
  for(problem.dot in names(PSJ$modelSelection.by.problem)){
    regions.dt <- PSJ$regions.by.problem[[problem.dot]]
    regions.dt[[problem.dot]] <- regions.dt$peaks
    if(!is.null(regions.dt)){
      viz.for$coverage <- viz.for$coverage+
        geom_tallrect(aes(
          xmin=chromStart/1e3,
          xmax=chromEnd/1e3,
          linetype=status),
          showSelected=c(problem.dot, "bases.per.problem"),
          data=data.frame(regions.dt),
          fill=NA,
          color="black")
    }
    if(problem.dot %in% names(PSJ$peaks.by.problem)){
      peaks <- PSJ$peaks.by.problem[[problem.dot]]
      peaks[[problem.dot]] <- peaks$peaks
      prob.peaks.names <- c(
        "bases.per.problem", "problem.i", "problem.name",
        "chromStart", "chromEnd", problem.dot)
      prob.peaks <- unique(data.frame(peaks)[, prob.peaks.names])
      prob.peaks$sample.id <- "problems"
      viz.for$coverage <- viz.for$coverage +
        geom_segment(aes(
          chromStart/1e3, 0,
          xend=chromEnd/1e3, yend=0),
          clickSelects="problem.name",
          showSelected=c(problem.dot, "bases.per.problem"),
          data=peaks, size=7, color="deepskyblue")+
        geom_segment(aes(
          chromStart/1e3, problem.i,
          xend=chromEnd/1e3, yend=problem.i),
          clickSelects="problem.name",
          showSelected=c(problem.dot, "bases.per.problem"),
          data=prob.peaks, size=7, color="deepskyblue")
    }
    modelSelection.dt <- PSJ$modelSelection.by.problem[[problem.dot]]
    modelSelection.dt[[problem.dot]] <- modelSelection.dt$peaks
    viz.for$modelSelection <- viz.for$modelSelection+
      geom_tallrect(aes(
        xmin=min.log.lambda, 
        xmax=max.log.lambda), 
        clickSelects=problem.dot,
        showSelected=c("problem.name", "bases.per.problem"),
        data=modelSelection.dt, alpha=0.5)
  }
}))
```

Note the timing of the code above. It takes a long time just to
evaluate the R code that defines this data viz, since it has so many
geoms. Next, we compile the data visualization.

```{r Ch14-for-compile}
print(timing.for.compile <- system.time({
  animint2dir(viz.for, "Ch14-for")
}))
```

Note that the compilation also takes a long time, since there are so
many geoms. The data viz can be viewed on
[Ch14-for/index.html](Ch14-for/index.html). In the next section we
will create the same data viz, but more efficiently.

## Define data viz using named clickSelects/showSelected {#define-using-named}

In this section we use named clickSelects/showSelected to create a
more efficient version of the previous data visualization. In general,
any data visualization defined using for loops in R code can be made
more efficient by instead using this method.

```{r Ch14-viz-named}
sample.peaks <- data.table(do.call(rbind, PSJ$peaks.by.problem))
prob.peaks.names <- c(
  "bases.per.problem", "problem.i", "problem.name", "peaks",
  "chromStart", "chromEnd")
problem.peaks <- unique(sample.peaks[, ..prob.peaks.names])
problem.peaks$sample.id <- "problems"
peakvar <- function(position){
  paste0(gsub("[-:]", ".", position), "peaks")
}
all.regions[, selector := peakvar(problem.name)]
sample.peaks[, selector := peakvar(problem.name)]
problem.peaks[, selector := peakvar(problem.name)]
all.modelSelection[, selector := peakvar(problem.name)]
print(timing.named.construct <- system.time({
  viz.named <- list(
    coverage=ggplot()+
      ggtitle("select problem")+
      geom_segment(aes(
        chromStart/1e3, problem.i,
        xend=chromEnd/1e3, yend=problem.i),
        showSelected="bases.per.problem",
        clickSelects="problem.name",
        data=prob.regions)+
      geom_text(aes(
        chromStart/1e3, problem.i,
        label=sprintf(
          "%d problems mean size %.1f kb",
          problems, mean.bases/1e3)),
        showSelected="bases.per.problem",
        data=PSJ$problem.labels,
        hjust=0)+
      geom_segment(aes(
        problemStart/1e3, problem.i,
        xend=problemEnd/1e3, yend=problem.i),
        showSelected="bases.per.problem",
        clickSelects="problem.name",
        size=5,
        data=PSJ$problems)+
      scale_y_continuous(
        "aligned read coverage",
        breaks=function(limits){
          floor(limits[2])
        })+
      scale_linetype_manual(
        "error type",
        limits=c(
          "correct", 
          "false negative",
          "false positive"),
        values=c(
          correct=0,
          "false negative"=3,
          "false positive"=1))+
      scale_x_continuous(paste(
        "position on chr11",
        "(kilo bases = kb)"))+
      coord_cartesian(xlim=c(118167.406, 118238.833))+
      geom_tallrect(aes(
        xmin=chromStart/1e3, xmax=chromEnd/1e3,
        fill=annotation),
        alpha=0.5,
        color="grey",
        data=PSJ$filled.regions)+
      scale_fill_manual(values=ann.colors)+
      theme_bw()+
      theme_animint(width=1500, height=facet.rows*100)+
      theme(panel.margin=grid::unit(0, "cm"))+
      facet_grid(sample.id ~ ., labeller=function(df){
        df$sample.id <- sub("McGill0", "", sub(" ", "\n", df$sample.id))
        df
      }, scales="free")+
      geom_line(aes(
        base/1e3, count),
        data=PSJ$coverage,
        color="grey50")+
      geom_tallrect(aes(
        xmin=chromStart/1e3,
        xmax=chromEnd/1e3,
        linetype=status),
        showSelected=c("selector"="peaks", "bases.per.problem"),
        data=all.regions,
        fill=NA,
        color="black")+
      geom_segment(aes(
        chromStart/1e3, 0,
        xend=chromEnd/1e3, yend=0),
        clickSelects="problem.name",
        showSelected=c("selector"="peaks", "bases.per.problem"),
        data=sample.peaks, size=7, color="deepskyblue")+
      geom_segment(aes(
        chromStart/1e3, problem.i,
        xend=chromEnd/1e3, yend=problem.i),
        clickSelects="problem.name",
        showSelected=c("selector"="peaks", "bases.per.problem"),
        data=problem.peaks, size=7, color="deepskyblue"),
    resError=ggplot()+
      ggtitle("select problem size")+
      ylab("minimum percent incorrect regions")+
      geom_tallrect(aes(
        xmin=min.bases.per.problem,
        xmax=max.bases.per.problem),
        clickSelects="bases.per.problem",
        alpha=0.5,
        data=res.error)+
      scale_x_log10()+
      geom_line(aes(
        bases.per.problem, errors/regions*100,
        color=chunks, size=chunks),
        data=data.frame(res.error, chunks="this"))+
      geom_line(aes(
        bases.per.problem, errors/regions*100,
        color=chunks, size=chunks),
        data=data.frame(PSJ$error.total.all, chunks="all")),
    modelSelection=ggplot()+
      geom_segment(aes(
        min.log.lambda, peaks,
        xend=max.log.lambda, yend=peaks),
        showSelected=c("problem.name", "bases.per.problem"),
        data=data.frame(all.modelSelection, what="peaks"),
        size=5)+
      geom_text(aes(
        min.log.lambda, peaks,
        label=sprintf(
          "%.1f kb in problem %s",
          (problemEnd-problemStart)/1e3, problem.name)),
        showSelected=c("problem.name", "bases.per.problem"),
        data=data.frame(modelSelection.labels, what="peaks"))+
      geom_segment(aes(
        min.log.lambda, as.integer(errors),
        xend=max.log.lambda, yend=as.integer(errors)),
        showSelected=c("problem.name", "bases.per.problem"),
        data=data.frame(modelSelection.errors, what="errors"),
        size=5)+
      ggtitle("select number of samples with 1 peak")+
      ylab("")+
      geom_tallrect(aes(
        xmin=min.log.lambda, 
        xmax=max.log.lambda), 
        clickSelects=c("selector"="peaks"),
        showSelected=c("problem.name", "bases.per.problem"),
        data=all.modelSelection, alpha=0.5)+
      facet_grid(what ~ ., scales="free"),
    title="Animint compiler with named clickSelects/showSelected",
    first=PSJ$first)
### For every problem there is a selector (called problem.name) for
### the number of peaks in that problem. The animint2dir compiler
### creates a selection variable for every unique value of
### clickSelects/showSelected names (and it uses corresponding values
### to set/update the selected value/geoms).
}))
```

It is clear that it takes much less time to evaluate the R code above
which uses the named clickSelects/showSelected. We compile it below.

```{r Ch14-named-compile}
print(timing.named.compile <- system.time({
  animint2dir(viz.named, "Ch14-named")
}))
```

The animint produced above can be viewed on
[Ch14-named/index.html](Ch14-named/index.html). Note that it
should appear to be the same as the other data viz above. The timings
above show that named clickSelects/showSelected are much faster
than for loops, in both the definition and compilation steps.

## Disk usage comparison {#disk-usage}

In this section we compute the disk usage of both methods.

```{r}
viz.dirs.vec <- c("Ch14-for", "Ch14-named")
viz.dirs.text <- paste(viz.dirs.vec, collapse=" ")
(cmd <- paste("du -ks", viz.dirs.text))
kb.dt <- fread(cmd)
setnames(kb.dt, c("kilobytes", "path"))
kb.dt
```

The table above shows that the data viz defined using for loops takes
about twice as much disk space as the data viz that used named
clickSelects/showSelected.

## Chapter summary and exercises {#exercises}

The table below summarizes the disk usage and timings presented in
this chapter. It is clear that named clickSelects/showSelected are
more efficient in both respects, and should be used instead of for
loops.

```{r}
data.frame(
  kilobytes=kb.dt$kilobytes,
  construct.seconds=c(
    timing.for.construct[["elapsed"]],
    timing.named.construct[["elapsed"]]),
  compile.seconds=c(
    timing.for.compile[["elapsed"]],
    timing.named.compile[["elapsed"]]),
  row.names=c("for", "named"))
```

Exercises:

* Use named clickSelects/showSelected to create a visualization
  of some data from your domain of expertise.

Next, [Chapter 15](Ch15-Newton.html) explains how to visualize
root-finding algorithms.
