---
title: Change-point detection
layout: default
output: bookdown::html_chapter
---

# Chapter 16, Change-point detection

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch16-figures/")
```

In this chapter we will explore several data visualizations of
supervised changepoint detection models.

Chapter outline:

* We begin by making several static visualizations of the `intreg`
  data set.
* We then create an interactive visualization in which one plot can be
  click to select the number of changepoints/segments, and the other
  plot shows the corresponding model.
* We end by showing a static visualization of the 

## Static figures {#static}

We begin by loading the `intreg` data set.

```{r intreg}
library(animint2)
data(intreg)
str(intreg)
```

As shown above, it is a named list of 7 related data.frames. We begin
our exploration of these data by plotting the signals in separate
facets.

```{r signals}
data.color <- "grey50"
gg.signals <- ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(signal ~ ., scales="free")+
  geom_point(aes(
    base/1e6, logratio,
    showSelected="signal"),
    color=data.color,
    data=intreg$signals)
print(gg.signals)
```

Each data point plotted above shows an approximate measurement of DNA
copy number (logratio), as a function of base position on a
chromosome. Such data come from high-throughput assays which are
important for diagnosing certain types of cancer such as
neuroblastoma. 

An important part of the diagnosis is detecting "breakpoints" or
abrupt changes, within a given chromosome (panel). It is clear from
the plot above that there are several breakpoint in these data. In
particular signal 4.2 appears to have three breakpoints, signal 4.3
appears to have one, etc. In fact these data come from medical doctors
at the Institute Curie (Paris, France) who have visually annotated
regions with and without breakpoints. These data are available as
`intreg$annotations` and are plotted below.

```{r annotations}
breakpoint.colors <- c("1breakpoint"="#ff7d7d", "0breakpoints"='#f6f4bf')
gg.ann <- gg.signals+
  scale_fill_manual(values=breakpoint.colors)+
  geom_tallrect(aes(
    xmin=first.base/1e6, xmax=last.base/1e6,
    fill=annotation),
    color="grey",
    alpha=0.5,
    data=intreg$annotations)
print(gg.ann)
```

The plot above shows yellow regions where the doctors have determined
that there are no significant breakpoints, and red regions where there
is one breakpoint. The goal in analyzing these data is to learn from
the limited labeled data (colored regions) and provide consistent
breakpoint predictions throughout (even in un-labeled regions).

In order to detect these breakpoints we have fit some maximum
likelihood segmentation models, using the efficient algorithm
implemented in `jointseg::Fpsn`. The segment means are available in
`intreg$segments` and the predicted breakpoints are available in
`intreg$breaks`. For each signal there is a sequence of models from 1
to 20 segments. First let's zoom in on one signal:

```{r one}
sig.name <- "4.2"
show.segs <- 7
sig.labels <- subset(intreg$annotations, signal==sig.name)
gg.one <- ggplot()+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  geom_tallrect(aes(
    xmin=first.base/1e6, xmax=last.base/1e6,
    fill=annotation),
    color="grey",
    alpha=0.5,
    data=sig.labels)+
  geom_point(aes(
    base/1e6, logratio),
    color=data.color,
    data=subset(intreg$signals, signal==sig.name))+
  scale_fill_manual(values=breakpoint.colors)
print(gg.one)
```

We plot some of these models for one of the signals below:

```{r models}
library(data.table)
sig.segs <- data.table(
  intreg$segments)[signal == sig.name & segments <= show.segs]
sig.breaks <- data.table(
  intreg$breaks)[signal == sig.name & segments <= show.segs]
model.color <- "green"
gg.models <- gg.one+
  facet_grid(segments ~ .)+
  geom_segment(aes(
    first.base/1e6, mean,
    xend=last.base/1e6, yend=mean),
    color=model.color,
    data=sig.segs)+
  geom_vline(aes(
    xintercept=base/1e6),
    color=model.color,
    linetype="dashed",
    data=sig.breaks)
print(gg.models)
```

The plot above shows the maximum likelihood segmentation models in
green (from one to six segments). Below we use the
`penaltyLearning::labelError` function to compute the label error,
which quantifies which models agree with which labels.

```{r labelerr}
sig.models <- data.table(segments=1:show.segs, signal=sig.name)
sig.errors <- penaltyLearning::labelError(
  sig.models, sig.labels, sig.breaks,
  change.var="base",
  label.vars=c("first.base", "last.base"),
  model.vars="segments",
  problem.vars="signal")
```

The `sig.errors$label.errors` data.table contains one row for every
(model,label) combination. The `status` column can be used to show the
label error: `false negative` for too few changes, `false positive`
for too many changes, or `correct` for the right number of changes.

```{r plotlabelerr}
gg.models+
  geom_tallrect(aes(
    xmin=first.base/1e6, xmax=last.base/1e6,
    linetype=status),
    data=sig.errors$label.errors,
    color="black",
    size=1,
    fill=NA)+
  scale_linetype_manual(
    "error type",
    values=c(
      correct=0,
      "false negative"=3,
      "false positive"=1))
```

Looking at the label error plot above, it is clear that the model with
four segments should be selected, because it achieves zero label
errors. There are a number of criteria that
can be used to select which one of these models is best. One way to do
that is by selecting the model with $s$ segments is
$S^*(\lambda)=L_s + \lambda*s$, where $L_s$ is the total loss of the
model with $s$ segments, and $\lambda$ is a non-negative penalty. In
the plot below we show the model selection function $S^*(\lambda)$ for
this data set:

```{r selection}
sig.selection <- data.table(
  intreg$selection)[signal == sig.name & segments <= show.segs]
gg.selection <- ggplot()+
  theme_bw()+
  geom_segment(aes(
    min.L, segments,
    xend=max.L, yend=segments),
    data=sig.selection)+
  xlab("log(lambda)")
print(gg.selection)
```

It is clear from the plot above that the model selection function is
decreasing. In the next section we make an interactive version of
these two plots where we can actually click on the model selection
plot in order to select the model. 

## Interactive figures for one signal {#interactive-one}

We will create an interactive figure for one signal by adding a
`geom_tallrect` with `clickSelects=segments` to the plot above:

```{r selectionClick}
interactive.selection <- gg.selection+
  geom_tallrect(aes(
    xmin=min.L, xmax=max.L),
    clickSelects="segments",
    data=sig.selection,
    color=NA,
    fill="black",
    alpha=0.5)
print(interactive.selection)
```

We will combine that with the non-facetted version of the data/models
plot below, in which we have added `showSelected=segments` to the
model geoms:

```{r selectionData}
interactive.models <- gg.one+
  geom_segment(aes(
    first.base/1e6, mean,
    xend=last.base/1e6, yend=mean),
    showSelected="segments",
    color=model.color,
    data=sig.segs)+
  geom_vline(aes(
    xintercept=base/1e6),
    showSelected="segments",
    color=model.color,
    linetype="dashed",
    data=sig.breaks)+
  geom_tallrect(aes(
    xmin=first.base/1e6, xmax=last.base/1e6,
    linetype=status),
    showSelected="segments",
    data=sig.errors$label.errors,
    size=2,
    color="black",
    fill=NA)+
  scale_linetype_manual(
    "error type",
    values=c(
      correct=0,
      "false negative"=3,
      "false positive"=1))
print(interactive.models)
```

Of course the plot above is not very informative because it is not
interactive. Below we combine the two interactive ggplots in a single
linked animint:

```{r interactiveOne}
animint(
  models=interactive.models+
    ggtitle("Selected model"),
  selection=interactive.selection+
    ggtitle("Click to select number of segments"))
```

Note that in the data viz above the model with 6 segments is not
selectable for any value of lambda, so there is no way to click on the
plot to select that model. However it is possible to select the model
using the segments selection menu (click "Show selection menus" at the
bottom of the data viz).

## Static max margin regression plot {#max-margin}

Another part of this data set is `intreg$intervals` which has one row
for every signal. The columns `min.L` and `max.L` indicate the min/max
values of the target interval, which is the largest range of
log(penalty) values with minimum label errors. Below we plot this
interval as a function of a feature of the data (log number of data
points):

```{r intervals}
gg.intervals <- ggplot()+
  geom_segment(aes(
    feature, min.L,
    xend=feature, yend=max.L),
    size=2,
    data=intreg$intervals)+
  geom_text(aes(
    feature, min.L, label=signal,
    color=ifelse(signal==sig.name, "black", "grey50")),
    vjust=1,
    data=intreg$intervals)+
  scale_color_identity()+
  ylab("output log(lambda)")+
  xlab("input feature x")
print(gg.intervals)
```

The target intervals in the plot above denote the region of
log(lambda) space that will select a model with minimum label
errors. There is one interval for each signal; we made an animint in
the previous section for the signal indicated in black text. Machine
learning algorithms can be used to find a penalty function that
intersects each of the intervals, and maximizes the margin (the
distance between the regression function and the nearest interval
limit). Data for the linear max margin regression function are in
`intreg$model` which is shown in the plot below:

```{r maxMargin}
gg.mm <- gg.intervals+
  geom_segment(aes(
    min.feature, min.L,
    xend=max.feature, yend=max.L,
    linetype=line),
    color="red",
    size=1,
    data=intreg$model)+
  scale_linetype_manual(
    values=c(
      regression="solid",
      margin="dotted",
      limit="dashed"))
print(gg.mm)    
```

The plot above shows the linear max margin regression function f(x) as
the solid red line. It is clear that it intersects each of the black
target intervals, and maximizes the margin (red vertical dotted
lines). For more information on the subject of supervised changepoint
detection, please see
[my useR 2017 tutorial](https://github.com/tdhock/change-tutorial).

Now that you know how to visualize each of the seven parts of the
`intreg` data set, the rest of the chapter is devoted to exercises.

## Chapter summary and exercises {#exercises}

Exercises:

* Add a `geom_text` which shows the currently selected signal name at
  the top of the plot, in `interactive.models` in the first animint
  above.
* Make an animint with two plots that shows the data set that
  corresponds to each interval on the max margin regression plot. One
  plot should show an interactive version of the max margin regression
  plot where you can click on an interval to select a signal. The
  other plot should show the data set for the currently selected
  signal.
* In the animint you created in the previous exercise, add a third
  plot with the model selection function for the currently selected
  signal. 
* Re-design the previous animint so that instead of using a third
  plot, add a facet to the max margin regression regression plot such
  that the log(lambda) axes are aligned. Add another facet that shows
  the number of incorrect labels (`intreg$selection$cost`) for each
  log(lambda) value.
* Add geoms for selecting the number of segments. Clicking the model
  selection plot should select the number of segments, which should
  update the displayed model and label errors on the plot of the data
  for the currently selected signal. Furthermore add a visual
  indication of the selected model to the max margin regression plot.
  The result should look something
  [like this](http://bl.ocks.org/tdhock/raw/10f27e4ace80bffa10a0/).
* Make another data viz by starting with the facetted `gg.signals`
  plot in the beginning of this chapter. Add a plot that can be used
  to select the number of segments for each signal. For each signal in
  the facetted plot of the data, show the currently selected model for
  that signal (there should be a separate selection variable for each
  signal -- you can use named clickSelects/showSelected as explained in
  [Chapter 14](Ch14-PeakSegJoint.html)). The result should look
  something
  [like this](http://bl.ocks.org/tdhock/raw/9311ca39d643d127e04a088814c81ee1/).

Next, [Chapter 17](Ch17-appendix.html) explains some R programming
idioms that are generally useful for interactive data visualization
design.
