---
title: Appendix, useful idioms
layout: default
output: bookdown::html_chapter
---

# Appendix, useful idioms

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch17-figures/")
```

This appendix describes several idioms which are useful for creating
animints.

## Space saving facets {#space-saving-facets}

To emphasize the plotted data in facetted ggplots, eliminate the space
between facets using the following idiom.

```{r, eval=FALSE}
ggplot()+
  geom_point(aes(Petal.Width, Sepal.Width), iris)+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))+
  facet_grid(. ~ Species)
```

There are three parts of this idiom:

* `panel.margin=0` eliminates space between panels.
* `theme_bw` activates a black and white theme (black panel borders
  and white panel backgrounds). This is necessary in order to see the
  boundaries between panels, since the ggplot default `theme_grey`
  uses grey panel backgrounds and no panel borders.
* `facet_*` creates a multi-panel ggplot.

Note that we use the grid unit `lines`, which equals the height of one
line of text at the default size. This is the only grid unit which
animint knows how to translate. It is not recommended to use other
units such as `cm`.

## List of data tables {#list-of-data-tables}

The list of data tables idiom is very useful for creating interactive
data visualizations of arbitrary complexity. The general form looks like

```{r, eval=FALSE}
library(data.table)
outer.data.list <- list()
inner.data.list <- list()
for(outer in outer.vec){
  outer.dt <- computeOuter(outer)
  outer.data.list[[paste(outer)]] <- data.table(outer, outer.dt)
  for(inner in inner.vec){
    inner.dt <- computeInner(outer.dt, inner)
    inner.data.list[[paste(outer, inner)]] <-
      data.table(outer, inner, inner.dt)
  }
}
outer.data <- do.call(rbind, outer.data.list)
inner.data <- do.call(rbind, inner.data.list)
```

Some comments:

* The first part of the idiom involves initializing empty lists. Here
  there are two, `outer.data.list` and `inner.data.list`. However
  there can be as many as necessary.
* The second part of the idiom is a bunch of nested for loops that
  assign data tables to elements of those lists.
  * Functions like `computeOuter` and `computeInner` can be used, or
    you can just do the computations directly inside the for loop.
  * To ensure that your code will run as fast as possible, use
    matrix-vector or vector-scalar operations in the innermost for
    loop. If you only do scalar-scalar operations in your innermost
    for loop, then you can definitely improve the performance of your
    code by removing that for loop and re-writing the computation in
    terms of vector-scalar operations.
  * The `paste` function is used to assign a `data.table` to a named
    list element. Although in principle one could use either
    `data.frame` or `data.table`, in practice `data.table` is often
    much faster during the last combination step.
* The last part of the idiom uses `do.call` with `rbind` to combine
  the data tables stored during the for loops. 

## addColumn then facet {#addColumn-then-facet}

This idiom is useful for creating multi-panel ggplots with aligned
axes. First, define a function which takes as input a data table and
one or more values which will be used to add factors to that data
table.

```{r, eval=FALSE}
addColumn <- function(df, time.period){
  data.frame(df, time.period=factor(time.period, c("1975", "1960-2010")))
}
animint(
  ggplot()+
  geom_point(aes(
    x=life.expectancy, y=fertility.rate, color=region),
    data=addColumn(WorldBank1975, "1975"))+
  geom_path(aes(
    x=life.expectancy, y=fertility.rate, color=region,
    group=country),
    data=addColumn(WorldBankBefore1975, "1975"))+
  geom_line(aes(
    x=year, y=fertility.rate, color=region, group=country),
    data=addColumn(WorldBank, "1960-2010"))+
  facet_grid(. ~ time.period, scales="free")+
  xlab(""))
```

Note that `scales="free"` and `xlab("")` are used since the x axes now
have very different units (year and life expectancy).

## Manual color legends {#manual-color-legends}

Color and fill legends in `ggplot2` can be manually specified via
`scale_color_manual` and `scale_fill_manual`. Typically we will choose
one of the ColorBrewer palettes:

```{r, fig.height=8}
RColorBrewer::display.brewer.all()
```

For example to get the R code for the Set1 palette, we can write

```{r}
dput(RColorBrewer::brewer.pal(Inf, "Set1"))
```

We can then copy that R code from the terminal and paste it into our
text editor

```{r}
data(WorldBank, package="animint2")
region.colors <- 
  c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00", "#FFFF33", 
    "#A65628", "#F781BF", "#999999")
names(region.colors) <- levels(WorldBank$region)
region.colors
```

Then we can use it with `scale_color_manual` 

```{r}
library(animint2)
ggplot()+
  scale_color_manual(values=region.colors)+
  geom_point(aes(
    x=life.expectancy, y=fertility.rate, color=region),
    data=WorldBank)
```


