---
title: Clustering and k-means
author: Toby Dylan Hocking
output: bookdown::html_chapter
---

# Chapter 17, K-means clustering

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.path="Ch17-figures/")
```

In this chapter we will explore several data visualizations of
K-means clustering, which is an unsupervised learning algorithm.

Chapter outline:

* We begin by visualizing two features of the iris data.
* We choose three random data points to use as cluster centers.
* We show how all distances between data points and cluster centers
  can be computed and visualized.
* We end by showing a visualization of how the k-means model
  parameters change with each iteration.

## Visualize iris data with labels {#viz-iris}

We begin with a typical visualization of the iris data, including a
color legend to indicate the Species.

```{r}
library(animint2)
color.code <- c(
  setosa="#1B9E77",
  versicolor="#D95F02",
  virginica="#7570B3",
  "1"="#E7298A",
  "2"="#66A61E",
  "3"="#E6AB02", 
  "4"="#A6761D")
ggplot()+
  scale_color_manual(values=color.code)+
  geom_point(aes(
    Petal.Length, Petal.Width, color=Species),
    data=iris)+
  coord_equal()
```

We will illustrate the K-means clustering algorithm using these two
dimensions,

```{r results=TRUE}
data.mat <- as.matrix(iris[,c("Petal.Width","Petal.Length")])
head(data.mat)
str(data.mat)
```

# K-means starts with three random cluster centers {#kmeans-init}

To run K-means, the number of clusters hyper-parameter (K) must be
fixed in advance. Then K random data points are selected as the
initial cluster centers,

```{r}
K <- 3
library(data.table)
data.dt <- data.table(data.mat)
set.seed(3)
centers.dt <- data.dt[sample(1:.N, K)]
(centers.mat <- as.matrix(centers.dt))
centers.dt[, cluster := factor(1:K)]
centers.dt
gg.centers <- ggplot()+
  scale_color_manual(values=color.code)+
  geom_point(aes(
    Petal.Length, Petal.Width),
    color="grey50",
    data=data.dt)+
  geom_point(aes(
    Petal.Length, Petal.Width, color=cluster),
    data=centers.dt)+
  coord_equal()
gg.centers
```

Above we displayed the two data sets (cluster centers and data) using
two instances of `geom_point`. Below we compute the distance between
each data point and each cluster center,

```{r}
pairs.dt <- data.table(expand.grid(
  centers.i=1:nrow(centers.mat),
  data.i=1:nrow(data.mat)))
```

These can be visualized via a `geom_point`,

```{r}
seg.dt <- pairs.dt[, data.table(
  data.i,
  data=data.mat[data.i,],
  center=centers.mat[centers.i,])]
gg.centers+
  geom_segment(aes(
    data.Petal.Length, data.Petal.Width,
    xend=center.Petal.Length, yend=center.Petal.Width),
    size=1,
    data=seg.dt)
```

There are `r nrow(seg.dt)` segments overplotted above, so
interactivity would be useful to emphasize the segments connected to a
particular data point. To do that we create a `data.i` selection
variable,

```{r Ch17-kmeans-distances}
animint(
  ggplot()+
    scale_color_manual(values=color.code)+
    geom_point(aes(
      Petal.Length, Petal.Width, color=cluster),
      size=4,
      data=centers.dt)+
    geom_point(aes(
      Petal.Length, Petal.Width),
      clickSelects="data.i",
      size=2,
      color="grey50",
      data=data.table(data.mat, data.i=1:nrow(data.mat)))+
    coord_equal()+
    geom_segment(aes(
      data.Petal.Length, data.Petal.Width,
      xend=center.Petal.Length, yend=center.Petal.Width),
      size=1,
      showSelected="data.i",
      data=seg.dt)
)
```

In the data viz above you can click on a data point to show the
distances from that data point to each cluster center.

Exercises for this section:

* edit the x/y scales so that the same ticks are shown.
* change the color of each segment so that it matches the
  corresponding cluster.
* add a tooltip that shows the distance value.
* make the segment width depend on its optimality (segment connected
  to closest cluster center should be emphasized with greater width).

## Visualizing iterations of algorithm {#viz-iterations}

Next we compute the closest cluster center for each data point,

```{r}
pairs.dt[, error := rowSums((data.mat[data.i,]-centers.mat[centers.i,])^2)]
(closest.dt <- pairs.dt[, .SD[which.min(error)], by=data.i])
(closest.data <- closest.dt[, .(
  data.dt[data.i],
  cluster=factor(centers.i)
)])
(both.dt <- rbind(
  data.table(type="centers", centers.dt),
  data.table(type="data", closest.data)))
ggplot()+
  scale_fill_manual(values=color.code)+
  scale_color_manual(values=c(centers="black", data="grey"))+
  scale_size_manual(values=c(centers=5, data=3))+
  geom_point(aes(
    Petal.Length, Petal.Width, fill=cluster, size=type, color=type),
    shape=21,
    data=both.dt)+
  coord_equal()+
  theme_bw()
```

Then we update the cluster centers,

```{r}
new.centers <- closest.dt[, data.table(
  t(colMeans(data.dt[data.i]))
), by=.(cluster=centers.i)]
(new.both <- rbind(
  data.table(type="centers", new.centers),
  data.table(type="data", closest.data)))
ggplot()+
  scale_fill_manual(values=color.code)+
  scale_color_manual(values=c(centers="black", data="grey"))+
  scale_size_manual(values=c(centers=5, data=3))+
  geom_point(aes(
    Petal.Length, Petal.Width, fill=cluster, size=type, color=type),
    shape=21,
    data=new.both)+
  coord_equal()+
  theme_bw()
```

So the visualizations above show the steps of k-means: (1) updating
cluster assignment based on closest center, then (2) updating center
based on data assigned to that cluster. To visualize several
iterations of the above two steps, we can use a for loop,

```{r}
set.seed(3)
centers.dt <- data.dt[sample(1:.N, K)]
(centers.mat <- as.matrix(centers.dt))
data.and.centers.list <- list()
iteration.error.list <- list()
for(iteration in 1:20){
  pairs.dt[, error := {
    rowSums((data.mat[data.i,]-centers.mat[centers.i,])^2)
  }]
  closest.dt <- pairs.dt[, .SD[which.min(error)], by=data.i]
  iteration.error.list[[iteration]] <- data.table(
    iteration, error=sum(closest.dt[["error"]]))
  iteration.both <- rbind(
    data.table(type="centers", centers.dt, cluster=1:K),
    closest.dt[, data.table(
      type="data", data.dt[data.i], cluster=factor(centers.i))])
  data.and.centers.list[[iteration]] <- data.table(
    iteration, iteration.both)
  new.centers <- closest.dt[, data.table(
    t(colMeans(data.dt[data.i]))
  ), keyby=.(cluster=centers.i)]
  centers.dt <- new.centers[, names(centers.dt), with=FALSE]
  centers.mat <- as.matrix(centers.dt)
}
(data.and.centers <- do.call(rbind, data.and.centers.list))
(iteration.error <- do.call(rbind, iteration.error.list))
```

First we create an overview plot with an error curve that will be used
to select the model size,

```{r}
gg.err <- ggplot()+
  theme_bw()+
  geom_point(aes(
    iteration, error),
    data=iteration.error)+
  make_tallrect(iteration.error, "iteration", alpha=0.3)
gg.err
```

We also make a plot which will show the current iteration,

```{r}
gg.iteration <- ggplot()+
  scale_fill_manual(values=color.code)+
  scale_color_manual(values=c(centers="black", data=NA))+
  scale_size_manual(values=c(centers=5, data=2))+
  geom_point(aes(
    Petal.Length, Petal.Width, fill=cluster, size=type, color=type),
    shape=21,
    showSelected="iteration",
    data=data.and.centers)+
  coord_equal()+
  theme_bw()
gg.iteration
```

Combining the two plots results in an interactive data viz,

```{r Ch17-kmeans-iterations}
animint(gg.err, gg.iteration)
```

## Chapter summary and exercises {#exercises}

Exercises:

* Make centers always show up in front (on top) of the data.
* Add smooth transitions.
* Add animation on iteration variable.
* Current code has fixed max number of iterations, so it is possible
  for the last few iterations to make no progress. For example in the
  viz above, iteration=16 was the last one that resulted in a decrease
  in error (iterations 17-20 resulted in no decrease). Modify the code
  so that it stops iterating if there is no decrease in error.
* Current viz has only one animation frame (showSelected subset) per
  iteration (the mean shown is before it is updated). Add another
  animation frame that shows the mean after the update.
* Add interactive segments that show the distance from each data point
  to each cluster center (as in first animint on this page).
* Add the features described in the exercises in the previous section
  on this page.
* Compute results for several different random seeds, then display
  error rates for each seed on the error overview plot, and allow the
  user to select any of those results.
* Compute results for several different numbers of clusters (K).
  Compute the Adjusted Rand Index using
  `pdfCluster::adj.rand.index(species, cluster)` for each different K
  and seed. Add an overview plot that shows the ARI value of each
  model, and allows selecting the number of clusters.
* Make a similar visualization using another data set such as
  `data("penguins", package="palmerpenguins")`.

Next, [Chapter 99](Ch99-appendix.html) explains some R
programming idioms that are generally useful for interactive data
visualization design.
