<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Cross-validation}
-->

# Cross-validation

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Comparing cross-validation methods

```{r}
library(data.table)
n.segments <- 5
set.seed(1)
(seg.mean.vec <- rnorm(n.segments))
data.per.segment <- 10
data.mean.vec <- rep(seg.mean.vec, each=data.per.segment)
n.data <- length(data.mean.vec)
total.validation.sets <- 100
n.folds.vec <- c(2, 5, 10)
prop.valid.vec <- 1/n.folds.vec
seg_dt_list <- list()
data_dt_list <- list()
residual_dt_list <- list()
loss_dt_list <- list()
selection_dt_list <- list()
selection_error_list <- list()
for(data_seed in 1:2){
  set.seed(data_seed)
  data.vec <- rnorm(n.data, data.mean.vec, 0.1)
  split_data <- data.table(
    data_seed, signal=data.vec,
    position=as.numeric(seq_along(data.vec))
  )[, pos := position]
  setkey(split_data, position, pos)
  data_dt_list[[data_seed]] <- split_data
  is.valid.vec.list <- list()
  for(n.folds in n.folds.vec){
    uniq.folds <- 1:n.folds
    n.seeds <- total.validation.sets/n.folds
    split_type <- sprintf("%d-fold", n.folds)
    for(cv_split_seed in 1:n.seeds){
      set.seed(cv_split_seed)
      fold.vec <- sample(rep(uniq.folds, l=n.data))
      for(valid.fold in uniq.folds){
        is.valid.vec.list[[split_type]][[paste(cv_split_seed, valid.fold)]] <-
          fold.vec==valid.fold
      }
    }
  }
  for(prop.valid in prop.valid.vec){
    split_type <- sprintf("%d%% valid", 100*prop.valid)
    prop.vec <- c(subtrain=1-prop.valid, validation=prop.valid)
    for(split_seed in 1:total.validation.sets){
      set.seed(split_seed)
      is.valid.vec.list[[split_type]][[split_seed]] <- binsegRcpp::random_set_vec(
        n.data, prop.vec) == "validation"
    }
  }
  for(split_type in names(is.valid.vec.list)){
    for(n_validation_sets in seq(5, total.validation.sets, by=5)){
      is.valid <- is.valid.vec.list[[split_type]][[n_validation_sets]]
      set(split_data, j="set", value=ifelse(is.valid, "validation", "subtrain"))
      bs.model <- split_data[, binsegRcpp::binseg_normal(signal, is.validation.vec=is.valid)]
      coef_dt <- coef(bs.model, 1:nrow(bs.model$splits))
      setkey(coef_dt, start.pos, end.pos)
      over_dt <- foverlaps(coef_dt, split_data)
      residual_dt_list[[paste(
        data_seed, split_type, n_validation_sets
      )]] <- over_dt[, .(
        data_seed, split_type, n_validation_sets,
        position, segments, signal, mean
      )]
      seg_dt_list[[paste(
        data_seed, split_type, n_validation_sets
      )]] <- data.table(
        data_seed, split_type, n_validation_sets,
        coef_dt
      )
      loss_dt_list[[paste(
        data_seed, split_type, n_validation_sets
      )]] <- data.table(
        data_seed, split_type, n_validation_sets,
        bs.model$splits[, rbind(
          data.table(set="subtrain", segments, loss),
          data.table(set="validation", segments, loss=validation.loss)
        )]
      )[, point := ifelse(loss==min(loss), "min", "other"), by=set][]
      selection_dt_list[[paste(
        data_seed, split_type, n_validation_sets
      )]] <- data.table(
        data_seed, split_type, n_validation_sets,
        segments=bs.model$splits[which.min(validation.loss), segments]
      )
    }
  }
}

sim.result <- do.call(rbind, sim.result.list)
(sim.err <- sim.result[, .(
  zero.one.loss=sum(selected != n.segments),
  L1.loss=sum(abs(selected-n.segments)),
  L2.loss=sum((selected-n.segments)^2)
), by=method][order(zero.one.loss)])


```

The code above compares several types of cross-validation for
selecting the number of segments in simulated random normal data. The
table above shows various error rates which compare the selected
number of segments to the true number of segments in the
simulation. The best methods appears to be the ones which use min err,
max times.

## How many times is sufficient?

```{r}
n.segments <- 20
seg.mean.vec <- 1:n.segments
data.per.segment <- 5
data.mean.vec <- rep(seg.mean.vec, each=data.per.segment)
n.data <- length(data.mean.vec)
n.validation.sets <- 200
prop.valid <- c(0.01, 0.05, 0.1, 0.25, 0.5)
if(interactive()){
  sim.result <- data.table(data.seed=1:100)[, {
    set.seed(data.seed)
    data.vec <- rnorm(n.data, data.mean.vec, 0.1)
    select.each.split <- CJ(split.i=1:n.validation.sets, prop.valid)[, {
      set.seed(split.i)
      prop.sets <- c(subtrain=1-prop.valid, validation=prop.valid)
      is.valid <- binsegRcpp::random_set_vec(
        n.data, prop.sets)=="validation"
      bs.model <- binsegRcpp::binseg_normal(
        data.vec, is.validation.vec=is.valid)
      bs.model$splits[, .(selected=segments[which.min(validation.loss)])]
    }, by=.(split.i, prop.valid)]
    data.table(n.splits=1:n.validation.sets)[, {
      select.each.split[split.i <= n.splits, .(
        times=.N
      ),
      by=.(prop.valid, selected)
      ][, .SD[which.max(times), .(selected)], by=prop.valid]
    }, by=n.splits]
  }, by=data.seed]

  if(require(ggplot2)){
    ggplot()+
      scale_color_gradient(low="red", high="black")+
      geom_line(aes(
        n.splits, selected,
        group=paste(data.seed, prop.valid),
        color=prop.valid),
        data=sim.result)+
      scale_y_continuous(breaks=seq(0, 100, by=10))
  }

  accuracy.dt <- sim.result[, .(
    correct=sum(selected==n.segments)
  ), by=.(prop.valid, n.splits)]
  if(require(ggplot2)){
    gg <- ggplot()+
      geom_line(aes(
        n.splits, correct,
        group=prop.valid,
        color=prop.valid),
        size=2,
        data=accuracy.dt)+
      scale_y_continuous("number of correctly chosen data sets")+
      scale_color_gradient(low="red", high="black")
    if(require(directlabels)){
      direct.label(gg, "right.polygons")
    }else{
      gg
    }
  }
}
```

The plot above suggests that 100 validation sets is sufficient.
