---
title: ggplot2 basics
layout: default
output: bookdown::html_chapter
---

# Chapter 2, data visualization using the grammar of graphics

This chapter explains the grammar of graphics, which is a powerful
model for describing a large class of data visualizations. After
reading this chapter, you will be able to

* State the advantages of the grammar of graphics relative to previous
  plotting systems
* Install the animint R package from GitHub
* Translate plot sketches into ggplot code in R
* Render ggplots on web pages using animint
* Specify manual scales for legends in ggplots
* Create multi-layer ggplots
* Create multi-panel ggplots

## History and purpose of the grammar of graphics

Most computer systems for data analysis provide functions for creating
plots to visualize patterns in data. The oldest systems provide very
general functions for drawing basic plot components such as lines and
points (e.g. the `graphics` and `grid` packages in R). If you use one
of these general systems, then it is your job to put the components
together to form a meaningful, interpretable plot. The advantage of
general systems is that they impose few limitations on what kinds of
plots can be created. The disadvantage is that general systems
typically do not provide functions for automating common plotting
tasks (axes, panels, legends).

To overcome the disadvantages of these general plotting systems,
charting packages such as `lattice` were developed (TODO: ref Sarkar
book). Such packages have several pre-defined chart types, and provide
a dedicated function for creating each chart type. For example,
`lattice` provides the `bwplot` function for making box and whisker
plots. The advantage of such systems is that they make it much easier
to create entire plots, including a legend and panels. The
disadvantage is the set of pre-defined chart types, which means that
it is not easy to create more complex graphics.

TODO: figure or table.

Newer plotting systems based on the grammar of graphics are situated
between these two extremes. Wilkinson proposed the grammar of graphics
in order to describe and create a large class of plots (TODO: cite
first edition of grammar of graphics book). Wickham later implemented
several ideas from the grammar of graphics in the `ggplot2` R package
(TODO: cite ggplot2 book). The `ggplot2` package has several
advantages with respect to previous plotting systems.

* Like general plotting systems, and unlike `lattice`, `ggplot2`
  imposes few limitations on the types of plots that can be created
  (there are no pre-defined chart types).
* Unlike general plotting systems, and like `lattice`, `ggplot2` makes
  it easy to include common plot elements such as axes, panels, and
  legends.
* Since `ggplot2` is based on the grammar of graphics, an explicit
  mapping of data variables to visual properties is required. Later in
  this chapter, we will explain how this mapping allows sketches of
  plot ideas to be directly translated into R code.

Finally, all of the previously discussed plotting systems are intended
for creating *static* graphics, which can be viewed equally well on a
computer screen or on paper. However, the main topic of this book is
`animint`, an R package for *interactive* graphics. In contrast to
static graphics, interactive graphics are best viewed on a computer
with a mouse and keyboard that can be used to interact with the plot.

Since many concepts from static graphics are also useful in
interactive graphics, the `animint` package is implemented as an
extension of `ggplot2`. In this chapter we will introduce the main
features of `ggplot2` which will also be useful for interactive plot
design in later chapters.

## Installing animint

The `animint` R package is hosted on GitHub, so you can install
`animint` by using the `devtools` package. To make sure that `animint`
is installed, open R and execute the following code.

```{r}
if(!require(animint)){
  install.packages("devtools")
  devtools::install_github("tdhock/animint", upgrade_dependencies=FALSE)
}
```

## Translating plot sketches into ggplots

This section explains how to translate a plot sketch into R `ggplot2`
code. We use a data set from the World Bank as an example, and we
begin by loading and looking at these data.

```{r}
data(WorldBank, package="animint")
tail(WorldBank)
dim(WorldBank)
```

The `WorldBank` data set consist of measures such as fertility rate
and life expectancy for each country over the period 1960-2010. The
code above prints the last few rows and the dimension of the data
table (11342 rows and 15 columns).

Suppose that we are interested to see if there is any relationship
between life expectancy and fertility rate. We could fix one year,
then use those two data variables in a scatterplot. Consider the
figure below which sketches the main components of that data
visualization.

TODO:SKETCH.

The sketch above shows life expectancy on the horizontal (x) axis,
fertility rate on the vertical (y) axis, and a legend for the
region. These elements of the sketch can be directly translated into R
code using the following method. First, we need to construct a data
table that has one row for every country in 1975, and columns named
`life.expectancy`, `fertility.rate`, and `region`. The `WorldBank`
data already has these columns, so all we need to do is consider the
subset for the year 1975:

```{r}
WorldBank1975 <- subset(WorldBank, year==1975)
head(WorldBank1975)
```

The code above prints the data for 1975, which clearly has the
appropriate columns, and one row for each country. The next step is to
use the notes in the sketch to code a ggplot with a corresponding
`aes` or aesthetic mapping of data variables to visual properties:

```{r}
library(ggplot2)
scatter <- ggplot()+
  geom_point(
    mapping=aes(x=life.expectancy, y=fertility.rate, color=region),
    data=WorldBank1975)
scatter
```

The `aes` function is called with names for visual properties (`x`,
`y`, `color`) and values for the corresponding data variables
(`life.expectancy`, `fertility.rate`, `region`). This mapping is
applied to the variables in the `WorldBank1975` data table, in order
to create the visual properties of the `geom_point`. The ggplot was
saved as the `scatter` object, which when printed on the R command
line shows the plot on a graphics device. Note that the `ggplot2`
system automatically shows a `region` color legend.

## Rendering ggplots on web pages using animint

This section explains how the `animint` package can be used to render
ggplots on web pages. The first step is to load the `animint` package:

```{r}
library(animint)
```

The ggplot from the previous section can be rendered with animint, by
first assigning it to a named element of a list. The R code below
assigns the ggplot `scatter` to a named element of the `scatter.viz`
list.

```{r}
scatter.viz <- list(scatter=scatter)
```

In this book we use `viz` (short for data visualization) to indicate
variables that are lists that contain ggplots. The `animint` package
implements a compiler that takes a `viz` list as input, and outputs a
web page with a data visualization. To activate the `animint`
compiler, we give that list the class `"animint"` in the R code below.

```{r}
structure(scatter.viz, class="animint")
```

When executing this code on the R command line, R will run the
`print.animint` function via the S3 object system
[http://adv-r.had.co.nz/OO-essentials.html#s3]. That will run the
`animint2dir` function, which compiles the animint `scatter.viz` list
to a directory of data and code files that can be rendered in a web
browser.

When viewed in a web browser, the plot should look mostly the same as
static versions produced by standard R graphics devices. One
difference is that the region legend is interactive: clicking a legend
entry will hide or show the points of that color.

## Multi-layer data visualization (multiple geoms)

Multi-layer data visualization is useful when you want to display
several different geoms or data sets in the same plot. For example,
consider the following R code which adds a `geom_path` to the previous
data visualization.

```{r}
WorldBankBefore1975 <- subset(WorldBank, 1970 <= year & year <= 1975)
two.layers <- scatter+
  geom_path(aes(x=life.expectancy, y=fertility.rate, color=region,
                group=country),
            data=WorldBankBefore1975)
two.layer.viz <- list(scatter=two.layers)
structure(two.layer.viz, class="animint")
```

This plot shows a data visualization with 2 geoms/layers: 

* the `geom_point` shows the life expectancy, fertility rate, and
  region of all countries in 1975.
* the `geom_path` shows the same variables for the previous 5 years.

The addition of the `geom_path` shows how the countries changed over
time. In particular, it shows that most countries moved to the right
and down, meaning higher life expectancy and lower fertility
rate. However, there are some exceptions. For example, the two East
Asian countries in the bottom left suffered a decrease in life
expectancy over this period. And there are some countries which showed
an increased fertility rate. Can we add the names of these countries
to this data viz? Below, we add another layer with a text label for
each country's name.

```{r}
three.layers <- two.layers+
  geom_text(aes(x=life.expectancy, y=fertility.rate, color=region,
                label=country),
            data=WorldBank1975)
three.layer.viz <- list(scatter=three.layers)
structure(three.layer.viz, class="animint")
```

This data viz is not so easy to read, since there are so many
overlapping text labels. The interactive region legend helps a little,
by allowing the user to hide data from selected regions. However, it
would be even better if the user could show and hide the text for
individual countries. That type of interaction can be achieved using
the showSelected and clickSelects aesthetics which we explain in
Chapters 3-4.

For now, we move on to discuss the other main advantage of ggplot2:
facets.

## Multi-panel data visualization (facets)

Panels or facets are sub-plots that show related data
visualizations. One of the main strengths of ggplot2 is that different
kinds of multi-panel plots are relatively easy to create. Multi-panel
data visualization is used in two different ways:

* You want to align the axes of several related plots containing
  different geoms. This facilitates comparison between several
  different geoms, and is a technique that is also useful for
  interactive data visualization.
* You want to divide the data from one geom into several panels. This
  facilitates comparison between data subsets, and is less useful for
  interactive data visualization (interactivity can often be used
  instead, to achieve the same effect of comparing data subsets).

  
### Different geoms in each panel (aligned axes)
  
Facets are useful to align the axes of related plots. For example,
consider the following data visualization which consists of two plots:
a scatterplot with data from 1975, and a time series with data from
1960-2010.
  
```{r}
viz.two.plots <- two.layer.viz
viz.two.plots$timeSeries <- ggplot()+
  geom_line(aes(x=year, y=fertility.rate, color=region, group=country),
            data=WorldBank)
structure(viz.two.plots, class="animint")
```

The data visualization above contains two ggplots. Both ggplots map
the fertility rate variable to the y axis. However, since they are
separate plots, the ranges of their y axes are computed
separately. That means that the two y axis are not exactly
aligned. That is a problem since it would make it easier to decode the
data visualization if each unit of vertical space was used to show the
same amount of fertility rate. To achieve that effect, we use facets
in the data visualization below.

```{r}
addColumn <- function(df, time.period){
  data.frame(df, time.period=factor(time.period, c("1975", "1960-2010")))
}
viz.aligned <- list(
  scatter=ggplot()+
    geom_point(aes(x=life.expectancy, y=fertility.rate, color=region),
               data=addColumn(WorldBank1975, "1975"))+
    geom_path(aes(x=life.expectancy, y=fertility.rate, color=region,
                  group=country),
              data=addColumn(WorldBankBefore1975, "1975"))+
    geom_line(aes(x=year, y=fertility.rate, color=region, group=country),
              data=addColumn(WorldBank, "1960-2010"))+
    facet_grid(. ~ time.period, scales="free")
  )
structure(viz.aligned, class="animint")
```

The data visualization above contains a single ggplot with two panels
and three layers. The left panel shows the `geom_point` and
`geom_path`, and the right panel shows the `geom_line`. The panels
have a shared axis for fertility rate, which ensures that the time
series can be directly compared with the scatterplot.

Note that we used the `addColumn` function to add a `time.period`
variable to each data set, and then we used that variable in
`facet_grid(scales="free")`. We call this the "addColumn-then-facet"
idiom, which is useful in general for creating a multi-panel data
visualization with aligned axes (TODO:ref appendix). In particular, if
we wanted to change the order of the panels in the data visualization,
we would only need to edit the order of the factor levels in the
definition of `addColumn`.

### Same geoms in each panel (compare data subsets)

The second reason for using plots with multiple panels in a data
visualization is to compare subsets of observations. This facilitates
comparison between data subsets, and can be used in at least two
different situations

* One geom's data set has too many observations to display
  informatively in one panel.
* You want to compare different subsets of data that is plotted for
  one geom.
  
For example, the code below creates two data sets based on three years
of the WorldBank data set. 

```{r}
show.point.list <- list()
show.path.list <- list()
for(show.year in c(1975, 1985, 1995)){
  show.point.list[[paste(show.year)]] <- data.frame(
    show.year, subset(WorldBank, year==show.year))
  five.before <- show.year - 5
  show.path.list[[paste(show.year)]] <- data.frame(
    show.year, subset(WorldBank, five.before <= year & year <= show.year))
}
show.point <- do.call(rbind, show.point.list)
show.path <- do.call(rbind, show.path.list)
```

We used a for loop over three values of `show.year`, the variable
which we will use later in `facet_grid`. For each value of
`show.year`, we store a data subset as a named element of a
list. After the for loop, we use `do.call` with `rbind` to combine the
data subsets. This is an example of the "list-of-data-tables" idiom,
which is also useful for interactive data visualization (TODO: ref
appendix).

Below, we facet on the `show.year` variable to create a data
visualization with three panels.

```{r}
viz.panels <- list(
  scatter=ggplot()+
    geom_point(aes(x=life.expectancy, y=fertility.rate, color=region),
               data=show.point)+
    geom_path(aes(x=life.expectancy, y=fertility.rate, color=region,
                  group=country),
              data=show.path)+
    facet_grid(. ~ show.year)
  )
structure(viz.panels, class="animint")
```

The data visualization above contains a single ggplot with three
panels. It shows more of the WorldBank data set than the previous
single-panel visualization with data only from 1975. However, it still
only shows a small subset of these data (from 1975, 1985, and
1995). 

You may be tempted to try using a panel to display every year (not
just 1975, 1985, and 1995). However, beware that this type of
multi-panel data visualization is especially useful if there are only
a few data subsets. With more than about 10 panels, it becomes
difficult to see all the data at once, and thus difficult to make
meaningful comparisons.

In the next chapter, we will show how the new `showSelected` keyword
can be used to achieve animation, and show more of this data set.

## Chapter summary and exercises

This chapter presented the basics of static data visualization using
ggplot2. We showed how animint can be used to render a list of ggplots
in a web browser. We explained two features of ggplot2 that make it
ideal for data visualization: multi-layer and multi-panel graphics.

Exercises:

* What are the three main advantages of `ggplot2` relative to previous
  plotting systems such as `grid` and `lattice`?
* What is the purpose of multi-layer graphics?
* What are the two different reasons for creating multi-panel
  graphics? Which of these two types is useful with interactivity?
* Let us define "A < B" to mean that "one B can contain several A."
  Which of the following statements is true?

	* ggplot < panel
	* panel < ggplot
	* ggplot < animint
	* animint < ggplot
	* layer < panel
	* panel < layer
	* layer < ggplot
	* ggplot < layer
	
* In the `viz.aligned` facets, why is it important to use the
  `scales="free"` argument?
* In `viz.aligned` we showed a ggplot with a scatterplot panel on the
  left and a time series panel on the right. Make another version of
  the data visualization with the time series panel on the left and
  the scatterplot panel on the right.
* In `viz.aligned` the scatterplot displays fertility rate and life
  expectancy, but the time series displays only fertility rate. Make
  another version of the data visualization that shows both time
  series. Hint: use both horizontal and vertical panels in
  `facet_grid`.
* Use `aes(size=population)` in the scatterplot to show the population
  of each country. Hint: `scale_size_animint(pixel.range=c(5, 10)`
  means that circles with a radius of 5/10 pixels should be used
  represent the minimum/maximum population.
* Create a multi-panel data visualization that shows each year of the
  `WorldBank` data set in a separate panel. What are the limitations
  of using static graphics to visualize these data?
* Create `viz.aligned` using a plotting system that is not based on
  the grammar of graphics. For example, you can use functions from the
  `graphics` package in R (`plot`, `points`, `lines`, etc), or
  matplotlib in Python. What are some advantages of ggplot2 and
  animint?
