# Chapter 4, clickSelects

This chapter explains clickSelects, one of the two main keywords that
animint introduces for interactive data visualization. The
clickSelects keyword specifies a geom for which clicking updates a
selection variable. Each geom in a data visualization has its own data
set, and its own definition of the clickSelects keyword. So clicking
on different geoms can change different selection variables.

After reading this chapter, you will be able to

* Understand how interactive legends implicitly use clickSelects.
* Use the clickSelects keyword in your plot sketches.
* Translate your plot sketches with clickSelects into R code.
* Use the `selector.types` option to specify multiple selection
  variables.
  
## Interactive legends implicitly use clickSelects

In this section, we will explain how the clickSelects keyword is
implicitly used in interactive legends. If you have read the previous
chapters, you have already implicitly used clickSelects, which was
automatically created for the interactive legends in the previous
chapters. For example, consider the sketch of the World Bank data viz
from the last chapter.

TODO: sketch.

Since the legend has `clickSelects=region`, clicking an entry of that
legend updates the `region` selection variable. Note that animint
automatically makes every discrete legend interactive, so you do not
need to explicitly specify `clickSelects=region` for the legend. In
fact, when we specified `color=region` for the `geom_point`, animint
does two things automatically:

* `showSelectedLegendColor=region` is assigned to the same
  `geom_point`.
* `clickSelects=region` is assigned to the color legend.

Note that `clickSelects` aesthetics are not limited to interactive
legends. Each geom has its own `clickSelects` variable, which
determines which selection variable is updated after clicking that
geom. In the next section we will give several examples of how
clickSelects can be used in combination with showSelected to create
interactive data visualizations.

## Use the clickSelects aesthetic to identify points on a scatterplot

The goal of this section is to create the following visualization of
the World Bank data.

TODO: sketch of scatterplot with one text label.

To start, consider the following R code which generates a
scatterplot of the World Bank data:

```{r}
data(WorldBank, package="animint")
library(animint)
scatter <- ggplot()+
    geom_point(aes(x=life.expectancy, y=fertility.rate, color=region,
                   key=country,
                   showSelected=year,
                   clickSelects=country),
               data=WorldBank)
scatter
```

First, note that we use the `addColumn` function to add a factor
variable to the `WorldBank` data set, which we will use later as a
facet variable. Also note that ggplot2 ignores the animint aesthetics
(key, showSelected, and clickSelects), so the plot above is not
interactive. In contrast, rendering the same ggplot using animint
results in the following interactive plot:

```{r}
viz.scatter <- list(
  scatter=scatter,
  duration=list(year=2000))
structure(viz.scatter, class="animint")
```

Try clicking data points in the scatterplot above. You should see the
value of the `country` selection menu change after clicking a data
point. You should also see that the data point for the selected
country is darker than the others. This serves to highlight the
current selection, and is performed automatically for each geom with
`clickSelects`. By default the selected point has alpha=1 (fully
opaque, no transparency), and the other points have alpha=0.5
(semi-transparent).

The data visualization above shows the currently selected country in
the selection menu, but it would be better to show it as a text label
on the scatterplot. We can do that by adding a `geom_text` layer with
several showSelected aesthetics:

```{r}
viz.text <- viz.scatter
viz.text$scatter <- scatter+
  geom_text(aes(x=life.expectancy, y=fertility.rate, label=country,
                showSelected=year,
                showSelected2=country,
                showSelected3=region),
            data=WorldBank)
structure(viz.text, class="animint")
```

After clicking a data point in the scatterplot above, you should see a
text label with the country name appear. Furthermore, try changing the
year using the selection menu. You should see the text label move in a
smooth transition along with the corresponding data point. Try
clicking the legend entry that corresponds to the region for the
currently selected country (e.g. if Canada is selected, try clicking
the North America legend entry). You should see the text label
disappear along with all of the points for that region.

As explained in the last chapter, any aesthetic that contains the
substring `showSelected` is treated as an interactive variable. In the
example above, we specified three `showSelected` variables for the
`geom_text`. This means to only draw a text label for the rows of the
`WorldBank` data set that match the current values of all three of
those selection variables. Since each combination of `country` and
`year` has one row in these data, only one text label will be shown at
a time.

Finally, note that specifying `showSelected3=region` causes the text
label to be hidden when changing the selected set of regions using the
interactive legend. Since we specified `color=region` for the
`geom_point`, it automatically gets a `showSelectedLegendColor=region`
aesthetic, which has the effect of hiding points after clicking the
region legend. However, since the `geom_text` does not have
`color=region`, we have to manually specify `showSelected3=region` if
we want to hide it after clicking the region legend.

In the last chapter, we introduced the terms "direct manipulation" and
"indirect manipulation" to describe interactions with legends and
menus. In the data viz above, we can change the value of the `country`
selection variable by either clicking a data point (direct
manipulation) or using the selection menu (indirect
manipulation). Both techniques are useful, but for different
purposes:

* Direct manipulation by clicking data points is useful to find the
  names of countries with extreme values of fertility rate and life
  expectancy. For example, for the year 1960, clicking the point at
  the bottom left of the plot reveals the country name Gabon.
* Indirect manipulation using menus is useful to see the plotted
  position of a country of interest. For example, it would be
  difficult to find France by clicking all the different points, but
  it is simple to find France by typing its name in the selection
  menu.
  
Note that in the data viz above, there is only one country selected at
a time. In the next section, we will explain how the `selector.types`
option can be used to change `country` to a multiple selection
variable.

## The selector.types and first options

In this section our goal is to produce a slightly more complicated
version of the scatterplot in the last section. The sketch below has
only one difference with respect to the sketch from the last section:
text labels are shown for more than one country.

TODO: sketch with 2 or 3 countries selected.

In animint, each selection variable has a type, either single or
multiple. Single selection means that only one value can be selected
at a time. Multiple selection means that any number of values can be
selected at a time. In the plots in the last section, multiple
selection was used for the `region` variable but not for the `year`
and `country` variables. Why is that?

By default, animint assigns multiple selection to all variables that
appear in interactive discrete legends, and single selection to other
variables. However, single or multiple selection can be specified by
using the `selector.types` option. In the R code below, we use the
`selector.types` option to specify that `country` should be treated as
a multiple selection variable.

```{r}
viz.multiple <- viz.text
viz.multiple$selector.types <- list(country="multiple")
structure(viz.multiple, class="animint")
```

The first thing you probably noticed in the data viz above is that
there is now a text label for each country. For each multiple
selection variable, animint selects all possible values at first, by
default. To change the set of values which is selected at first, we
can specify the `first` option as in the R code below.

```{r}
viz.first <- viz.multiple
viz.first$first <- list(
  year=1970,
  country=c("Canada", "India", "Pakistan", "Japan"),
  region=c("North America", "South Asia"))
structure(viz.first, class="animint")
```

When the data viz above is first rendered, it shows data points from
the year 1970, for each country in North America and South Asia. It
also shows text labels for Canada, India, and Pakistan. It does not
render a text label or a data point for Japan, because the East Asia
region is not selected at first. Try selecting the East Asia region,
and you should see the text label for Japan appear, along with all of
the other data points for that region.

You may have noticed that it is easy to add countries to the current
selection, by clicking data points. Normally, clicking a selected data
point will remove that country from the current selection. However, in
this particular data viz, it is not so easy to remove them, since the
text labels are rendered on top of the data points. To fix that
problem, we specify `clickSelects=country` for the `geom_text` in the
R code below.

```{r}
viz.text.click <- viz.first
viz.text.click$scatter <- ggplot()+
    geom_point(aes(x=life.expectancy, y=fertility.rate, color=region,
                   key=country,
                   showSelected=year,
                   clickSelects=country),
               data=WorldBank)+
    geom_text(aes(x=life.expectancy, y=fertility.rate, label=country,
                  showSelected=year,
                  showSelected2=country,
                  showSelected3=region,
                  ## The next two lines are the only additions:
                  key=country,
                  clickSelects=country),
              data=WorldBank)
structure(viz.text.click, class="animint")
```

Since we used `clickSelects=country` for the `geom_text` in the data
viz above, clicking a text label de-selects that country, making that
text label disappear. We also added `key=country` to make sure that
the text labels always use the correct country for their smooth
transitions between years.

In general the `key` aesthetic should be specified for all geoms that
use `showSelected` with a variable that appears in the `duration`
option. In this example, we used the `duration` option to specify a
smooth transition for the `year` variable. Since we use
`showSelected=year` in both the `geom_point` and the `geom_text`, we
also specified the `key` aesthetic for these geoms.

Note that in the data viz above, the year variable can only be changed
via the selection menu. In the next section, we will add a facet with
a geom that can be directly clicked to change the year variable.

## Selecting a year on a time series facet

The goal of this section is to add a facet with a time series plot
that can be clicked to change the selected year.

TODO: sketch.

Note that the sketch above includes `geom_tallrect`, a new geom
introduced in animint. It is "tall" because it occupies the entire
vertical space of the plot, and thus only requires definition of its
horizontal limits via the `xmin` and `xmax` aesthetics. Specifying
`clickSelects=year` means that we want to be able to draw one tallrect
for each year, and click a tallrect to change the selected year. Thus
we need to create a new data set called `years` with one row for each
unique year of the `WorldBank` data.

```{r}
years <- data.frame(year=unique(WorldBank$year))
head(years)
```

Next, we re-create the scatterplot from the previous section using the
"addColumn then facet" idiom, which is useful for creating ggplots
with aligned axes (TODO: ref appendix).

```{r}
add.time.period <- function(df, time.period){
  data.frame(
    df, time.period=factor(time.period, c("selected year", "all years")))
}
scatterFacet <- ggplot()+
  geom_point(aes(x=life.expectancy, y=fertility.rate, color=region,
                 key=country,
                 showSelected=year,
                 clickSelects=country),
             data=add.time.period(WorldBank, "selected year"))+
  geom_text(aes(x=life.expectancy, y=fertility.rate, label=country,
                key=country,
                clickSelects=country,
                showSelected=year,
                showSelected2=country,
                showSelected3=region),
            data=add.time.period(WorldBank, "selected year"))+
  facet_grid(. ~ time.period, scales="free")+
  theme_bw()+
  theme(panel.margin=grid::unit(0, "lines"))
scatterFacet
```

Note that the ggplot above uses the same `aes` definitions as the
scatterplot from the previous section. The only difference is that we
have used an augmented `WorldBank` data set with an additional
`time.period` variable that we use with `facet_grid`. Below, we add
geoms for a time series plot that is aligned on the fertility rate
axis.

```{r}
scatterTS <- scatterFacet+
  geom_tallrect(aes(xmin=year-0.5, xmax=year+0.5,
                    clickSelects=year),
                alpha=0.5,
                data=add.time.period(years, "all years"))+
  geom_line(aes(x=year, y=fertility.rate, group=country, color=region,
                clickSelects=country),
            size=3,
            alpha=0.6,
            data=add.time.period(WorldBank, "all years"))
scatterTS
```

The two geoms defined above occupy a new facet for the `"all years"`
value of the `time.period` variable (defined by the `add.time.period`
function). Since these two geoms have different definitions of
`clickSelects`, clicking each geom will update the plot in a different
way. Note that for the `geom_line` we specify `size=3`, which means a
line stroke width of 3 pixels. In general it is a good idea to
increase the size of geoms with `clickSelects`, to make them easier to
click.

Also note that we specified `alpha=0.5` for the `geom_tallrect` and
`alpha=0.6` for the `geom_line`. Since both of these geoms define
`clickSelects`, some plotted lines and tallrect will be selected, and
others will not be selected. The alpha values in R code specify the
opacity of the selected objects, and other objects will have an
`alpha` opacity which is 0.5 less than that value. In the example
above, the un-selected lines will have `alpha=0.1`un-selected
tallrects will have `alpha=0` (completely transparent), and the
un-selected tallrects will have `alpha=0` (completely transparent).

Finally, we use the R code below to render the new aligned scatterplot
and time series using animint.

```{r}
viz.facets <- viz.text.click
viz.facets$scatter <- scatterTS
structure(viz.facets, class="animint")
```

The interactive data viz above contains a new panel with lines that
show a fertility rate time series over all years. Since we specified
`clickSelects=country` for the `geom_line`, clicking a line updates
the set of selected countries. Since we specified `clickSelects=year`
for the `geom_tallrect`, clicking on a tallrect updates the selected
year. 

## Chapter summary and exercises

This chapter explained `clickSelects`, one of the two main keywords
that animint introduces for interactive data visualization design. We
used the World Bank data set to show how `clickSelects` can be used to
specify different interactions for each of the plotted geoms. We
explained how the `selector.types` option can be used to specify
multiple selection variables. We also explained how the `first` option
can be used to specify the selected values that are used when the
animint is first rendered.

Exercises:

* So far we have seen three different ways to change selection
  variables: (1) interactive legends, (2) selection menus, and (3)
  clicking data with `clickSelects`. Order these three techniques in
  terms from most to least direct manipulation. Which technique is
  preferable in what circumstances?
* When `geom_point(aes(clickSelects=something), alpha=0.75)` is
  rendered with ggplot2, how much opacity/transparency is present for
  all data points? When animint renders the same geom, some points
  will be selected and others not. What is the opacity/transparency of
  selected points? What is the opacity/transparency of points which
  are not selected?
* Add `aes(size=population)` to the points in the World Bank
  scatterplot. Is the size legend interactive? Why?
* Add a `geom_text` to the World Bank scatterplot that shows the
  selected year.
* Add a `geom_text` to the World Bank time series to show the names of
  the selected countries.
* Add a `geom_path` to the World Bank scatterplot to show data for the
  last 5 years.
* Use the `time` option to make an animated version of `viz.facets`.
