# Chapter 3, the showSelected keyword

This chapter explains showSelected, one of the two main keywords that
animint introduces for interactive data visualization. After reading
this chapter, you will be able to

* Use the showSelected keyword in your plot sketches to specify geoms
  for which only a subset of data should be plotted at any time.
* Use selection menus in animint to change the subset of plotted data.
* Specify smooth transitions between data subsets using the duration
  option and key aesthetic.
* Create animated data visualizations using the time option.

## Sketching with showSelected

In this section, we will explain how the showSelected keyword can be
used in plot sketches. The showSelected keyword specifies a variable
to use for subsetting the data before plotting.

In fact, we have already seen an example of the showSelected keyword,
in the interactive legends that we created in the previous two
chapters. For example, consider the sketch below of the Keeling Curve
data visualization from Chapter 1.

TODO: SKETCH co2.

The sketch above includes `showSelected=month` for the `geom_point`,
meaning that it should show the subset of data for the selected
months. In contrast, since the `geom_line` does not include
`showSelected` keywords, it always shows the entire data set
(regardless of the selected months).

As another example, consider the sketch below of the `basic` WorldBank
data visualization from Chapter 2.

TODO: SKETCH WorldBank.

The sketch above specifies `showSelected=region` for the `geom_point`,
meaning that it should show the subset of data for the selected
regions. Note that the code we used in chapter 2 did not explicitly
specify `showSelected=region`. Instead, we specified
`aes(color=region)`, and animint automatically assigned
`aes(showSelectedLegendColor=region)`. In general, animint will assign
a showSelected aesthetic for each variable that is used in a
categorical legend.

Each variable that is used with a showSelected aesthetic is treated by
animint as a *selection variable*. For example, the Keeling Curve data
viz has one selection variable (month), and so does the WorldBank data
viz (region). For each selection variable, animint keeps track of the
currently selected values. When the selection changes, animint updates
the subset of data that is shown.

Each of the data visualizations sketched above has only one selection
variable. However, a data visualization can have any number of
selection variables. In the next section, we will explore a
visualization of the World Bank data that has selection variables for
`region` and `year`.

## Selecting data subsets using menus 

The R code below produces a scatterplot of life expectancy and
fertility rate for each country in the WorldBank data set. 

```{r}
data(WorldBank, package="animint")
library(ggplot2)
basic <- ggplot()+
  geom_point(aes(x=life.expectancy, y=fertility.rate, color=region,
                 showSelected=year),
             data=WorldBank)
basic
```

Note that the ggplot above contains the `showSelected` aesthetic, one
of the two main features introduced in animint. The `showSelected`
aesthetic is ignored by ggplot2, which produced a scatterplot with one
point for every country and year.

In constrast, rendering the same ggplot using animint yields the
interactive data visualization below.

```{r}
library(animint)
viz.basic <- list(scatter=basic)
structure(viz.basic, class="animint")
```

Note that the data viz above has two selection variables: region and
year. Each variable has a menu that can be used to change the current
selection.

Discrete legend variables such as region default to multiple
selection, so several values are selected and shown at once. Try
changing the selected region in the interactive legend and the
selection menu. When you change the selection using either method,
both the interactive legend and the selection menu should update to
reflect the current selection. 

Other selection variables such as year default to single selection, so
only one value is selected and shown at any time. Try changing the
selected value of the year variable using the selection menu. You
should see the points in the scatterplot immediately update to show
the fertility rate and life expectancy of all the countries in the
year that you selected.

## Transitions: the duration option and key aesthetic

You may have noticed that there are three buttons at the bottom of
each data visualization created by animint. Try clicking the "Show
animation controls" button above. This table contains a row for each
selection variable. The text boxes show the number of milliseconds
that are used for transition durations after updating each selection
variable. The default transition duration for each selection variable
is 0, meaning data will be immediately placed at their new positions
after updating each variable.

To illustrate the significance of transition durations, try changing
the transition duration of the year variable to 2000. Then, change the
selected value of the year variable. You should see the data points
move slowly to their new positions, over a duration of 2 seconds.

Some transitions result in points moving only a little bit, to nearby
positions (e.g. 1979-1980). Other transitions result in points moving
a lot more, to far away locations (e.g. 1980-1981). Why is that?

Smooth transitions only make sense for data points that exist both
before and after changing the selection. In the R code below we
compute a table of counts of data points that can be plotted in each
of these three years.

```{r}
three.years <- subset(WorldBank, 1979 <= year & year <= 1981)
can.plot <- with(three.years, {
  (!is.na(life.expectancy)) & (!is.na(fertility.rate))
})
table(three.years$year, can.plot)
```

It is clear from the table above that there are 187 points that can be
plotted in 1979 and 1980. However, in 1981 there is one more data
point, corresponding to a country for which we did not have data in
1980. Below we show the data for that country, Kosovo.

```{r}
subset(three.years, country=="Kosovo")
```

Indeed, the table above shows that fertility rate and life expectancy
are missing for Kosovo during 1979-1980. Thus it does not make sense
to do a smooth transition for countries such as Kosovo which would not
be plotted either before or after the transition. How to specify that
in the data visualization? In the code below, we use
`aes(key=country)` to specify that the `country` variable should be
used to match data points before and after changing the selection.

```{r}
scatter.key <- ggplot()+
  geom_point(aes(x=life.expectancy, y=fertility.rate, color=region,
                 key=country,
                 showSelected=year),
             data=WorldBank)
```

Like `showSelected`, the `key` aesthetic is only meaningful for
interactive data visualization, so ggplot2 will ignore it. However, if
we render this ggplot using animint, the `country` variable will be
used to make sure transtion durations are meaningful. To specify a
default transition duration for the `year` variable, we use the
`duration` option in the data viz below.

```{r}
viz.duration <- list(
  scatter=scatter.key,
  duration=list(year=2000))
structure(viz.duration, class="animint")
```

The `duration` option must be a named list. Each name should be a
selection variable, and each value should specify the number of
milliseconds to use for a transition duration when the selected value
of that variable is changed.

If you click "Show animation controls" in the data viz above, you will
see that the text box for the year variable is 2000, as specified in
the R code. If you change the selection from 1980 to 1981, you should
see a proper transition.

## Animation: the time option

The `time` option is used to specify a variable to use for animation.

```{r}
viz.duration.time <- viz.duration
viz.duration.time$time <- list(variable="year", ms=2000)
structure(viz.duration.time, class="animint")
```

## Chapter summary and exercises

This chapter explained the showSelected aesthetic, selection menus,
transition durations, and animation.

Exercises:

* Make an improved version of `viz.aligned` from the previous
  chapter. Instead of fixing the year at 1975, use
  `aes(showSelected=year)` so that the user can select a year. Add
  geoms that show the selected year: a `geom_text` on the scatterplot,
  and a `geom_vline` on the time series.
* Translate an example from the animation package. Hint: in the code
  for the animation package there is always a for loop over the time
  variable. Instead of calling a plotting function inside the for
  loop, use the list-of-data-tables idiom to store the data that
  should be plotted (TODO:ref appendix). Then use those data along
  with `showSelected` aesthetics to create ggplots, and render them
  using animint.
