---
title: Motivation
layout: default
output: bookdown::html_chapter
---

<!-- comment -->

# Chapter 1, Motivation

<!-- comment -->

 
 
 
 
 
 ```{r setup, echo=FALSE}
 ```{r, echo=FALSE}
 ```{r, echo=FALSE}
 ```{r, echo=FALSE}
 ```{r, echo=FALSE}
 ```{r Ch01-viz, echo=FALSE}
 knitr::opts_chunk$set(fig.path="Ch01-figures/")
 TeaTasting <- matrix(c(
    3, 1, 1, 3),
    nrow = 2,
    dimnames = list(
        Guess = c("Milk", "Tea"),
        Truth = c("Milk", "Tea")))
print(TeaTasting)
 year.int <- as.integer(floor(time(co2)))
month.int <- as.integer(time(co2) %% 1 * 12 + 1)
year.month.chr <- sprintf("%d-%02d-15", year.int, month.int)
year.month.POSIXct <- as.POSIXct(strptime(year.month.chr, "%Y-%m-%d"))
Sys.setlocale("LC_TIME", "C")
month.str <- strftime(year.month.POSIXct, "%B")
month <- factor(month.str, month.str[1:12])
library(data.table)
co2.details <- data.table(
  year.int,
  month.int,
  month,
  year.month.POSIXct,
  ppm=as.numeric(co2))
co2.details
 library(animint2)
extremes <- co2.details[month.int %in% c(5, 10),]
ggplot()+
  geom_line(aes(year.month.POSIXct, ppm),
            data=co2.details)+
  geom_point(aes(year.month.POSIXct, ppm, color=month),
             data=extremes)
 TimeSeries <- ggplot()+
  geom_line(aes(year.month.POSIXct, ppm),
            size=1,
            data=co2.details)+
  geom_point(aes(year.month.POSIXct, ppm, color=month),
             data=co2.details)+
  xlab("Time of measurement")+
  ylab("Atmospheric carbon dioxide (parts per million)")
TimeSeries
 animint(
  TimeSeries,
  first=list(month=c("May", "October")))
 

<!-- comment -->

The purpose of this manual is to explain the usage of animint2, an R
package for interactive data visualization. This introductory chapter
answers the following questions:

<!-- comment -->

* What are data, and how are they analyzed?
* What is data visualization, and when is it useful for data analysis?
* What is interactive data visualization, and when is it useful?

<!-- comment -->

This introductory chapter uses the following outline:

<!-- comment -->

* What is data?
* Small data -- data visualization is not necessary.
* Medium data -- static data visualization is sufficient.
* Large data -- interactive data visualization is useful.

<!-- comment -->

## What is data? {#data}

<!-- comment -->

Data are any pieces of information that are systematically recorded,
either on paper or on a computer. Anybody can create data, just by
systematically writing things down. Typically, data are created in
order to help answer a specific question, and are organized into
tables with rows for observations and columns for variables or
different types of information. The word "data" is the plural form of
"datum," which we use to refer to one observation/row of a data
table. We use the term "data set" to refer to a subset of
observations/rows, or the entire data table.

<!-- comment -->

There are many examples of data that could be created to answer
questions based on everyday experiences:

<!-- comment -->

* How does the weather this year compare to previous years? Have we
  had more or less rain than usual? To answer these questions, we
  could create a data table with column for measurements of different
  weather conditions: temperature, rainfall, etc. There should also be
  columns for the date and time of each observation, and a row for
  each observation.
* How is this new diet affecting me? If you are trying a new diet, you
  may want to record what you eat and how you feel after each meal. In
  that case you could make a table with a row for each meal and four
  columns: date, time, what you ate, and how you felt after.
* Does this new lung cancer treatment work better than the old
  treatment? A doctor who conducts the clinical trial would randomly
  assign patients to receive either the new or old treatment. The
  doctor would then create a data table with a row for each patient,
  and several columns: years the patient has smoked, treatment type
  (new or old), patient age at treatment, patient age at death.

<!-- comment -->

We define "data analysis" as the process of answering these questions
by converting the raw data table into other, more comprehensible
forms. One highly effective class of methods for data analysis is
called "data visualization," which seeks to provide answers to
questions by converting a data set into an informative picture. The
term "data visualization" refers to both the picture itself (also
known as a plot, chart, figure, graph, graphic, or data viz), and the
process of creating the picture.

<!-- comment -->

There are many different ways to perform data analysis, and data sets
of different sizes should be analyzed using different
techniques. There are many different ways to characterize the size of
data sets, and every author uses a slightly different definition. In
this manual we will use a classification of data sets into three sizes:
small, medium, and large. We begin by discussing small data sets, for
which data visualization is not necessary.

<!-- comment -->

## Small data analysis without visualization {#small-data}

<!-- comment -->

In this section, we will discuss "small data," which are small enough
such that data analysis can be done by simply looking at the entire
data table. For small data sets, there is no need to use data
visualization. Instead, the data can simply be presented for visual
inspection in a table.

<!-- comment -->

As a concrete example, consider
[the famous tea tasting experiment proposed by Ronald Fisher](https://en.wikipedia.org/wiki/Lady_tasting_tea). A
Lady claimed that she could taste the difference when milk is added to
the teacup before or after the tea. Fisher asked the question, can the
Lady really taste the difference between the two types of tea?

<!-- comment -->

To answer that question, Fisher prepared four cups of tea with milk
added after, and four cups of tea with milk added before. Fisher then
placed the cups in a random order, had the Lady taste all eight cups
of tea, and asked her to identify the four in which milk was added
after the tea. According to `help(fisher.test)` in R, the Lady
correctly identified three of the four cups in which milk was added
after the tea. Fisher than wrote down the following data: the total
number of cups (8), the total number of cups with milk added (4), and
the total number of cups that the Lady correctly identified (3). In R,
this data set can be viewed by printing a contingency table of count
data:
 

<!-- comment -->

In this case, the data set is small enough such that Fisher's question
can be answered by simply looking at the data table itself. If the
Lady had been able to correctly identify all four cups, then that
would have been a very convincing demonstration of her
ability. However, she was apparently only able to correctly identify
three out of the four cups, which is less convincing. 

<!-- comment -->

The main topic of this manual is data visualization, which is not
necessary for such small data sets. Instead, we will focus on data
sets that are too big to be analyzed by manual visual inspection of
the data table.

<!-- comment -->

## Medium data analysis with static data visualizations {#medium-data}

<!-- comment -->

For medium sized data sets, simply inspecting the data table is no
longer sufficient to answer the questions posed during data
analysis. Medium data are big enough such that we need to use
visualization to understand the data. 

<!-- comment -->

For example, consider the following data on atmospheric carbon dioxide
(CO2) concentrations, recorded monthly between 1959 and 1997.
 

<!-- comment -->

Printing these data on the R command line shows that there are 468
rows/observations total. This is not a huge number of observations,
but it is already big enough so that answering questions is not easy
by simple visual inspection of the data table. Instead, we will create
a static data visualization:
 

<!-- comment -->

The static data visualization shows that CO2 concentrations increased
over the second half of the twentieth century. This particular data
visualization is called a
[Keeling Curve](https://en.wikipedia.org/wiki/Keeling_Curve). It is
named after Charles David Keeling, the pioneering scientist who
collected the first frequent regular data on atmospheric CO2. The
general increasing trend can be explained by considering the chemical
process of combustion, which converts oxygen to CO2. Keeling noted
that "the observed rate of increase is nearly that to be expected from
the combustion of fossil fuel" (REF: The Concentration and Isotopic
Abundances of Carbon Dioxide in the Atmosphere, Keeling 1960).

<!-- comment -->

The data visualization also reveals the interesting seasonal trend
that attains a local maximum each May and a local minimum each
October. This seasonal trend can be explained by considering the
forests in the Northern Hemisphere. The leaves on the trees in these
forests perform photosynthesis, the chemical conversion of CO2 to
oxygen. During the winter months there are no leaves on the trees, so
CO2 accumulates in the atmosphere until it peaks in May of each
year. When the leaves come back each year, they perform photosynthesis
throughout Spring and Summer, which causes the atmospheric CO2
concentration to drop until it reaches its yearly minimum in October.

<!-- comment -->

We say that this data visualization is "static" or "non-interactive"
because the reader can view it but can not change what is
displayed. That is fine for medium sized data sets, in which we can
see all the details of the data set.  However, as we discuss in the
following section, static data visualization is not sufficient to show
all the details in larger data sets.

<!-- comment -->

## Large data analysis with interactive data visualization {#large-data}

<!-- comment -->

Some data sets are so large that it is not possible or desirable to
plot all of the data at once in a static data visualization. For such
"large data" sets, traditional approaches to data analysis include
summarizing the data, and then visualizing the summary. However, the
summary can be misleading, because it does not show all the details of
the original data. In such situations, "interactive data
visualization" becomes useful.

<!-- comment -->

First, let us consider a slightly more complicated form of the Keeling
Curve data visualization.
 

<!-- comment -->

The plot above shows colored points for each month of the year, rather
than for only May and September, the yearly local minima that we
wanted to emphasize. Since it lacks this emphasis, this static plot is
not as informative as the previous plot. This is an example where it
is not desirable to plot all of the data at once. We can solve this
problem using the following interactive plot.
 

<!-- comment -->

In the plot above, the default emphasis is May and October, but the
user can click the legend to update the emphasis. This simple example
illustrates the main idea of interactive data visualization using
animint. There are many choices that must be made to show details of
big data sets. For example, the choice of which months to emphasize in
the plot above. Rather than fixing such choices in a static plot, the
goal of interactive data visualization is to allow the reader to see
what the plot looks like when different choices are made. In the
example above, we used an interactive legend which allows the user to
select different months and see what the plot looks like after
changing the selection.

<!-- comment -->

The example above also provides a good example of `clickSelects` and
`showSelected`, the two keywords that animint introduces to allow
interaction. Without going into too many details, the plot above uses
`clickSelects=month` for the interactive legend, meaning that clicking
the legend should change the selected months. Furthermore, we used
`showSelected=month` for the points, meaning that we should only plot
the set of points which corresponds to the currently selected
months. In Chapters 3-4, we will explain how to design data
visualizations by writing R code using these two new keywords.

<!-- comment -->

## Chapter summary and exercises {#exercises}

<!-- comment -->

This chapter explained some basic facts about data, and gave
definitions of different sizes of data: small, medium, and large.

<!-- comment -->

* Based on the definitions introduced in this chapter, what is the
  difference between small and medium data?
* What is the difference between medium and large data?

<!-- comment -->

Next, [Chapter 2](Ch02-ggplot2.html) explains how the grammar of
graphics can be used to create data visualizations.